### 2024-04-13 manuscript.qmd
add manuscript markdown

---
title: Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
author:
  - name: Kitty B Murphy
    orcid: 
    corresponding: true
    email: kitty.murphy@ukdri.ac.uk
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Brian M Schilder
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Nathan G Skene
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
keywords:
  - HPO
  - chatGPT
  - rare disease 
bibliography: bibliography.bib
---

## Abstract

The Human phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised database for phenotypic abnormalities. However, there is currently no information pertaining to the severity of each phenotype, making systematic analyses and prioritisation of results difficult. With 18,082 abnormalities now corresponding to over 10,000 rare diseases, manual curation of such phenotypic annotations by experts would be labor-intensive and time-consuming. Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 model with Python to systematically annotate the severity of ~ 17,000 phenotypic abnormalities in the HPO. By checking that phenotypes with guaranteed outcomes were appropriately annotated, we demonstrate the potential for natural language processing technologies to automate the curation process accurately. For example, phenotypes such as "decreased male fertility" were used to compute a true positive rate, as they would be expected to be annotated as often, if not always, causing reduced fertility. Across the annotated outcomes, we observed > 73 % annotation accuracy. Using a novel approach, we developed a severity scoring system that incorporates both the nature of the phenotype outcome and the frequency of its occurrence. These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human well being, and best targets for therapeutic intervention.

## Introduction

Comprehensive annotation of phenotypic abnormalities is invaluable for defining, diagnosing, prognosing, and treating human disease. Since 2008, the Human phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes @Gargano2024-nj. Through developing open community resources, the depth and breadth of the HPO has continued to expand and there are now ~ 18,000 phenotypic abnormalities, corresponding to > 10,000 rare diseases, described. In recent years, the HPO has expanded its disease annotations so that each HPO term can have metadata including typical age of onset and frequency. In addition, there are the Clinical modifier (put this in italics) and Clinical course (also italics) subontologies, which contains terms to describe factors including severity and triggers, and mortality and progression, respectively. Describing the severity-related attributes of a disease is crucial for attaining significant objectives in rare diseases. This includes enhancing diagnostic capabilities, as well as prioritising and guiding gene therapy trials.

To date, the HPO has largely been manually curated by experts including clinicians, clinical geneticists, and researchers. Although this approach ensures the quality and accuracy of the ontology, it is time-consuming and labour-intensive. As a result, less than 1% of terms within the HPO have metadata pertaining to their features such as time course and severity. As artificial intelligence (AI) capabilities advance, there is an opportunity to integrate natural language processing technologies into assisting in the curation process. Here, we have used the OpenAI GPT-4 model (https://openai.com/) with Python to systematically annotate the severity of > 17,000 phenotypic abnormalities within the HPO. Our severity annotation framework was developed based on previously defined criteria developed through consultation with clinicians 2. The authors consulted 192 healthcare professionals for their opinions on the relative severity of various clinical characteristics: they used this to create a system for categorising the severity of  diseases. Briefly, each healthcare professional was sent a survey asking them to first rate how important a disease characteristic was for determining disease severity, and then to rate the severity of a set of given disease. Using the responses, the authors were able to categorise clinical characteristics into 4 ‘severity tiers’. While characteristics such as shortened lifespan in infancy and intellectual disability were identified as highly severe and placed into tier 1, sensory impairment and reduced lifespan were categorised as less severe and placed into tier 4. Being able to quickly ascribe severity measures based on those criteria to HPO phenotypes, will assist with interpreting phenome-wide studies.

Almost 800 phenotypes were annotated twice to evaluate annotation consistency, and a true positive rate of annotations was calculated to assess annotation accuracy. Additionally, based on the clinical characteristics and their occurrence, we have quantified the severity of each phenotype, providing an example of how these clinical characteristic annotations can be used to guide prioritisation of gene therapy trials. Ultimately, we hope that our resource will be of utility to those working in rare diseases, as well as the wider rare disease community.

## Results

### Annotating the HPO using OpenAI GPT-4
We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO). Our annotation framework was developed based on previously defined criteria for classifying disease severity @Lazarin2014-rz. We sought to evaluate the impact of phenotypes on outcomes including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset. Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response. For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred. Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance was achieved when inputting no more than two or three phenotypes per prompt. An example prompt can be seen in Figure 1. The phenotype outcome occurrence varied by category. While > 50% of annotated phenotypes never caused blindness, cancer, immunodeficiency, intellectual disability, reduced fertility, and sensory impairment, a similar proportion often or always had congenital onset (Fig. 1).

```{r occurence_plot}
#| echo: false
#| warning: false

# read in hpo gpt annotations 
hpo_annot <- HPOExplorer::gpt_annot_read()

# subset to clinical characteristics 
occurr_df <- hpo_annot[,c("hpo_name", "intellectual_disability", "death", "impaired_mobility", "blindness",
                          "sensory_impairments", "immunodeficiency", "cancer", "reduced_fertility",
                          "congenital_onset")]

cols_of_interest <- names(occurr_df)[-1]

# Create an empty dataframe to store the counts
count_df <- data.frame(matrix(0, nrow = length(cols_of_interest), ncol = 4))
colnames(count_df) <- c("always", "often", "rarely", "never")
rownames(count_df) <- cols_of_interest

# Loop through each column of interest and count the occurrences of 'always', 'often', 'rarely', 'never'
for (col in cols_of_interest) {
  counts <- table(occurr_df[[col]])
  count_df[col, ] <- counts[match(colnames(count_df), names(counts))]
}

count_df_long <- tidyr::gather(count_df, key = "condition", value = "count", always:never)
count_df_long$category <- rownames(count_df)

count_df_long$category <- gsub("_", " ", count_df_long$category)
count_df_long$condition <- factor(count_df_long$condition, levels=c("always", "often", "rarely", "never"))

count_df_long <- count_df_long %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(percentage = count / sum(count) * 100)

# set annotation order for plots 
annotation_order <- c("death", "intellectual disability", "impaired mobility", "blindness", 
                      "congenital onset", "physical malformations", "sensory impairments", 
                      "immunodeficiency", "cancer", "reduced fertility")

count_df_long$category <- factor(count_df_long$category, levels = annotation_order)

# Stacked bar plot of clinical characteristic occurrence
ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
  geom_bar(stat = "identity") +
  labs(title = "Clinical characteristic occurrence",
       x = "", y = "HPO phenotypes (n)", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "right", plot.title = element_text(hjust=0.5), 
        text = element_text(size=16),
        axis.text.x = element_text(angle=45, size=16, hjust=1), 
        title = element_text(size=15)) +
  scale_fill_brewer(palette = "GnBu", direction = -1)

```

![Figure 1: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics. a An example of the prompt input given to to GPT-4. The phenotypes listed in the last sentence were changed to allow ~17,000 phenotypes to be annotated. b Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all ~17,000 phenotypes annotated.The terms shown on the X-axis are the clinical characteristics, for which GPT-4 was asked to determine whether each phenotype caused them. For example, it annotated ‘Respiratory arrest’ as always causing death. Only 75 phenotypes were annotated as always causing death, despite 9,544 phenotypes often or rarely causing it.](figures/clinical_characteristic_occurrence.png)
### Annotation consistency and accuracy
To assess annotation consistency, 793 phenotypes underwent duplicate annotations, with two metrics employed to determine the 'consistency rate'. The first, less stringent metric, defined consistency as the duplicate annotations being either 'always' and 'often', or 'never' and 'rarely'. The second, more stringent metric, required exact agreement in annotation occurrences, e.g. 'always' and 'always'. For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least 80%, and for the more stringent metric, the lowest consistency rate was 57%, for congenital onset. An example of where annotations were inconsistent was for the HPO term ‘Acute leukaemia’. One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that ‘weakness and fatigue from leukaemia and its treatment can impair mobility’. The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that ‘acute leukaemia rarely impairs mobility directly’. Despite specifying in the prompt for GPT-4 not to take into consideration indirect effects, here is an example of where it failed to do so.

```{r annot_checks}
# Annotation checks
checks <- HPOExplorer::gpt_annot_check()

# Define items to be checked
items <- c("consistency_count", "consistency_rate",
           "consistency_stringent_count", "consistency_stringent_rate",
           "true_pos_count", "true_pos_rate")

metric_types <- c("Rate", "Count")[1]

# Get data frame with checks values
check_df <- lapply(checks[items], data.table::as.data.table, keep.rownames = TRUE) |>
  data.table::rbindlist(idcol = "metric") |>
  data.table::setnames(c("metric", "annotation", "value"))

check_df$metric <- factor(check_df$metric,
                          levels = items,
                          ordered = TRUE)

check_df$annotation <- factor(check_df$annotation,
                              levels = unique(check_df$annotation),
                              ordered = TRUE)

check_df[, metric_type := ifelse(grepl("count", metric),
                                 "Count", ifelse(grepl("rate", metric), "Rate", NA))]

check_df[, metric_category := gsub("_count|_rate", "", metric)]

check_df$metric_type <- factor(check_df$metric_type, ordered = TRUE)

check_df <- check_df[annotation != "pheno_count"]

check_df[, n := value[metric_type == "Count"],
         by = c("metric_category", "annotation")]

check_df$annotation <- gsub("_", " ", check_df$annotation)
annotation_order <- unique(check_df$annotation)
check_df$annotation <- factor(check_df$annotation, levels = annotation_order)
check_df$metric_category <- gsub("_", " ", check_df$metric_category)

metric_categories <- c("consistency", "consistency stringent")

```

```{r figure_2}
#| label: fig-checks
#| fig-cap: GPT-4 annotations are consistent and accurate across annotations. a Barplot showing the annotation consistency for 793 phenotypes that were annotated twice. On the left hand side, annotations were defined as consistent if the given phenotype was annotated to the same outcome occurrence both times, or if it was annotated to ‘always’ and ‘often’, or, ‘never’ and ‘rarely. On the right hand side, annotations were only defined as consistent if the given phenotype was annotated to the same outcome occurrence both times. b Bar plot of the true positive rate for each annotation. The true positives were determined based on the assumption that all descendants of some HPO phenotypes should be related to those outcomes. For instance, all descendents of the HPO term for “Neoplasm” were considered to be true positives for cancer (an assumption which is not always the case, as GPT-4 detects, as some neoplasms are benign). The labels above each bar indicate the number of phenotypes tested. 
#| fig-height: 4
#| fig-width: 6
consistency_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
                           aes(x = annotation, y = value,
                               fill = metric_category,
                               label = round(value, 2))) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  facet_grid(facets = ~metric_category, scales = "free") +
  theme_bw() +
  labs(x = "", y = "Value") +
  ggtitle("Annotation consistency") +
  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
        legend.position = "none",
        text = element_text(size = 16),
        title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(x = 7.5, y = 1.05, label = "N phenotypes = 793"), size = 4)

metric_categories <- "true pos"

accuracy_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
                        aes(x = annotation, y = value,
                            fill = metric_category,
                            label = n)) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  theme_bw() +
  labs(x = "", y = "Value") +
  ggtitle("True positive rate") +
  scale_fill_manual(values = c("#B1D5BBFF")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
        legend.position = "none",
        text = element_text(size = 16),
        title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  geom_label(fill = "white", size = 4)

consistency_plot + accuracy_plot + plot_layout(widths = c(1, 0.7)) +
  plot_annotation(tag_levels = 'a') & theme(plot.tag = element_text(size = 18, face = "bold"))
```

In order to evaluate the accuracy of the annotations, we calculated a true positive rate. This involved identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the presence of certain conditions. For instance, the phenotypes 'decreased fertility in females' and 'decreased fertility in males' should, in the annotations, often, if not always, cause reduced fertility. We observed an encouraging true positive rate exceeding 70% across all phenotype outcomes. The lowest rate was observed for cancer (73%), for which all descendants of the HPO term ‘Neoplasm’ were considered to be true positives. However, this assumption, as GPT-4 detects, is not always true as these abnormal tissue/cell growths can be either cancerous or benign. For example, the HPO term ‘multiple digital exostoses’, which are benign bone tumours, is a descendant of the HPO term ‘Neoplasm’, and GPT-4 annotated it as rarely causing cancer, giving the justification that ‘Bone growths may rarely become cancerous’. This high level of accuracy underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information. However, we acknowledge that the number of testable phenotypes for some of these categories are low. Furthermore, some of the ‘true positive’ phenotypes are synonymous to the clinical characteristic itself, for instance, for testing the true positive rate of the characteristic ‘Intellectual disability’, the phenotype ‘Severe intellectual disability’ was used.

### Quantifying phenotypic severity
Quantifying the severity of phenotypes can have important implications for diagnosis, prognosis, and treatment, for instance, guiding prioritisation of gene therapy trials for the most severe phenotypes. First, we created a dictionary to map each phenotype outcome (e.g. blindness) and its response (always, often, rarely, never) to numeric values from 0-4. Then, the phenotype outcome values were multiplied by their associated response values. Importantly, the values reflected the severity of each outcome based on both the outcome itself and the frequency of the response. For instance, a phenotype always causing death would have a higher multiplied value than a phenotype rarely causing reduced fertility. Next, we computed an average score for each phenotype by aggregating the multiplied values across all phenotype outcomes and then calculating the mean. This was then normalised by the theoretical maximum severity score, so that all phenotypes were on a 0-100 severity scale (where 100 is the most severe phenotype possible). This average normalised score represents the overall severity of the phenotype based on the severity of its individual outcomes.

Based on these scores we evaluated the top 50 severe phenotypes. The most severe phenotype, with a score of 58, was ‘Anencephaly’ (HP:0002323). Anencephaly is a birth defect where the baby is born without a portion of its brain and skull, often these babies are stillborn. In fact, many of the most severe phenotypes were related to developmental brain and neural tube defects. Comparison of the severity scores for each response, across the phenotype outcomes annotated, revealed consistent trends: as the response of the phenotype outcome increased (from never to always), the severity score also increased (Supplementary Fig. 1). We also evaluated the severity score distribution by HPO branch and calculated the mean severity score using all phenotypes within a HPO branch (Fig. 4). After the term ‘Phenotypic abnormality’, the highest severity score was for the HPO branch ‘Abnormal cellular phenotype’, followed by ‘Abnormality of prenatal development or birth’, which would include the high severity scoring phenotypes seen in Fig. 3.

```{r severity_scoring}
# get all descendants of the term phenotypic abnormality 
keep_descendants="Phenotypic abnormality"
keep_ont_levels=seq(3,17)

ancestor_name <- variable <- hpo_id <- hpo_name <-
  value <- severity_score_gpt <- mean_severity_score_gpt <- NULL;

#### Prepare annotation results ####
gpt_annot <- HPOExplorer::gpt_annot_read()
res_coded <- HPOExplorer::gpt_annot_codify(annot = gpt_annot)
dat1 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)

#### Filter out ont levels ####
dat1 <- add_ancestor(dat1,
                     keep_descendants = keep_descendants)
data.table::setorderv(dat1,"severity_score_gpt",-1)

dat1 <- HPOExplorer::add_ont_lvl(dat1, keep_ont_levels = keep_ont_levels)
dat1$variable <- gsub("_", " ", dat1$variable)

# get all unique phenotypes
total_unique <- length(unique(dat1$hpo_id))

# get the top N phenotypes
top_n=15
dat_top <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(top_n)]]

# Get the bottom N phenotypes
bottom_n <- 15
dat_bottom <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(total_unique - bottom_n + 1, total_unique)]]

# Define the range for selecting phenotypes from the middle
start_index <- ceiling(total_unique / 2)  
end_index <- start_index + 12

# Get phenotypes from the middle
dat_middle <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(start_index, end_index)]]

# define labels
dat_top$position <- "Top"
dat_middle$position <- "Middle"
dat_bottom$position <- "Bottom"

dat_all <- rbind(dat_top, dat_middle, dat_bottom)
dat_all$position <- factor(dat_all$position, levels=c("Top", "Middle", "Bottom"))

dat_all$variable <- factor(dat_all$variable, levels = annotation_order)
```

```{r figure_3}
#| label: fig-severity
#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions. Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score. Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic (death = 5, intellectual disability = 5, impaired mobility = 4, blindness = 4, congenital onset = 4, physical malformations = 3, sensory impairments = 3, immunodeficiency = 3,  cancer = 3, reduced fertility = 1) by their corresponding response frequencies (Always = 4, Often = 3, Rarely = 2, Never = 0). The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score. The x-axis is scaled by the severity of the clinical characterisation.
#| fig-height: 4
#| fig-width: 6

p1 <- ggplot2::ggplot(data = dat_top,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right', 
                 text = element_text(size=16)) + 
  labs(x = "Annotation",
       y = "HPO term", 
       fill = "")

# Heatmap of severity scores for top phenotypes
p2 <- ggplot2::ggplot(data = dat_middle,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right', 
                 text = ggplot2::element_text(size=16)) + 
  ggplot2::labs(x = "Annotation",
                y = "HPO term", 
                fill = "")

# Heatmap of severity scores for top phenotypes
p3 <- ggplot2::ggplot(data = dat_bottom,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right', 
                 text = ggplot2::element_text(size=16)) + 
  ggplot2::labs(x = "Annotation",
                y = "HPO term", 
                fill = "")

# Combine the plots
p_combined <- patchwork::wrap_plots(p1, p2, p3, ncol = 3)

# Print the combined plot
print(p_combined)
```

## Discussion

Here, we present a novel approach leveraging the OpenAI GPT-4 model, to systematically annotate the severity of 17,000 phenotypic abnormalities within the HPO. Our findings highlight the potential of natural language processing technologies in significantly contributing to the automation and refinement of the curation process in genomics and rare disease research.
Manual curation of ontologies has traditionally relied on expert input, a process that is both time-consuming and labor-intensive. By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without compromising accuracy. Our validation approach yielded a high true positive rate exceeding 73% across the phenotypes tested. Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.

A key contribution of our study is the development of a severity scoring system that integrates both the nature of the phenotype outcome and the frequency of its occurrence. By quantifying the phenotypic severity this way, we highlight the most impactful conditions and provide a comprehensive framework for prioritising gene therapy trials and guiding clinical decision-making in rare diseases.

** Brian to pick examples from severity scoring **

Given its scalability it can be easily applied as more phenotypes are added to the HPO, but can also be extended to include different annotations and to be used with other ontologies. For instance, we could have also tried to use GPT-4 to annotate how prevalent a phenotype is in the human population, or what the cost to the health system is of a phenotype. Currently there are no systematic annotations of these on a phenotypic level. However, it’s unclear whether GPT-4 could currently be applied for such annotations, or whether they require much more extensive data collection.

Importantly, there are additional factors that need to be considered when trying to prioritise phenotypes for their suitability for gene therapy development. First, our scoring is in some sense subjective, for instance, one could ask is something that always causes death, always worse than something which causes a lifetime of other severe clinical characteristics. In terms of viability for a clinical trial, a phenotype which causes a lifetime of expensive burden on the medical system may be a better target than a phenotype which causes a quick death with little or no other clinical characteristics before. It would be worth exploring whether GPT-4 could estimate such costs. Another factor that affects the viability of developing a treatment for a therapy is how quickly a clinical trial can be run. For instance, measuring a high risk of respiratory failure over ten years would require a long trial. However, testing for total reversal of an existing severe phenotype would be much faster. As AI capabilities continue to advance, perhaps this is something GPT-4 and other models of the like could test for.

An additional annotation that we didn’t take into consideration here is that GPT-4 might be aware of whether there are existing treatments for these phenotypes. Therefore, some of the top-ranking severe phenotype we have identified may already have treatment options available. It would therefore be useful to know whether a phenotype, although highly severe, is currently treatable, and also whether there exist known clinical endpoints for clinical trial testing. GPT-4 does seem to take into account quality of care, i.e. the degree to which health services increase the likelihood of desired outcomes (World Health Organization). For example, many of the cancer phenotypes are justified as always or often causing death unless treated or caught early. On the other hand, some cancers are justified as rarely causing death if appropriate treatment is provided, which may not always be the case for individuals or populations with access to less advanced medical systems. In future efforts, we might want to ask GPT-4 more explicitly to distinguish between whether a phenotype would cause death without treatment, or with best available standard of care.

While our study demonstrates the feasibility and utility of AI-driven phenotypic annotation, several limitations must be acknowledged. The reliance on computational algorithms may introduce biases or inaccuracies inherent to the training data, necessitating ongoing validation and refinement of our approach. Additionally, our severity scoring system, while comprehensive, may not capture the full spectrum of phenotypic variability or account for complex gene-environment interactions. Future research should focus on further optimising AI-driven annotation methodologies, incorporating additional data modalities such as genomic and clinical data to enhance accuracy.

In conclusion, our study represents a significant step towards harnessing the power of AI to advance phenotypic annotation and severity assessment in rare diseases. This resource aims to provide researchers and clinicians with actionable insights that can inform rare disease research and improve the lives of individuals affected by rare diseases.

## Methods

### Annotating the HPO using OpenAI GPT-4
We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO). Our annotation framework was developed based on previously defined criteria for classifying disease severity (Lazarin et al., 2014). We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset. Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response. For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.

### Calculating the true positive rate
A true positive rate was calculated as a measure of the accuracy of the GPT-4 annotations. This was achieved by identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the occurrence of certain clinical characteristics, and using all descendants of this HPO branch as true positives. For example, all descendants of the terms Abnormal central motor function and abnormality of movement should always or often cause impaired mobility (Table 1).

| Clinical characteristics | True positive HPO branch |
| ------------------------ | ------------------------ |
| Intellectual disability  | Intellectual disability (HP:0001249)
| Impaired mobility.       | Abnormal central motor function (HP:0011442) and Abnormality of movement (HP:0100022)
| Physical malformations   | Malformation (string search)
| Blindness                | Blindness (HP:0000618)
| Sensory impairments      | Abnormality of vision (HP:0000504), Abnormality of the sense of smell (HP:0004408), Abnormality of taste sensation (HP:0000223), Somatic sensory dysfunction (HP:0003474), and Hearing abnormality (HP:0000364)
| Immunodeficiency         | Immunodeficiency (HP:0002721)
| Cancer                   | Neoplasm (HP:0002664)
| Reduced fertility        | Decreased fertility (HP:0000144)

### Quantifying phenotypic severity
The GPT-4 generated annotation occurrences were converted into a semi-quantitative scoring system, with ‘always’ corresponding to 3, ‘often’ to 2, ‘rarely’ to 1, and ‘never’ to 0. These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided annotations (Table 2). Subsequently, the weighted scores underwent normalisation to yield a final quantitative severity score ranging from 0-100, with 100 signifying the maximum severity score attainable.

| Clinical characteristic     | Always (4) | Often (3) | Rarely (2) | Never (0) |
| Intellectual_disability (5) |     20     |     15    |    10      |     0     | 
| Impaired_mobility (5)       |     20     |     15    |    10      |     0     | 
| Blindness (4)               |     16     |     12    |    8       |     0     | 
| Congenital onset (4)        |     16     |     12    |    8       |     0     | 
| Physical_malformations (3)  |     12     |     9     |    6       |     0     | 
| Sensory_impairments (3)     |     12     |     9     |    6       |     0     | 
| Immunodeficiency (3)        |     12     |     9     |    6       |     0     | 
| Cancer (3)                  |     12     |     9     |    6       |     0     | 
| Reduced_fertility (1)       |     4      |     3     |    2       |     0     |

## Supplementary figures

```{r supp_figure_1}
dat2 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
dat2 <- add_ancestor(dat2,
                     keep_descendants = keep_descendants)
dat2[,mean_severity_score_gpt:=mean(severity_score_gpt, na.rm=TRUE),
     by="ancestor_name"] |>
  data.table::setorderv("mean_severity_score_gpt", -1, na.last = TRUE)
dat2[,ancestor_name:=factor(ancestor_name,
                            levels = unique(dat2$ancestor_name),
                            ordered = TRUE)]

ggplot2::ggplot(dat2, ggplot2::aes(x=severity_score_gpt
                                   # fill=factor(congenital_onset)
)) +
  ggplot2::geom_histogram(bins = 50, fill="slateblue") +
  ggplot2::geom_vline(ggplot2::aes(xintercept=mean_severity_score_gpt),
                      color="red") +
  ggplot2::geom_label(data = unique(
    dat2[,list(mean_severity_score_gpt), by="ancestor_name"]
  ),
  ggplot2::aes(x=mean_severity_score_gpt,
               y=Inf,
               label=round(mean_severity_score_gpt,2)),
  color="red", size=3, hjust = 0, vjust = 1.5, alpha=.65) +
  ggplot2::facet_wrap(facets = "ancestor_name~.", scales = "free_y",
                      ncol = 3) +
  ggplot2::labs(y="Frequency", x="Severity score") +
  ggplot2::theme_bw() +
  ggplot2::theme(strip.background = ggplot2::element_rect(
    fill = "transparent"), 
    text = element_text(size=16))
```

## References {.unnumbered}

:::{#refs}

:::

### 2024-04-15 _manuscript/index.qmd
add manuscript files

--- 2024-04-13 manuscript.qmd
@@
   - rare disease 
-bibliography: bibliography.bib
 ---
@@
 Comprehensive annotation of phenotypic abnormalities is invaluable for defining, diagnosing, prognosing, and treating human disease.
-Since 2008, the Human phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes @Gargano2024-nj.
+Since 2008, the Human phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes1.
 Through developing open community resources, the depth and breadth of the HPO has continued to expand and there are now ~ 18,000 phenotypic abnormalities, corresponding to > 10,000 rare diseases, described.
@@
 
-## Results
-
-### Annotating the HPO using OpenAI GPT-4
-We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO).
-Our annotation framework was developed based on previously defined criteria for classifying disease severity @Lazarin2014-rz.
-We sought to evaluate the impact of phenotypes on outcomes including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
-Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
-Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance was achieved when inputting no more than two or three phenotypes per prompt.
-An example prompt can be seen in Figure 1.
-The phenotype outcome occurrence varied by category.
-While > 50% of annotated phenotypes never caused blindness, cancer, immunodeficiency, intellectual disability, reduced fertility, and sensory impairment, a similar proportion often or always had congenital onset (Fig.
-1).
-
-```{r occurence_plot}
-#| echo: false
-#| warning: false
-
-# read in hpo gpt annotations 
-hpo_annot <- HPOExplorer::gpt_annot_read()
-
-# subset to clinical characteristics 
-occurr_df <- hpo_annot[,c("hpo_name", "intellectual_disability", "death", "impaired_mobility", "blindness",
-                          "sensory_impairments", "immunodeficiency", "cancer", "reduced_fertility",
-                          "congenital_onset")]
-
-cols_of_interest <- names(occurr_df)[-1]
-
-# Create an empty dataframe to store the counts
-count_df <- data.frame(matrix(0, nrow = length(cols_of_interest), ncol = 4))
-colnames(count_df) <- c("always", "often", "rarely", "never")
-rownames(count_df) <- cols_of_interest
-
-# Loop through each column of interest and count the occurrences of 'always', 'often', 'rarely', 'never'
-for (col in cols_of_interest) {
-  counts <- table(occurr_df[[col]])
-  count_df[col, ] <- counts[match(colnames(count_df), names(counts))]
-}
-
-count_df_long <- tidyr::gather(count_df, key = "condition", value = "count", always:never)
-count_df_long$category <- rownames(count_df)
-
-count_df_long$category <- gsub("_", " ", count_df_long$category)
-count_df_long$condition <- factor(count_df_long$condition, levels=c("always", "often", "rarely", "never"))
-
-count_df_long <- count_df_long %>%
-  dplyr::group_by(category) %>%
-  dplyr::mutate(percentage = count / sum(count) * 100)
-
-# set annotation order for plots 
-annotation_order <- c("death", "intellectual disability", "impaired mobility", "blindness", 
-                      "congenital onset", "physical malformations", "sensory impairments", 
-                      "immunodeficiency", "cancer", "reduced fertility")
-
-count_df_long$category <- factor(count_df_long$category, levels = annotation_order)
-
-# Stacked bar plot of clinical characteristic occurrence
-ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
-  geom_bar(stat = "identity") +
-  labs(title = "Clinical characteristic occurrence",
-       x = "", y = "HPO phenotypes (n)", fill = NULL) +
-  theme_minimal() +
-  theme(legend.position = "right", plot.title = element_text(hjust=0.5), 
-        text = element_text(size=16),
-        axis.text.x = element_text(angle=45, size=16, hjust=1), 
-        title = element_text(size=15)) +
-  scale_fill_brewer(palette = "GnBu", direction = -1)
-
+```{r}
+eruptions <- c(1492, 1585, 1646, 1677, 1712, 1949, 1971, 2021)
+n_eruptions <- length(eruptions)
 ```
 
-![Figure 1: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics.
-a An example of the prompt input given to to GPT-4.
-The phenotypes listed in the last sentence were changed to allow ~17,000 phenotypes to be annotated.
-b Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all ~17,000 phenotypes annotated.The terms shown on the X-axis are the clinical characteristics, for which GPT-4 was asked to determine whether each phenotype caused them.
-For example, it annotated ‘Respiratory arrest’ as always causing death.
-Only 75 phenotypes were annotated as always causing death, despite 9,544 phenotypes often or rarely causing it.](figures/clinical_characteristic_occurrence.png)
-### Annotation consistency and accuracy
-To assess annotation consistency, 793 phenotypes underwent duplicate annotations, with two metrics employed to determine the 'consistency rate'.
-The first, less stringent metric, defined consistency as the duplicate annotations being either 'always' and 'often', or 'never' and 'rarely'.
-The second, more stringent metric, required exact agreement in annotation occurrences, e.g.
-'always' and 'always'.
-For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least 80%, and for the more stringent metric, the lowest consistency rate was 57%, for congenital onset.
-An example of where annotations were inconsistent was for the HPO term ‘Acute leukaemia’.
-One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that ‘weakness and fatigue from leukaemia and its treatment can impair mobility’.
-The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that ‘acute leukaemia rarely impairs mobility directly’.
-Despite specifying in the prompt for GPT-4 not to take into consideration indirect effects, here is an example of where it failed to do so.
-
-```{r annot_checks}
-# Annotation checks
-checks <- HPOExplorer::gpt_annot_check()
-
-# Define items to be checked
-items <- c("consistency_count", "consistency_rate",
-           "consistency_stringent_count", "consistency_stringent_rate",
-           "true_pos_count", "true_pos_rate")
-
-metric_types <- c("Rate", "Count")[1]
-
-# Get data frame with checks values
-check_df <- lapply(checks[items], data.table::as.data.table, keep.rownames = TRUE) |>
-  data.table::rbindlist(idcol = "metric") |>
-  data.table::setnames(c("metric", "annotation", "value"))
-
-check_df$metric <- factor(check_df$metric,
-                          levels = items,
-                          ordered = TRUE)
-
-check_df$annotation <- factor(check_df$annotation,
-                              levels = unique(check_df$annotation),
-                              ordered = TRUE)
-
-check_df[, metric_type := ifelse(grepl("count", metric),
-                                 "Count", ifelse(grepl("rate", metric), "Rate", NA))]
-
-check_df[, metric_category := gsub("_count|_rate", "", metric)]
-
-check_df$metric_type <- factor(check_df$metric_type, ordered = TRUE)
-
-check_df <- check_df[annotation != "pheno_count"]
-
-check_df[, n := value[metric_type == "Count"],
-         by = c("metric_category", "annotation")]
-
-check_df$annotation <- gsub("_", " ", check_df$annotation)
-annotation_order <- unique(check_df$annotation)
-check_df$annotation <- factor(check_df$annotation, levels = annotation_order)
-check_df$metric_category <- gsub("_", " ", check_df$metric_category)
-
-metric_categories <- c("consistency", "consistency stringent")
-
+```{r}
+#| label: fig-timeline
+#| fig-cap: Timeline of recent earthquakes on La Palma
+#| fig-alt: An event plot of the years of the last 8 eruptions on La Palma.
+#| fig-height: 1.5
+#| fig-width: 6
+par(mar = c(3, 1, 1, 1) + 0.1)
+plot(eruptions, rep(0, n_eruptions), 
+  pch = "|", axes = FALSE)
+axis(1)
+box()
 ```
 
-```{r figure_2}
-#| label: fig-checks
-#| fig-cap: GPT-4 annotations are consistent and accurate across annotations.
-a Barplot showing the annotation consistency for 793 phenotypes that were annotated twice.
-On the left hand side, annotations were defined as consistent if the given phenotype was annotated to the same outcome occurrence both times, or if it was annotated to ‘always’ and ‘often’, or, ‘never’ and ‘rarely.
-On the right hand side, annotations were only defined as consistent if the given phenotype was annotated to the same outcome occurrence both times.
-b Bar plot of the true positive rate for each annotation.
-The true positives were determined based on the assumption that all descendants of some HPO phenotypes should be related to those outcomes.
-For instance, all descendents of the HPO term for “Neoplasm” were considered to be true positives for cancer (an assumption which is not always the case, as GPT-4 detects, as some neoplasms are benign).
-The labels above each bar indicate the number of phenotypes tested.
-#| fig-height: 4
-#| fig-width: 6
-consistency_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
-                           aes(x = annotation, y = value,
-                               fill = metric_category,
-                               label = round(value, 2))) +
-  geom_bar(stat = "identity", show.legend = TRUE) +
-  facet_grid(facets = ~metric_category, scales = "free") +
-  theme_bw() +
-  labs(x = "", y = "Value") +
-  ggtitle("Annotation consistency") +
-  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
-        legend.position = "none",
-        text = element_text(size = 16),
-        title = element_text(size = 15),
-        plot.title = element_text(hjust = 0.5)) +
-  geom_text(aes(x = 7.5, y = 1.05, label = "N phenotypes = 793"), size = 4)
-
-metric_categories <- "true pos"
-
-accuracy_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
-                        aes(x = annotation, y = value,
-                            fill = metric_category,
-                            label = n)) +
-  geom_bar(stat = "identity", show.legend = TRUE) +
-  theme_bw() +
-  labs(x = "", y = "Value") +
-  ggtitle("True positive rate") +
-  scale_fill_manual(values = c("#B1D5BBFF")) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
-        legend.position = "none",
-        text = element_text(size = 16),
-        title = element_text(size = 15),
-        plot.title = element_text(hjust = 0.5)) +
-  geom_label(fill = "white", size = 4)
-
-consistency_plot + accuracy_plot + plot_layout(widths = c(1, 0.7)) +
-  plot_annotation(tag_levels = 'a') & theme(plot.tag = element_text(size = 18, face = "bold"))
+```{r}
+#| output: false
+avg_years_between_eruptions <- mean(diff(eruptions[-n_eruptions]))
+avg_years_between_eruptions
 ```
 
-In order to evaluate the accuracy of the annotations, we calculated a true positive rate.
-This involved identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the presence of certain conditions.
-For instance, the phenotypes 'decreased fertility in females' and 'decreased fertility in males' should, in the annotations, often, if not always, cause reduced fertility.
-We observed an encouraging true positive rate exceeding 70% across all phenotype outcomes.
-The lowest rate was observed for cancer (73%), for which all descendants of the HPO term ‘Neoplasm’ were considered to be true positives.
-However, this assumption, as GPT-4 detects, is not always true as these abnormal tissue/cell growths can be either cancerous or benign.
-For example, the HPO term ‘multiple digital exostoses’, which are benign bone tumours, is a descendant of the HPO term ‘Neoplasm’, and GPT-4 annotated it as rarely causing cancer, giving the justification that ‘Bone growths may rarely become cancerous’.
-This high level of accuracy underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information.
-However, we acknowledge that the number of testable phenotypes for some of these categories are low.
-Furthermore, some of the ‘true positive’ phenotypes are synonymous to the clinical characteristic itself, for instance, for testing the true positive rate of the characteristic ‘Intellectual disability’, the phenotype ‘Severe intellectual disability’ was used.
+Based on data up to and including 1971, eruptions on La Palma happen every `r round(avg_years_between_eruptions, 1)` years on average.
 
-### Quantifying phenotypic severity
-Quantifying the severity of phenotypes can have important implications for diagnosis, prognosis, and treatment, for instance, guiding prioritisation of gene therapy trials for the most severe phenotypes.
-First, we created a dictionary to map each phenotype outcome (e.g.
-blindness) and its response (always, often, rarely, never) to numeric values from 0-4.
-Then, the phenotype outcome values were multiplied by their associated response values.
-Importantly, the values reflected the severity of each outcome based on both the outcome itself and the frequency of the response.
-For instance, a phenotype always causing death would have a higher multiplied value than a phenotype rarely causing reduced fertility.
-Next, we computed an average score for each phenotype by aggregating the multiplied values across all phenotype outcomes and then calculating the mean.
-This was then normalised by the theoretical maximum severity score, so that all phenotypes were on a 0-100 severity scale (where 100 is the most severe phenotype possible).
-This average normalised score represents the overall severity of the phenotype based on the severity of its individual outcomes.
+Studies of the magma systems feeding the volcano, such as @marrero2019, have proposed that there are two main magma reservoirs feeding the Cumbre Vieja volcano; one in the mantle (30-40km depth) which charges and in turn feeds a shallower crustal reservoir (10-20km depth).
 
-Based on these scores we evaluated the top 50 severe phenotypes.
-The most severe phenotype, with a score of 58, was ‘Anencephaly’ (HP:0002323).
-Anencephaly is a birth defect where the baby is born without a portion of its brain and skull, often these babies are stillborn.
-In fact, many of the most severe phenotypes were related to developmental brain and neural tube defects.
-Comparison of the severity scores for each response, across the phenotype outcomes annotated, revealed consistent trends: as the response of the phenotype outcome increased (from never to always), the severity score also increased (Supplementary Fig.
-1).
-We also evaluated the severity score distribution by HPO branch and calculated the mean severity score using all phenotypes within a HPO branch (Fig.
-4).
-After the term ‘Phenotypic abnormality’, the highest severity score was for the HPO branch ‘Abnormal cellular phenotype’, followed by ‘Abnormality of prenatal development or birth’, which would include the high severity scoring phenotypes seen in Fig.
-3.
+Eight eruptions have been recorded since the late 1400s (@fig-timeline).
 
-```{r severity_scoring}
-# get all descendants of the term phenotypic abnormality 
-keep_descendants="Phenotypic abnormality"
-keep_ont_levels=seq(3,17)
+Data and methods are discussed in @sec-data-methods.
 
-ancestor_name <- variable <- hpo_id <- hpo_name <-
-  value <- severity_score_gpt <- mean_severity_score_gpt <- NULL;
+Let $x$ denote the number of eruptions in a year. Then, $x$ can be modeled by a Poisson distribution
 
-#### Prepare annotation results ####
-gpt_annot <- HPOExplorer::gpt_annot_read()
-res_coded <- HPOExplorer::gpt_annot_codify(annot = gpt_annot)
-dat1 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
+$$
+p(x) = \frac{e^{-\lambda} \lambda^{x}}{x !}
+$$ {#eq-poisson}
 
-#### Filter out ont levels ####
-dat1 <- add_ancestor(dat1,
-                     keep_descendants = keep_descendants)
-data.table::setorderv(dat1,"severity_score_gpt",-1)
+where $\lambda$ is the rate of eruptions per year. Using @eq-poisson, the probability of an eruption in the next $t$ years can be calculated.
 
-dat1 <- HPOExplorer::add_ont_lvl(dat1, keep_ont_levels = keep_ont_levels)
-dat1$variable <- gsub("_", " ", dat1$variable)
+| Name                 | Year   |
+| -------------------- | ------ |
+| Current              | 2021   |
+| Teneguía             | 1971   |
+| Nambroque            | 1949   |
+| El Charco            | 1712   |
+| Volcán San Antonio   | 1677   |
+| Volcán San Martin    | 1646   |
+| Tajuya near El Paso  | 1585   |
+| Montaña Quemada      | 1492   |
 
-# get all unique phenotypes
-total_unique <- length(unique(dat1$hpo_id))
+: Recent historic eruptions on La Palma {#tbl-history}
 
-# get the top N phenotypes
-top_n=15
-dat_top <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(top_n)]]
+@tbl-history summarises the eruptions recorded since the colonization of the islands by Europeans in the late 1400s.
 
-# Get the bottom N phenotypes
-bottom_n <- 15
-dat_bottom <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(total_unique - bottom_n + 1, total_unique)]]
+## Data & Methods {#sec-data-methods}
 
-# Define the range for selecting phenotypes from the middle
-start_index <- ceiling(total_unique / 2)  
-end_index <- start_index + 12
-
-# Get phenotypes from the middle
-dat_middle <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(start_index, end_index)]]
-
-# define labels
-dat_top$position <- "Top"
-dat_middle$position <- "Middle"
-dat_bottom$position <- "Bottom"
-
-dat_all <- rbind(dat_top, dat_middle, dat_bottom)
-dat_all$position <- factor(dat_all$position, levels=c("Top", "Middle", "Bottom"))
-
-dat_all$variable <- factor(dat_all$variable, levels = annotation_order)
-```
-
-```{r figure_3}
-#| label: fig-severity
-#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions.
-Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score.
-Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic (death = 5, intellectual disability = 5, impaired mobility = 4, blindness = 4, congenital onset = 4, physical malformations = 3, sensory impairments = 3, immunodeficiency = 3,  cancer = 3, reduced fertility = 1) by their corresponding response frequencies (Always = 4, Often = 3, Rarely = 2, Never = 0).
-The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score.
-The x-axis is scaled by the severity of the clinical characterisation.
-#| fig-height: 4
-#| fig-width: 6
-
-p1 <- ggplot2::ggplot(data = dat_top,
-                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
-  ggplot2::geom_tile() +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
-  ggplot2::theme_bw() +
-  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right', 
-                 text = element_text(size=16)) + 
-  labs(x = "Annotation",
-       y = "HPO term", 
-       fill = "")
-
-# Heatmap of severity scores for top phenotypes
-p2 <- ggplot2::ggplot(data = dat_middle,
-                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
-  ggplot2::geom_tile() +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
-  ggplot2::theme_bw() +
-  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right', 
-                 text = ggplot2::element_text(size=16)) + 
-  ggplot2::labs(x = "Annotation",
-                y = "HPO term", 
-                fill = "")
-
-# Heatmap of severity scores for top phenotypes
-p3 <- ggplot2::ggplot(data = dat_bottom,
-                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
-  ggplot2::geom_tile() +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
-  ggplot2::theme_bw() +
-  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right', 
-                 text = ggplot2::element_text(size=16)) + 
-  ggplot2::labs(x = "Annotation",
-                y = "HPO term", 
-                fill = "")
-
-# Combine the plots
-p_combined <- patchwork::wrap_plots(p1, p2, p3, ncol = 3)
-
-# Print the combined plot
-print(p_combined)
-```
-
-## Discussion
-
-Here, we present a novel approach leveraging the OpenAI GPT-4 model, to systematically annotate the severity of 17,000 phenotypic abnormalities within the HPO.
-Our findings highlight the potential of natural language processing technologies in significantly contributing to the automation and refinement of the curation process in genomics and rare disease research.
-Manual curation of ontologies has traditionally relied on expert input, a process that is both time-consuming and labor-intensive.
-By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without compromising accuracy.
-Our validation approach yielded a high true positive rate exceeding 73% across the phenotypes tested.
-Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.
-
-A key contribution of our study is the development of a severity scoring system that integrates both the nature of the phenotype outcome and the frequency of its occurrence.
-By quantifying the phenotypic severity this way, we highlight the most impactful conditions and provide a comprehensive framework for prioritising gene therapy trials and guiding clinical decision-making in rare diseases.
-
-** Brian to pick examples from severity scoring **
-
-Given its scalability it can be easily applied as more phenotypes are added to the HPO, but can also be extended to include different annotations and to be used with other ontologies.
-For instance, we could have also tried to use GPT-4 to annotate how prevalent a phenotype is in the human population, or what the cost to the health system is of a phenotype.
-Currently there are no systematic annotations of these on a phenotypic level.
-However, it’s unclear whether GPT-4 could currently be applied for such annotations, or whether they require much more extensive data collection.
-
-Importantly, there are additional factors that need to be considered when trying to prioritise phenotypes for their suitability for gene therapy development.
-First, our scoring is in some sense subjective, for instance, one could ask is something that always causes death, always worse than something which causes a lifetime of other severe clinical characteristics.
-In terms of viability for a clinical trial, a phenotype which causes a lifetime of expensive burden on the medical system may be a better target than a phenotype which causes a quick death with little or no other clinical characteristics before.
-It would be worth exploring whether GPT-4 could estimate such costs.
-Another factor that affects the viability of developing a treatment for a therapy is how quickly a clinical trial can be run.
-For instance, measuring a high risk of respiratory failure over ten years would require a long trial.
-However, testing for total reversal of an existing severe phenotype would be much faster.
-As AI capabilities continue to advance, perhaps this is something GPT-4 and other models of the like could test for.
-
-An additional annotation that we didn’t take into consideration here is that GPT-4 might be aware of whether there are existing treatments for these phenotypes.
-Therefore, some of the top-ranking severe phenotype we have identified may already have treatment options available.
-It would therefore be useful to know whether a phenotype, although highly severe, is currently treatable, and also whether there exist known clinical endpoints for clinical trial testing.
-GPT-4 does seem to take into account quality of care, i.e.
-the degree to which health services increase the likelihood of desired outcomes (World Health Organization).
-For example, many of the cancer phenotypes are justified as always or often causing death unless treated or caught early.
-On the other hand, some cancers are justified as rarely causing death if appropriate treatment is provided, which may not always be the case for individuals or populations with access to less advanced medical systems.
-In future efforts, we might want to ask GPT-4 more explicitly to distinguish between whether a phenotype would cause death without treatment, or with best available standard of care.
-
-While our study demonstrates the feasibility and utility of AI-driven phenotypic annotation, several limitations must be acknowledged.
-The reliance on computational algorithms may introduce biases or inaccuracies inherent to the training data, necessitating ongoing validation and refinement of our approach.
-Additionally, our severity scoring system, while comprehensive, may not capture the full spectrum of phenotypic variability or account for complex gene-environment interactions.
-Future research should focus on further optimising AI-driven annotation methodologies, incorporating additional data modalities such as genomic and clinical data to enhance accuracy.
-
-In conclusion, our study represents a significant step towards harnessing the power of AI to advance phenotypic annotation and severity assessment in rare diseases.
-This resource aims to provide researchers and clinicians with actionable insights that can inform rare disease research and improve the lives of individuals affected by rare diseases.
-
-## Methods
-
-### Annotating the HPO using OpenAI GPT-4
-We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO).
-Our annotation framework was developed based on previously defined criteria for classifying disease severity (Lazarin et al., 2014).
-We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
-Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
-
-### Calculating the true positive rate
-A true positive rate was calculated as a measure of the accuracy of the GPT-4 annotations.
-This was achieved by identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the occurrence of certain clinical characteristics, and using all descendants of this HPO branch as true positives.
-For example, all descendants of the terms Abnormal central motor function and abnormality of movement should always or often cause impaired mobility (Table 1).
-
-| Clinical characteristics | True positive HPO branch |
-| ------------------------ | ------------------------ |
-| Intellectual disability  | Intellectual disability (HP:0001249)
-| Impaired mobility.       | Abnormal central motor function (HP:0011442) and Abnormality of movement (HP:0100022)
-| Physical malformations   | Malformation (string search)
-| Blindness                | Blindness (HP:0000618)
-| Sensory impairments      | Abnormality of vision (HP:0000504), Abnormality of the sense of smell (HP:0004408), Abnormality of taste sensation (HP:0000223), Somatic sensory dysfunction (HP:0003474), and Hearing abnormality (HP:0000364)
-| Immunodeficiency         | Immunodeficiency (HP:0002721)
-| Cancer                   | Neoplasm (HP:0002664)
-| Reduced fertility        | Decreased fertility (HP:0000144)
-
-### Quantifying phenotypic severity
-The GPT-4 generated annotation occurrences were converted into a semi-quantitative scoring system, with ‘always’ corresponding to 3, ‘often’ to 2, ‘rarely’ to 1, and ‘never’ to 0.
-These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided annotations (Table 2).
-Subsequently, the weighted scores underwent normalisation to yield a final quantitative severity score ranging from 0-100, with 100 signifying the maximum severity score attainable.
-
-| Clinical characteristic     | Always (4) | Often (3) | Rarely (2) | Never (0) |
-| Intellectual_disability (5) |     20     |     15    |    10      |     0     | 
-| Impaired_mobility (5)       |     20     |     15    |    10      |     0     | 
-| Blindness (4)               |     16     |     12    |    8       |     0     | 
-| Congenital onset (4)        |     16     |     12    |    8       |     0     | 
-| Physical_malformations (3)  |     12     |     9     |    6       |     0     | 
-| Sensory_impairments (3)     |     12     |     9     |    6       |     0     | 
-| Immunodeficiency (3)        |     12     |     9     |    6       |     0     | 
-| Cancer (3)                  |     12     |     9     |    6       |     0     | 
-| Reduced_fertility (1)       |     4      |     3     |    2       |     0     |
-
-## Supplementary figures
-
-```{r supp_figure_1}
-dat2 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
-dat2 <- add_ancestor(dat2,
-                     keep_descendants = keep_descendants)
-dat2[,mean_severity_score_gpt:=mean(severity_score_gpt, na.rm=TRUE),
-     by="ancestor_name"] |>
-  data.table::setorderv("mean_severity_score_gpt", -1, na.last = TRUE)
-dat2[,ancestor_name:=factor(ancestor_name,
-                            levels = unique(dat2$ancestor_name),
-                            ordered = TRUE)]
-
-ggplot2::ggplot(dat2, ggplot2::aes(x=severity_score_gpt
-                                   # fill=factor(congenital_onset)
-)) +
-  ggplot2::geom_histogram(bins = 50, fill="slateblue") +
-  ggplot2::geom_vline(ggplot2::aes(xintercept=mean_severity_score_gpt),
-                      color="red") +
-  ggplot2::geom_label(data = unique(
-    dat2[,list(mean_severity_score_gpt), by="ancestor_name"]
-  ),
-  ggplot2::aes(x=mean_severity_score_gpt,
-               y=Inf,
-               label=round(mean_severity_score_gpt,2)),
-  color="red", size=3, hjust = 0, vjust = 1.5, alpha=.65) +
-  ggplot2::facet_wrap(facets = "ancestor_name~.", scales = "free_y",
-                      ncol = 3) +
-  ggplot2::labs(y="Frequency", x="Severity score") +
-  ggplot2::theme_bw() +
-  ggplot2::theme(strip.background = ggplot2::element_rect(
-    fill = "transparent"), 
-    text = element_text(size=16))
-```
+## Conclusion
 

### 2024-04-18 manuscript.qmd
add manuscript qmd

@@
   - rare disease 
-bibliography: bibliography.bib
+bibliography: references.bib
 ---
+
+```{r setup}
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(HPOExplorer)
+```
 
@@
 
-p1 <- ggplot2::ggplot(data = dat_top,
+p1 <- ggplot2::ggplot(data = dat_all,
                       ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
-  ggplot2::geom_tile() +
+  ggplot2::geom_tile(colour = "white", lwd = 0.5, linetype =1) +
   ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
-  ggplot2::theme_bw() +
+  scale_fill_brewer(palette = "GnBu", direction = -1) +
+  ggplot2::theme_classic() +
   ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
@@
                  text = element_text(size=16)) +
-  labs(x = "Annotation",
+  labs(x = "Clinical characteristic",
        y = "HPO term",
-       fill = "")
+       fill = "") +
+  ggplot2::facet_grid(position ~ ., scales = "free_y")
 
-# Heatmap of severity scores for top phenotypes
-p2 <- ggplot2::ggplot(data = dat_middle,
-                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
+p2 <- ggplot2::ggplot(data = dat_all,
+                      ggplot2::aes(x="severity_score_gpt", y=hpo_name,
+                                   fill=severity_score_gpt)) +
   ggplot2::geom_tile() +
   ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
-  ggplot2::theme_bw() +
-  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right', 
-                 text = ggplot2::element_text(size=16)) + 
-  ggplot2::labs(x = "Annotation",
-                y = "HPO term", 
-                fill = "")
+  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = "reverse") +
+  ggplot2::labs(x=NULL, y=NULL, fill = "Severity score") +
+  ggplot2::theme_void() +
+  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
+                 axis.ticks.y = ggplot2::element_blank(),
+                 legend.position = 'right') +
+  guides(fill = guide_colorbar(reverse = TRUE))
 
-# Heatmap of severity scores for top phenotypes
-p3 <- ggplot2::ggplot(data = dat_bottom,
-                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
-  ggplot2::geom_tile() +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
-  ggplot2::theme_bw() +
-  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right', 
-                 text = ggplot2::element_text(size=16)) + 
-  ggplot2::labs(x = "Annotation",
-                y = "HPO term", 
-                fill = "")
+p3 <- patchwork::wrap_plots(p1, p2, ncol = 2,
+                            widths = c(1,.2),
+                            guides = "collect")
 
-# Combine the plots
-p_combined <- patchwork::wrap_plots(p1, p2, p3, ncol = 3)
+print(p3)
 
-# Print the combined plot
-print(p_combined)
 ```

### 2024-04-18 _manuscript/manuscript.qmd
add manuscript files

--- 2024-04-18 manuscript.qmd
@@
 ```{r setup}
+#| echo: false
+#| warning: false
+
 library(dplyr)
@@
 #| warning: false
+#| message: false
 
@@
 # Stacked bar plot of clinical characteristic occurrence
-ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
+figure1 <- ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
   geom_bar(stat = "identity") +
@@
 ```{r annot_checks}
+#| echo: false
+#| message: false
+
 # Annotation checks
@@
 ```{r figure_2}
+#| echo: false
+#| message: false
 #| label: fig-checks
@@
 The labels above each bar indicate the number of phenotypes tested.
-#| fig-height: 4
-#| fig-width: 6
+#| fig-height: 7
+#| fig-width: 10
 consistency_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
@@
 ```{r severity_scoring}
+#| echo: false
+#| message: false
+
 # get all descendants of the term phenotypic abnormality 
@@
 ```{r figure_3}
+#| echo: false
+#| message: false
 #| label: fig-severity
@@
 The x-axis is scaled by the severity of the clinical characterisation.
-#| fig-height: 4
-#| fig-width: 6
+#| fig-height: 11
+#| fig-width: 10
 
@@
 | Clinical characteristic     | Always (4) | Often (3) | Rarely (2) | Never (0) |
+| --------------------------- | ---------- | --------- | ---------- | --------- |
 | Intellectual_disability (5) |     20     |     15    |    10      |     0     | 
@@
 ```{r supp_figure_1}
+#| echo: false
+#| message: false
+#| fig-cap: The distribution of phenotypic severity scores across various branches of the HPO.
+The histogram plots show the frequency of severity scores for all phenotypes  descending from the specified HPO branches.
+The x-axis represents the severity scores, and the y-axis indicates the count of phenotypes corresponding to each score.
+The average severity score for each HPO branch is highlighted in red.
+#| fig-height: 10
+#| fig-width: 10
 dat2 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)

### 2024-04-19 _manuscript/manuscript.qmd
BMS updates to GPT HPO manuscript

--- 2024-04-18 manuscript.qmd
@@
   - name: Kitty B Murphy
-    orcid: 
+    orcid: 0000-0002-8669-3076
     corresponding: true
@@
   - name: Brian M Schilder
-    orcid: 
-    corresponding: true
+    orcid: 0000-0001-5949-2191
+    corresponding: false
+    email: brian_schilder@alumni.brown.edu
     roles: 
@@
   - name: Nathan G Skene
-    orcid: 
+    orcid: 0000-0002-6807-3180
+    email: n.skene@imperial.ac.uk
     corresponding: true
@@
 keywords:
-  - HPO
-  - chatGPT
-  - rare disease 
+  - Ontology
+  - Human Phenotype Ontology
+  - Large Language Model
+  - GPT-4
+  - Artificial intelligence
+  - Generative AI
+  - Rare disease 
+  - Automation
+  - Medicine
 bibliography: references.bib
+date: last-modified
+citation:
+  container-title: medRxiv
 ---
 
-```{r setup}
+```{r setup, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
 library(dplyr)
 library(ggplot2)
+library(ggtext)
 library(patchwork)
 library(HPOExplorer)
+library(KGExplorer)
+library(data.table)
+```
+
+```{r import-data}
+#### Load HPO data
+tag <- "v2024-02-08"
+save_dir <- "data"
+keep_descendants <- "Phenotypic abnormality"
+p2g <- HPOExplorer::load_phenotype_to_genes(tag = tag,
+                                            save_dir=save_dir)
+hpo <- HPOExplorer::get_hpo(method="github",
+                            tag=tag, 
+                            save_dir=save_dir)
+hpo <- KGExplorer::filter_ontology(ont = hpo, 
+                                   keep_descendants = keep_descendants)
+#### Read HPO GPT annotations 
+gpt_annot <- HPOExplorer::gpt_annot_read(hpo = hpo)
+#### Compute IDs counts
+ids_matched <- intersect(c(hpo@terms,names(hpo@alternative_terms)), 
+                         gpt_annot$hpo_id)
+names_matched <- intersect(hpo@elementMetadata$name,
+                           trimws(gsub("obsolete","",gpt_annot$hpo_name, ignore.case = TRUE)))
+
+n_ids_hpo <- hpo@n_terms
+n_ids_gpt <- length(unique(fcoalesce(gpt_annot$hpo_name,gpt_annot$hpo_id)))
+```
+
+```{r gpt_annot_check}
+query_hits <- HPOExplorer::search_hpo(search_cols = c("name"),
+                                      hpo = hpo)
+#### Run annotation checks
+checks <- HPOExplorer::gpt_annot_check(annot = gpt_annot, 
+                                       query_hits = query_hits)
+# checks$true_pos_rate
+```
+
+```{r gpt_annot_codify}
+res_coded <- HPOExplorer::gpt_annot_codify(annot = gpt_annot)
+annot_melt <- HPOExplorer::gpt_annot_melt(res_coded = res_coded)
+```
+
+```{r gpt_annot_plot}
+gpt_annot_plot_out <- HPOExplorer::gpt_annot_plot(
+  annot = gpt_annot[hpo_name!=keep_descendants],
+  keep_descendants = keep_descendants, 
+  width = 10)
 ```
@@
 
-The Human phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised database for phenotypic abnormalities.
-However, there is currently no information pertaining to the severity of each phenotype, making systematic analyses and prioritisation of results difficult.
-With 18,082 abnormalities now corresponding to over 10,000 rare diseases, manual curation of such phenotypic annotations by experts would be labor-intensive and time-consuming.
-Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 model with Python to systematically annotate the severity of ~ 17,000 phenotypic abnormalities in the HPO.
-By checking that phenotypes with guaranteed outcomes were appropriately annotated, we demonstrate the potential for natural language processing technologies to automate the curation process accurately.
-For example, phenotypes such as "decreased male fertility" were used to compute a true positive rate, as they would be expected to be annotated as often, if not always, causing reduced fertility.
-Across the annotated outcomes, we observed > 73 % annotation accuracy.
+The Human Phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised framework for describing phenotypic abnormalities.
+However, documentation on the severity of each phenotype is currently very sparse, making systematic analyses and prioritisation of results challenging.
+With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming.
+Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 large language model (LLM) to systematically annotate the severity of all phenotypic abnormalities in the HPO.
+First, we benchmarked the generative LLM annotations against ground-truth labels within the HPO (e.g.
+phenotypes in the 'Cancer' HPO branch were annotating as causing cancer by GPT-4).
+True positive recall rates across different annotations ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity.
 Using a novel approach, we developed a severity scoring system that incorporates both the nature of the phenotype outcome and the frequency of its occurrence.
-These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human well being, and best targets for therapeutic intervention.
+These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.
 
@@
 Comprehensive annotation of phenotypic abnormalities is invaluable for defining, diagnosing, prognosing, and treating human disease.
-Since 2008, the Human phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes @Gargano2024-nj.
-Through developing open community resources, the depth and breadth of the HPO has continued to expand and there are now ~ 18,000 phenotypic abnormalities, corresponding to > 10,000 rare diseases, described.
+Since 2008, the Human Phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
+Through developing open community resources, the depth and breadth of the HPO has continued to expand and there are now \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases, described.
 In recent years, the HPO has expanded its disease annotations so that each HPO term can have metadata including typical age of onset and frequency.
-In addition, there are the Clinical modifier (put this in italics) and Clinical course (also italics) subontologies, which contains terms to describe factors including severity and triggers, and mortality and progression, respectively.
-Describing the severity-related attributes of a disease is crucial for attaining significant objectives in rare diseases.
+In addition, there are the *Clinical modifier* and *Clinical course* subontologies, which contains terms to describe factors including severity, triggers, mortality and progression, respectively.
+Describing the severity-related attributes of a disease is crucial for both research and clinical care of individuals with rare diseases.
 This includes enhancing diagnostic capabilities, as well as prioritising and guiding gene therapy trials.
 
-To date, the HPO has largely been manually curated by experts including clinicians, clinical geneticists, and researchers.
+To date, the HPO has largely relied on manual curation by experts including clinicians, clinical geneticists and researchers.
 Although this approach ensures the quality and accuracy of the ontology, it is time-consuming and labour-intensive.
 As a result, less than 1% of terms within the HPO have metadata pertaining to their features such as time course and severity.
-As artificial intelligence (AI) capabilities advance, there is an opportunity to integrate natural language processing technologies into assisting in the curation process.
-Here, we have used the OpenAI GPT-4 model (https://openai.com/) with Python to systematically annotate the severity of > 17,000 phenotypic abnormalities within the HPO.
-Our severity annotation framework was developed based on previously defined criteria developed through consultation with clinicians 2.
+Artificial intelligence (AI) capabilities have advanced considerably in recent years, presenting new opportunities to integrate natural language processing technologies into assisting in the curation process.
+Specifically, there have recently been considerable advances in large language model (LLM) and their application to biomedical problems, in some cases performing as well or better than human clinicians on standardised medical exams and patient diagnosis tasks [@vanveenAdaptedLargeLanguage2024; @boltonBioMedLM7BParameter2024; @zhangBiomedGPTUnifiedGeneralist2023; @labrakBioMistralCollectionOpenSource2024; @guDomainSpecificLanguageModel2021; @singhalLargeLanguageModels2023; @shinBioMegatronLargerBiomedical2020; @chengExploringPotentialGPT42023; @singhalLargeLanguageModels2023; @oneilPhenomicsAssistantInterface2024; @luoBioGPTGenerativePretrained2022; @mcduffAccurateDifferentialDiagnosis2023; @singhalExpertLevelMedicalQuestion2023].
+Recent work has demonstrated that the Generative Pre-trained Transformer 4 (GPT-4) foundation model [@openaiGPT4TechnicalReport2024], when combined with strategic prompt engineering, can outperform even specialist LLMs that are explicitly fine-tuned for biomedical tasks [@noriCanGeneralistFoundation2023].
+In a landmark achievement, GPT-4 was the first LLM to surpass a score of 90% in the United States Medical Licensing Examination (USML) [@noriCanGeneralistFoundation2023].
+
+Here, we have used GPT-4 to systematically annotate the severity of `r n_ids_gpt` / `r n_ids_hpo` (`r round(n_ids_gpt/n_ids_hpo*100,1)`%) phenotypic abnormalities within the HPO.
+Our severity annotation framework was adapted from previously defined criteria developed through consultation with clinicians [@Lazarin2014-rz].
 The authors consulted 192 healthcare professionals for their opinions on the relative severity of various clinical characteristics: they used this to create a system for categorising the severity of diseases.
 Briefly, each healthcare professional was sent a survey asking them to first rate how important a disease characteristic was for determining disease severity, and then to rate the severity of a set of given disease.
-Using the responses, the authors were able to categorise clinical characteristics into 4 ‘severity tiers’.
+Using the responses, the authors were able to categorise clinical characteristics into 4 'severity tiers'.
 While characteristics such as shortened lifespan in infancy and intellectual disability were identified as highly severe and placed into tier 1, sensory impairment and reduced lifespan were categorised as less severe and placed into tier 4.
-Being able to quickly ascribe severity measures based on those criteria to HPO phenotypes, will assist with interpreting phenome-wide studies.
+Standardised metrics of severity allow clinicians to quickly assess the urgency of treating a given phenotype, as well as prognosing what outcomes might be expected.
 
-Almost 800 phenotypes were annotated twice to evaluate annotation consistency, and a true positive rate of annotations was calculated to assess annotation accuracy.
+`r checks$consistency_count[[1]]` phenotypes were annotated multiple times to evaluate the consistency of responses generated by GPT-4.
+For a subset of phenotypes with known expected annotations, true positive rates were calculated to assess recall.
 Additionally, based on the clinical characteristics and their occurrence, we have quantified the severity of each phenotype, providing an example of how these clinical characteristic annotations can be used to guide prioritisation of gene therapy trials.
-Ultimately, we hope that our resource will be of utility to those working in rare diseases, as well as the wider rare disease community.
+Ultimately, we hope that our resource will be of utility to those working in rare diseases, as well as the wider healthcare community.
 
@@
 
-### Annotating the HPO using OpenAI GPT-4
-We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO).
-Our annotation framework was developed based on previously defined criteria for classifying disease severity @Lazarin2014-rz.
-We sought to evaluate the impact of phenotypes on outcomes including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
-Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
-Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance was achieved when inputting no more than two or three phenotypes per prompt.
-An example prompt can be seen in Figure 1.
-The phenotype outcome occurrence varied by category.
-While > 50% of annotated phenotypes never caused blindness, cancer, immunodeficiency, intellectual disability, reduced fertility, and sensory impairment, a similar proportion often or always had congenital onset (Fig.
-1).
+### Annotating the HPO using GPT-4
 
-```{r occurence_plot}
-#| echo: false
-#| warning: false
-
-# read in hpo gpt annotations 
-hpo_annot <- HPOExplorer::gpt_annot_read()
+```{r fig-occurrence}
+#| label: fig-occurrence
+#| fig-cap: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics.
+**a**, An example of the prompt input given to to GPT-4.
+The phenotypes listed in the second to last sentence (*italicised*) were changed to allow all HPO phenotypes to be annotated.
+**b**, Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all annotated HPO phenotypes.
+The terms shown on the x-axis are the clinical characteristics for which GPT-4 was asked to determine whether each phenotype caused them.
+#| fig-height: 7
+#| fig-width: 8
 
 # subset to clinical characteristics 
-occurr_df <- hpo_annot[,c("hpo_name", "intellectual_disability", "death", "impaired_mobility", "blindness",
+occurr_df <- gpt_annot[,c("hpo_name", "intellectual_disability", "death", "impaired_mobility", "blindness",
                           "sensory_impairments", "immunodeficiency", "cancer", "reduced_fertility",
@@
 
-# Loop through each column of interest and count the occurrences of 'always', 'often', 'rarely', 'never'
+## Loop through each column of interest and count the occurrences of:
+## 'always', 'often', 'rarely', 'never'
 for (col in cols_of_interest) {
@@
 
-count_df_long <- tidyr::gather(count_df, key = "condition", value = "count", always:never)
+count_df_long <- tidyr::gather(count_df,
+                               key = "condition",
+                               value = "count", 
+                               always:never)
 count_df_long$category <- rownames(count_df)
@@
 count_df_long$category <- gsub("_", " ", count_df_long$category)
-count_df_long$condition <- factor(count_df_long$condition, levels=c("always", "often", "rarely", "never"))
+count_df_long$condition <- factor(count_df_long$condition, 
+                                  levels=c("always", "often", "rarely", "never"))
 
-count_df_long <- count_df_long %>%
-  dplyr::group_by(category) %>%
+count_df_long <- count_df_long |>
+  dplyr::group_by(category) |>
   dplyr::mutate(percentage = count / sum(count) * 100)
@@
                       "immunodeficiency", "cancer", "reduced fertility")
+count_df_long$category <- factor(count_df_long$category, 
+                                 levels = annotation_order)
 
-count_df_long$category <- factor(count_df_long$category, levels = annotation_order)
+# Load the prompt generation function 
+source("code/prompt_gen.R")
+prompts <- prompt_gen_table(hpo = hpo)
+txt_prompt <- prompts$prompt[1688]
+for(x in prompts$terms[[1688]]){
+  txt_prompt <- sub(x,paste0("*",x,"*"),txt_prompt)
+}
+for(x in c(
+  eval(formals(prompt_gen_table)$effects),
+  eval(formals(prompt_gen_table)$responses),
+  "justification")){
+  txt_prompt <- sub(x,paste0("**",x,"**"),txt_prompt)
+}
+## Subplot a: Example prompt
+f1a <- ggplot() + 
+  labs(title="Example GPT-4 prompt") +
+  ggtext::geom_textbox(aes(x=.5,y=1),
+                       box.padding = margin(20, 10, 20, 10),
+                       box.margin = margin(0, 0, 0, 0),
+                       box.r = grid::unit(30, "pt"),
+                       width = unit(.8, "npc"),
+                       halign=.5,
+                       fill="grey90",
+                       label=txt_prompt) +
+  theme_void() +
+  theme(plot.title = element_text(hjust=0.5))
 
-# Stacked bar plot of clinical characteristic occurrence
-ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
+## Subplot b: Stacked bar plot of clinical characteristic occurrence.
+f1b <- ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
   geom_bar(stat = "identity") +
@@
   theme_minimal() +
-  theme(legend.position = "right", plot.title = element_text(hjust=0.5), 
-        text = element_text(size=16),
-        axis.text.x = element_text(angle=45, size=16, hjust=1), 
-        title = element_text(size=15)) +
+  theme(legend.position = "right", 
+        plot.title = element_text(hjust=0.5),  
+        axis.text.x = element_text(angle=45, hjust=1),  
+        ) +
   scale_fill_brewer(palette = "GnBu", direction = -1)
 
+(
+  f1a| 
+    (f1b/plot_spacer() ) + plot_layout(heights = c(1, .1))
+) + 
+  patchwork::plot_layout(widths = c(2,1)) +
+  patchwork::plot_annotation(tag_levels = letters) 
 ```
 
-![Figure 1: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics.
-a An example of the prompt input given to to GPT-4.
-The phenotypes listed in the last sentence were changed to allow ~17,000 phenotypes to be annotated.
-b Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all ~17,000 phenotypes annotated.The terms shown on the X-axis are the clinical characteristics, for which GPT-4 was asked to determine whether each phenotype caused them.
-For example, it annotated ‘Respiratory arrest’ as always causing death.
-Only 75 phenotypes were annotated as always causing death, despite 9,544 phenotypes often or rarely causing it.](figures/clinical_characteristic_occurrence.png)
-### Annotation consistency and accuracy
-To assess annotation consistency, 793 phenotypes underwent duplicate annotations, with two metrics employed to determine the 'consistency rate'.
+We employed the OpenAI GPT-4 model with Python to annotate `r length(unique(gpt_annot$hpo_name))` terms within the HPO (`r tag`) [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
+Our annotation framework was developed based on previously defined criteria for classifying disease severity [@Lazarin2014-rz].
+We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
+Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
+For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
+
+```{r always_death}
+never_thresh <- 50
+never_dt <- subset(count_df_long, condition=="never" & percentage>=never_thresh)
+always_death <- gpt_annot[,list(always_death=all(death=="always"),
+                                hpo_name=hpo_name[1]),
+                          by="hpo_id"][always_death==TRUE,]
+```
+
+Phenotype outcome occurrence varied across annotation categories.
+\>`r never_thresh`% of phenotypes never caused `r paste(never_dt$category[-1], collapse=", ")` or `r never_dt$category[1]`.
+Only a minority of phenotypes (`r round(sum(subset(count_df_long, condition %in% c("never") & category=="congenital onset")$percentage),1)`%) never had a congenital onset, which is expected as rare disorders tend to be early onset genetic conditions [Fig.
+@fig-occurrence].
+
+Less than `r ceiling(subset(count_df_long, condition=="always" & category=="death")$percentage)`% of phenotype always directly resulted in death, such as 'Stillbirth', 'Anencephaly' and 'Bilateral lung agenesis'.
+Only `r uniqueN(gpt_annot[death=="always"]$hpo_name)` phenotypes were annotated as always causing death, while `r uniqueN(gpt_annot[death %in% c("often","rarely")]$hpo_name)` phenotypes were annotated as often or rarely causing death.
+`r uniqueN(gpt_annot[death %in% c("never")]$hpo_name)` phenotypes were annotated as never causing death.
+This included `r sum(grepl("syndactyly",gpt_annot[death %in% c("never")]$hpo_name, ignore.case = TRUE))` unique forms of syndactyly, a non-lethal condition that causes fused or webbed fingers (occurring 1 in 1,200–15,000 live births).
+While not life-threatening itself, syndactyly is a symptom of genetic disorders that can cause life-threatening cardiovascular and neurodevelopmental defects, such as Apert Syndrome [@garagnaniSyndromesAssociatedSyndactyly2013].
+This example highlights the ability of GPT-4 to successfully distinguish between phenotypes that directly cause lethality, and those that are often associated with diseases that cause lethality.
+
+### Annotation consistency and recall
+
+To assess annotation consistency, `r checks$consistency_count[[1]]` phenotypes underwent duplicate annotations.
+We employed two different metrics to determine the *consistency rate*.
 The first, less stringent metric, defined consistency as the duplicate annotations being either 'always' and 'often', or 'never' and 'rarely'.
@@
 'always' and 'always'.
-For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least 80%, and for the more stringent metric, the lowest consistency rate was 57%, for congenital onset.
-An example of where annotations were inconsistent was for the HPO term ‘Acute leukaemia’.
-One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that ‘weakness and fatigue from leukaemia and its treatment can impair mobility’.
-The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that ‘acute leukaemia rarely impairs mobility directly’.
+For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least `r floor(min(checks$consistency_rate)*100)`%, and for the more stringent metric, the lowest consistency rate was `r round(min(checks$consistency_stringent_rate)*100)`% for congenital onset.
+An example of where annotations were inconsistent was for the HPO term 'Acute leukaemia'.
+One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that 'weakness and fatigue from leukaemia and its treatment can impair mobility'.
+The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that 'acute leukaemia rarely impairs mobility directly'.
 Despite specifying in the prompt for GPT-4 not to take into consideration indirect effects, here is an example of where it failed to do so.
@@
 ```{r annot_checks}
-# Annotation checks
-checks <- HPOExplorer::gpt_annot_check()
-
 # Define items to be checked
@@
            "true_pos_count", "true_pos_rate")
-
 metric_types <- c("Rate", "Count")[1]
@@
 check_df$metric_category <- gsub("_", " ", check_df$metric_category)
-
-metric_categories <- c("consistency", "consistency stringent")
-
+check_df[metric_category=="consistency",metric_category:="consistency lenient"]
 ```
 
-```{r figure_2}
+```{r fig-checks}
 #| label: fig-checks
 #| fig-cap: GPT-4 annotations are consistent and accurate across annotations.
-a Barplot showing the annotation consistency for 793 phenotypes that were annotated twice.
-On the left hand side, annotations were defined as consistent if the given phenotype was annotated to the same outcome occurrence both times, or if it was annotated to ‘always’ and ‘often’, or, ‘never’ and ‘rarely.
-On the right hand side, annotations were only defined as consistent if the given phenotype was annotated to the same outcome occurrence both times.
-b Bar plot of the true positive rate for each annotation.
-The true positives were determined based on the assumption that all descendants of some HPO phenotypes should be related to those outcomes.
-For instance, all descendents of the HPO term for “Neoplasm” were considered to be true positives for cancer (an assumption which is not always the case, as GPT-4 detects, as some neoplasms are benign).
+**a**, Barplot showing the annotation consistency within phenotypes that were annotated more than once.
+In the lenient metric, annotations were collapsed into two groups ('always'/'often' and 'never'/'rarely').
+For a given clinical characteristic within a given phenotype, if an annotation was always within the same group it was considered consistent.
+In the stringent metric, all four annotation categories were considered to be different from one another.
+Thus, annotations were only defined as consistent if they were all identical.
+The blue dashed line indicates the probability of two annotations being consistent by chance in the lenient metric (~1/2).
+The gold dashed line is the probability of two annotations being consistent by chance in the stringent metric (~1/4).
+**b**, Bar plot of the true positive rate for each annotation.
 The labels above each bar indicate the number of phenotypes tested.
-#| fig-height: 4
+#| fig-height: 5
 #| fig-width: 6
-consistency_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
+
+lab_size <- 3
+consistency_plot <- ggplot(check_df[metric_type %in% metric_types &
+                                    grepl("consistency",metric_category)],
+                           aes(x = annotation, y = value,
+                               fill = stringr::str_to_sentence(
+                                 gsub("consistency ", "", metric_category)
+                                 ),
+                               label = round(value, 2))) +
+  geom_bar(stat = "identity",position = "dodge") +
+  labs(title="Annotation consistency",
+       x = NULL, y = "Mean consistency", 
+       fill=NULL) +
+  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1),
+        legend.position = "top", 
+        strip.background = element_rect(fill="white"),
+        plot.title = element_text(hjust = 0.5)) +
+  geom_text(aes(x = 1, y = 1, label = paste0("n=",n," phenotypes")),
+            size=lab_size,
+            check_overlap=TRUE, hjust=0) +
+  ## Compute the empirical probability of annotations being consistent by chance 
+  ## with 2 options and N duplicates
+  geom_hline(yintercept = mean(1/2^checks$annot$pheno_count), 
+             color="dodgerblue", linetype="dashed")+
+  ## Compute the empirical probability of annotations being consistent by chance 
+  ## with 4 options and N duplicates
+  geom_hline(yintercept = mean(1/4^checks$annot$pheno_count), 
+             color="gold3", linetype="dashed")
+
+recall_plot <- ggplot(check_df[metric_type %in% metric_types &
+                               metric_category %in% c("true pos"), ],
                         aes(x = annotation, y = value,
                             fill = metric_category,
-                               label = round(value, 2))) +
-  geom_bar(stat = "identity", show.legend = TRUE) +
-  facet_grid(facets = ~metric_category, scales = "free") +
-  theme_bw() +
-  labs(x = "", y = "Value") +
-  ggtitle("Annotation consistency") +
-  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
-        legend.position = "none",
-        text = element_text(size = 16),
-        title = element_text(size = 15),
-        plot.title = element_text(hjust = 0.5)) +
-  geom_text(aes(x = 7.5, y = 1.05, label = "N phenotypes = 793"), size = 4)
-
-metric_categories <- "true pos"
-
-accuracy_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
-                        aes(x = annotation, y = value,
-                            fill = metric_category,
-                            label = n)) +
+                            label = n#paste0("n\n=\n",n)
+                        )
+                      ) +
   geom_bar(stat = "identity", show.legend = TRUE) +
   theme_bw() +
-  labs(x = "", y = "Value") +
+  labs(x = NULL, y = "Recall") +
   ggtitle("True positive rate") +
   scale_fill_manual(values = c("#B1D5BBFF")) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
+  theme(axis.text.x = element_text(angle = 45, hjust = 1),
         legend.position = "none", 
-        text = element_text(size = 16),
-        title = element_text(size = 15),
         plot.title = element_text(hjust = 0.5)) +
-  geom_label(fill = "white", size = 4)
+  geom_label(fill = alpha("white",1),
+            size = lab_size)
 
-consistency_plot + accuracy_plot + plot_layout(widths = c(1, 0.7)) +
-  plot_annotation(tag_levels = 'a') & theme(plot.tag = element_text(size = 18, face = "bold"))
+consistency_plot + recall_plot + 
+  plot_layout(widths = c(1, 1)) +
+  plot_annotation(tag_levels = 'a')
 ```
 
-In order to evaluate the accuracy of the annotations, we calculated a true positive rate.
+In order to evaluate the validity of the annotations, we calculated a true positive rate.
 This involved identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the presence of certain conditions.
-For instance, the phenotypes 'decreased fertility in females' and 'decreased fertility in males' should, in the annotations, often, if not always, cause reduced fertility.
-We observed an encouraging true positive rate exceeding 70% across all phenotype outcomes.
-The lowest rate was observed for cancer (73%), for which all descendants of the HPO term ‘Neoplasm’ were considered to be true positives.
-However, this assumption, as GPT-4 detects, is not always true as these abnormal tissue/cell growths can be either cancerous or benign.
-For example, the HPO term ‘multiple digital exostoses’, which are benign bone tumours, is a descendant of the HPO term ‘Neoplasm’, and GPT-4 annotated it as rarely causing cancer, giving the justification that ‘Bone growths may rarely become cancerous’.
-This high level of accuracy underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information.
-However, we acknowledge that the number of testable phenotypes for some of these categories are low.
-Furthermore, some of the ‘true positive’ phenotypes are synonymous to the clinical characteristic itself, for instance, for testing the true positive rate of the characteristic ‘Intellectual disability’, the phenotype ‘Severe intellectual disability’ was used.
+For instance, the phenotypes 'Decreased fertility in females' and 'Decreased fertility in males' should often or always cause reduced fertility.
+We observed an encouraging true positive rate exceeding `r floor(min(checks$true_pos_rate)*100)`% across in every clinical characteristic and achieving perfect recall (100%) in `r sum(checks$true_pos_rate==1)`/`r length(checks$true_pos_rate)` characteristics.
+
+```{r}
+lowest_recall <- names(checks$true_pos_rate)[checks$true_pos_rate==min(checks$true_pos_rate)]
+lowest_recall_dt <- checks$annot[get(lowest_recall)=="never" & hpo_id %in% query_hits[[lowest_recall]]]
+```
+
+The lowest true positive rate was observed for `r gsub("_"," ",lowest_recall)`, with `r round(checks$true_pos_rate[[lowest_recall]]*100,1)`% recall across `r checks$true_pos_count[[lowest_recall]]` HPO phenotypes.
+Some cases in which the GPT-4 annotations disagreed with the HPO ground truth included: `r paste(shQuote(head(lowest_recall_dt$hpo_name, 6)), collapse=", ")`.
+In the case of `r shQuote(lowest_recall_dt$hpo_name[1])` it provided the justification that `r shQuote(lowest_recall_dt$physical_malformations_justification[1])`.
+In another instance, GPT-4 noted that `r shQuote(lowest_recall_dt$hpo_name[2])` is `r shQuote(lowest_recall_dt$physical_malformations_justification[2])`.
+This indicates that while technically incorrect according to our predefined benchmarks, a case could in fact be made that mild skin conditions do not rise to the level of physical malformations.
+
+This high level of recall underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information.
+However, we acknowledge that the number of testable true positive phenotypes for some of these categories are low, especially `r shQuote(names(checks$checkable_count)[checks$checkable_count==min(checks$checkable_count)])` for which there is only `r min(checks$checkable_count)` phenotype in the HPO (after excluding terms pertaining to colour or night blindness).
+Furthermore, some of the true positive phenotypes are lexically similar to the name of the clinical characteristic itself.
+In these cases, annotating 'Severe intellectual disability' as always causing intellectual disability is a relatively trivial task.
+Nevertheless, even these scenarios provide a clear and interpretable benchmark for the model's performance.
+In addition, were numerous phenotypes with lexically non-obvious relationships to the clinical characteristic that were annotated correctly by GPT-4.
+For example, 'Molar tooth sign on MRI' was correctly annotated as causing intellectual disability.
 
 ### Quantifying phenotypic severity
+
+```{r anenceph_dt}
+anenceph_dt <- res_coded$annot_weighted[hpo_name=='Anencephaly',]
+```
+
 Quantifying the severity of phenotypes can have important implications for diagnosis, prognosis, and treatment, for instance, guiding prioritisation of gene therapy trials for the most severe phenotypes.
 First, we created a dictionary to map each phenotype outcome (e.g.
-blindness) and its response (always, often, rarely, never) to numeric values from 0-4.
+blindness) and its response (always, often, rarely, never) to numeric values from 0-3.
 Then, the phenotype outcome values were multiplied by their associated response values.
 Importantly, the values reflected the severity of each outcome based on both the outcome itself and the frequency of the response.
-For instance, a phenotype always causing death would have a higher multiplied value than a phenotype rarely causing reduced fertility.
+For instance, a phenotype always causing death would have a higher multiplied value than a phenotype rarely causing reduced fertility (see @tbl-metric-weights).
 Next, we computed an average score for each phenotype by aggregating the multiplied values across all phenotype outcomes and then calculating the mean.
@@
 Based on these scores we evaluated the top 50 severe phenotypes.
-The most severe phenotype, with a score of 58, was ‘Anencephaly’ (HP:0002323).
+One of the most severe phenotype was `r shQuote(anenceph_dt$hpo_name)` (`r anenceph_dt$hpo_id`) with a composite severity score of `r round(anenceph_dt$severity_score_gpt, 1)`.
 Anencephaly is a birth defect where the baby is born without a portion of its brain and skull, often these babies are stillborn.
 In fact, many of the most severe phenotypes were related to developmental brain and neural tube defects.
-Comparison of the severity scores for each response, across the phenotype outcomes annotated, revealed consistent trends: as the response of the phenotype outcome increased (from never to always), the severity score also increased (Supplementary Fig.
-1).
-We also evaluated the severity score distribution by HPO branch and calculated the mean severity score using all phenotypes within a HPO branch (Fig.
-4).
-After the term ‘Phenotypic abnormality’, the highest severity score was for the HPO branch ‘Abnormal cellular phenotype’, followed by ‘Abnormality of prenatal development or birth’, which would include the high severity scoring phenotypes seen in Fig.
-3.
+Comparison of the severity scores for each response, across the phenotype outcomes annotated, revealed consistent trends: as the response of the phenotype outcome increased (from never to always), the severity score also increased ([Supplementary Fig.
+@fig-severity-boxplot]).
+We also evaluated the severity score distribution by HPO branch and calculated the mean severity score using all phenotypes within each major HPO branch ([Fig.
+@fig-severity-histo]).
+The HPO branch with the greatest mean severity score was `r shQuote(unique(gpt_annot_plot_out$data$dat2$ancestor_name)[1])` (mean=`r round(unique(gpt_annot_plot_out$data$dat2$mean_severity_score_gpt)[1],1)`), followed by `r shQuote(unique(gpt_annot_plot_out$data$dat2$ancestor_name)[2])` (mean=`r round(unique(gpt_annot_plot_out$data$dat2$mean_severity_score_gpt)[2],1)`), which would include the highly ranked phenotypes seen in @fig-top-phenos.
 
 ```{r severity_scoring}
+## code adapted from HPOExplorer::gpt_annot_plot()
 # get all descendants of the term phenotypic abnormality 
-keep_descendants="Phenotypic abnormality"
 keep_ont_levels=seq(3,17)
 
-ancestor_name <- variable <- hpo_id <- hpo_name <-
-  value <- severity_score_gpt <- mean_severity_score_gpt <- NULL;
+#### Filter out ont levels ####
+annot_melt <- HPOExplorer::add_ancestor(annot_melt,
+                                        keep_descendants = keep_descendants)
+data.table::setorderv(annot_melt,"severity_score_gpt",-1)
 
-#### Prepare annotation results ####
-gpt_annot <- HPOExplorer::gpt_annot_read()
-res_coded <- HPOExplorer::gpt_annot_codify(annot = gpt_annot)
-dat1 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
-
-#### Filter out ont levels ####
-dat1 <- add_ancestor(dat1,
-                     keep_descendants = keep_descendants)
-data.table::setorderv(dat1,"severity_score_gpt",-1)
-
-dat1 <- HPOExplorer::add_ont_lvl(dat1, keep_ont_levels = keep_ont_levels)
-dat1$variable <- gsub("_", " ", dat1$variable)
+annot_melt <- HPOExplorer::add_ont_lvl(annot_melt, keep_ont_levels = keep_ont_levels)
+annot_melt$variable <- gsub("_", " ", annot_melt$variable)
 
 # get all unique phenotypes
-total_unique <- length(unique(dat1$hpo_id))
+total_unique <- length(unique(annot_melt$hpo_id))
 
 # get the top N phenotypes
-top_n=15
-dat_top <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(top_n)]]
+top_n <- 15
+dat_top <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(top_n)]]
 
@@
 bottom_n <- 15
-dat_bottom <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(total_unique - bottom_n + 1, total_unique)]]
+dat_bottom <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(total_unique - bottom_n + 1, total_unique)]]
 
@@
 # Get phenotypes from the middle
-dat_middle <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(start_index, end_index)]]
+dat_middle <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(start_index, end_index)]]
 
@@
 dat_all <- rbind(dat_top, dat_middle, dat_bottom)
-dat_all$position <- factor(dat_all$position, levels=c("Top", "Middle", "Bottom"))
-
+dat_all$position <- factor(dat_all$position, levels=unique(dat_all$position))
 dat_all$variable <- factor(dat_all$variable, levels = annotation_order)
@@
 
-```{r figure_3}
-#| label: fig-severity
+```{r fig-top-phenos}
+#| label: fig-top-phenos
 #| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions.
 Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score.
-Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic (death = 5, intellectual disability = 5, impaired mobility = 4, blindness = 4, congenital onset = 4, physical malformations = 3, sensory impairments = 3, immunodeficiency = 3,  cancer = 3, reduced fertility = 1) by their corresponding response frequencies (Always = 4, Often = 3, Rarely = 2, Never = 0).
+Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights.
 The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score.
 The x-axis is scaled by the severity of the clinical characterisation.
-#| fig-height: 4
-#| fig-width: 6
+#| fig-height: 7.5
+#| fig-width: 8
 
@@
   ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right',
-                 text = element_text(size=16)) +
+                 legend.position = 'right') +
   labs(x = "Clinical characteristic",
-       y = "HPO term",
-       fill = "") +
+       y = "HPO phenotype",
+       fill = "Occurrence") +
   ggplot2::facet_grid(position ~ ., scales = "free_y")
@@
   scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = "reverse") +
-  ggplot2::labs(x=NULL, y=NULL, fill = "Severity score") +
+  ggplot2::labs(x=NULL, y=NULL, fill = "Severity\nscore") +
   ggplot2::theme_void() +
+  guides(fill = guide_colorbar(reverse = TRUE)) +
   ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                  axis.ticks.y = ggplot2::element_blank(),
-                 legend.position = 'right') +
-  guides(fill = guide_colorbar(reverse = TRUE))
+                 legend.position = 'right')
 
@@
 
-print(p3)
-
+methods::show(p3)
 ```
@@
 
-Here, we present a novel approach leveraging the OpenAI GPT-4 model, to systematically annotate the severity of 17,000 phenotypic abnormalities within the HPO.
-Our findings highlight the potential of natural language processing technologies in significantly contributing to the automation and refinement of the curation process in genomics and rare disease research.
-Manual curation of ontologies has traditionally relied on expert input, a process that is both time-consuming and labor-intensive.
-By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without compromising accuracy.
-Our validation approach yielded a high true positive rate exceeding 73% across the phenotypes tested.
+The creation and annotation of biomedical knowledge has traditionally relied on manual or semi-manual curation by human experts [@putmanMonarchInitiative20242024; @ochoaOpenTargetsPlatform2021; @mungallMonarchInitiativeIntegrative2017a; @kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
+Performing such manual curation and review tasks at scale is often infeasible for human biomedical experts given limited time and resources.
+LLMs have the capacity to effectively encode, retrieve, and synthesise vast amounts of diverse information in a highly scalable manner [@vanveenAdaptedLargeLanguage2024; @singhalLargeLanguageModels2023; @openaiGPT4TechnicalReport2024].
+This makes them powerful tools that can be applied in a rapidly expanding variety of scenarios, including medical practice, research and data curation [@singhalLargeLanguageModels2023; @toroDynamicRetrievalAugmented2023; @panLargeLanguageModels2023; @oneilPhenomicsAssistantInterface2024; @caufieldStructuredPromptInterrogation2023].
+
+Here, we introduce a novel framework to leverage the current best-in-class LLM, GPT-4 [@openaiGPT4TechnicalReport2024], to systematically annotate the severity of `r n_ids_gpt` phenotypic abnormalities within the HPO.
+By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without substantially compromising accuracy.
+Our validation approach yielded a high true positive rate exceeding `r floor(min(checks$true_pos_rate)*100)`% across the phenotypes tested.
 Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.
 
-A key contribution of our study is the development of a severity scoring system that integrates both the nature of the phenotype outcome and the frequency of its occurrence.
-By quantifying the phenotypic severity this way, we highlight the most impactful conditions and provide a comprehensive framework for prioritising gene therapy trials and guiding clinical decision-making in rare diseases.
+Throughout this study, we observed that GPT-4 was capable of reliably recovering deep semantic relationships from the medical domain, far beyond making superficial inferences based on lexical similarities.
+An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt`), which GPT-4 annotated as causing intellectual disability.
+At first glance, we ourselves assumed this was a false positive as the term appeared to be related to dentition.
+However, upon further inspection we realised that molar tooth sign is in fact a pattern of abnormal brain morphology that happens to bear some resemblance to molar dentition when observed in radiological scans.
+This phenotype is a known sign of neurodevelopmental defects that can indeed cause severe intellectual disability [@gleesonMolarToothSign2004].
 
-** Brian to pick examples from severity scoring **
+In addition to rapidly synthesising and summarising vast amounts of information, LLMs can also be steered to provide justifications for each particular response.
+This makes LLMs amenable to direct interrogation as a means of recovering explainability, especially when designed to retain information about previous requests and interactions as they use these to iteratively improve and update their predictions [@janikAspectsHumanMemory2024].
+This represents a categorical advance over traditional natural language processing models based on more shallow forms of statistical or machine learning (e.g.
+Term Frequency-Inverse Document Frequency [@jonesStatisticalInterpretationTerm1972], Word2vec [@mikolovEfficientEstimationWord2013]) which lack the ability to provide chains of causal reasoning to justify their predictions.
+This highlights the fundamental trade-off between simpler models with high explainability (the ability humans to understand the inner workings of the model) but low interpretability (the ability of humans to trace the decision process of the model, analogous to human 'reasoning'), and deeper more complex models with low explainability but high interpretability [@marcinkevicsInterpretabilityExplainabilityMachine2023].
 
-Given its scalability it can be easily applied as more phenotypes are added to the HPO, but can also be extended to include different annotations and to be used with other ontologies.
-For instance, we could have also tried to use GPT-4 to annotate how prevalent a phenotype is in the human population, or what the cost to the health system is of a phenotype.
-Currently there are no systematic annotations of these on a phenotypic level.
-However, it’s unclear whether GPT-4 could currently be applied for such annotations, or whether they require much more extensive data collection.
+A key contribution of our study is the introduction of a quantitative severity scoring system that integrates both the nature of the phenotype outcome and the frequency of its occurrence.
+By encoding the concept of severity in this way, we are able to prioritise phenotypes based on their impact on patients.
+The methodology allowed us to transition from low-throughput qualitative assessments of severity (e.g.
+@Lazarin2014-rz) to high-throughput quantitative assessments of severity.
+One of the most severe phenotypes in the HPO is 'Fetal akinesia sequence' (FAS; `r annot_melt[hpo_name=='Fetal akinesia sequence']$hpo_id[1]`, severity score= `r round(annot_melt[hpo_name=='Fetal akinesia sequence']$severity_score_gpt[1],1)`), and extremely rare condition that is almost always lethal.
+FAS is a complex, multi-system phenotype that can be caused by at least `r uniqueN(p2g[hpo_name=='Fetal akinesia sequence']$disease_id)` different genetic disorders.
+Despite the complex and heterogenous aetiology of this phenotype, GPT-4 was able to provide accurate annotations alongside explainable justifications for those annotations (see @tbl-fas).
+For example, this phenotype almost always results in death, either *in utero* or shortly after birth.
+Not only did GPT-4 correctly provide the annotation death as `r shQuote(gpt_annot[hpo_name=='Fetal akinesia sequence']$death)`, when asked whether FAS causes sensory impairments it provided the response `r shQuote(gpt_annot[hpo_name=='Fetal akinesia sequence']$sensory_impairments)` with the justification `r shQuote(gpt_annot[hpo_name=='Fetal akinesia sequence']$sensory_impairments_justification)` Neurodevelopmental disruption is indeed a hallmark component of FAS (e.g.
+hydrocephalus, cerebellar hypoplasia) that causes severe impairments across multiple sensory systems [@chenPrenatalDiagnosisGenetic2012].
+This demonstrates that GPT-4 was able to recover the correct chain of causality from phenotype to clinical outcome.
 
-Importantly, there are additional factors that need to be considered when trying to prioritise phenotypes for their suitability for gene therapy development.
-First, our scoring is in some sense subjective, for instance, one could ask is something that always causes death, always worse than something which causes a lifetime of other severe clinical characteristics.
-In terms of viability for a clinical trial, a phenotype which causes a lifetime of expensive burden on the medical system may be a better target than a phenotype which causes a quick death with little or no other clinical characteristics before.
-It would be worth exploring whether GPT-4 could estimate such costs.
-Another factor that affects the viability of developing a treatment for a therapy is how quickly a clinical trial can be run.
-For instance, measuring a high risk of respiratory failure over ten years would require a long trial.
-However, testing for total reversal of an existing severe phenotype would be much faster.
-As AI capabilities continue to advance, perhaps this is something GPT-4 and other models of the like could test for.
+Our findings highlight the potential of this next generation of natural language processing technologies in significantly contributing to the automation and refinement of data curation in biomedical research.
+These results have a large number of useful real-world applications, such as prioritising gene therapy candidates [@murphyIdentificationCellTypespecific2023] and guiding clinical decision-making in rare diseases.
+As the HPO continues to grow over time, this approach can easily be repeated to keep pace with the rapidly evolving biomedical landscape.
+Furthermore, this approach can be extended to produce different sets of annotations or be used with any other ontology.
+Additional use cases include gathering data on the prevalence of each phenotype to approximate their social and financial costs.
 
-An additional annotation that we didn’t take into consideration here is that GPT-4 might be aware of whether there are existing treatments for these phenotypes.
-Therefore, some of the top-ranking severe phenotype we have identified may already have treatment options available.
-It would therefore be useful to know whether a phenotype, although highly severe, is currently treatable, and also whether there exist known clinical endpoints for clinical trial testing.
-GPT-4 does seem to take into account quality of care, i.e.
-the degree to which health services increase the likelihood of desired outcomes (World Health Organization).
-For example, many of the cancer phenotypes are justified as always or often causing death unless treated or caught early.
-On the other hand, some cancers are justified as rarely causing death if appropriate treatment is provided, which may not always be the case for individuals or populations with access to less advanced medical systems.
-In future efforts, we might want to ask GPT-4 more explicitly to distinguish between whether a phenotype would cause death without treatment, or with best available standard of care.
+One key limitation of our study is the fact that we did not explicitly interrogate GPT-4 to assess how the availability of treatments affected the annotations it produced.
+For example, there are some very severe conditions for which highly effective treatments and early detection screens are widely available (e.g.
+syphilis, some forms of melanoma), thus rendering them fully treatable or even curable provided access to modern healthcare.
+It would therefore be useful to further interrogate GPT-4 to uncover how the availability of treatments influences its responses.
+Many of our findings here seem to indicate that GPT-4 does take into account quality of care to the extent that health services increase the likelihood of desired outcomes.
+For example, many of the cancer phenotypes are justified as always or often causing death unless detected and treated early in the disease course.
+On the other hand, some cancers are justified as rarely causing death if appropriate treatment is provided, which may not always be the case for individuals or populations with access to less access to quality healthcare services.
+Future efforts could more explicitly ask GPT-4 whether the phenotype would cause death with no or suboptimal treatment.
+
+In addition to these technical challenges, there are multiple factors that need to be considered when trying to prioritise phenotypes for their suitability for gene therapy development.
+First, while we have attempted to formalise severity here, this is an inherently subjective concept that may vary considerably across different individuals and contexts.
+For instance, one could ask whether a condition that always causes death is worse than a condition that causes a lifetime of severe disability (e.g.
+paralysis, blindness, intellectual disability).
+Metrics such as quality-adjusted life years (QALYs) have been proposed in the past to address these dilemmas by defining health as a function of both the length and quality of life [@prietoProblemsSolutionsCalculating2003].
+With regards to the financial burden of diseases, in some situations phenotypes which require many years of expensive medical care may be prioritised over those that result in extremely early onset lethality and little opportunity for therapeutic intervention.
+Another factor that affects the viability of a therapeutic program is the speed, cost and other practical considerations of a clinical trial.
+For instance, measuring risk of ageing-related respiratory failure over a ten-year period may be impractical in some cases.
+However, testing for total reversal of an existing severe phenotype could potentially yield faster and more immediately impactful results.
+If performed in close collaboration with medical ethicists, governmental organisations, advocacy groups and patient families, such cost/benefit assessments could be aided by LLMs through the scalable gathering of relevant data.
+As AI capabilities continue to advance, the range of applications in which they can be used effectively will continue to grow.
 
@@
 
-In conclusion, our study represents a significant step towards harnessing the power of AI to advance phenotypic annotation and severity assessment in rare diseases.
+In conclusion, our study represents a significant step towards harnessing the power of AI to advance phenotypic annotation and severity assessment across all rare diseases.
 This resource aims to provide researchers and clinicians with actionable insights that can inform rare disease research and improve the lives of individuals affected by rare diseases.
@@
 ### Annotating the HPO using OpenAI GPT-4
-We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO).
-Our annotation framework was developed based on previously defined criteria for classifying disease severity (Lazarin et al., 2014).
-We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
-Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
+
+We wrote a Python script to iteratively query GPT-4 via the OpenAI application programming interface (API).
+The ultimately yielded consistently formatted annotations for `r n_ids_gpt` terms within the HPO.
+Our annotation framework was developed based on previously defined criteria for classifying disease severity [@Lazarin2014-rz].
+We sought to evaluate whether each phenotype directly caused a given severity-related clinical characteristic, including: intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and/or had a congenital onset.
+Through prompt engineering we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
+We asked how frequently the given phenotype directly causes each clinical characteristic - whether it never, rarely, often, or always occurred.
+Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance (in terms of producing the desired annotations, and adhering to the formatting requirements) was achieved when inputting no more than two or three phenotypes per prompt.
+An example prompt can be seen in @fig-occurrence.
+Only 2 phenotypes were included per prompt in order to 1) avoid exceeding per-query token limits, and 2) prevent the breakdown of GPT-4 performance due to long-form text input, which is presently a known limitation common to many LLMs including GPT-4 [@weiLongformFactualityLarge2024].
 
 ### Calculating the true positive rate
-A true positive rate was calculated as a measure of the accuracy of the GPT-4 annotations.
+
+```{r tbl-true-positives}
+#| label: tbl-true-positives
+#| tbl-cap: For each clinical characteristic, any terms that matched
+
+queries <- eval(formals(HPOExplorer::search_hpo)$queries)
+query_hits_dt <- data.table(
+  "Clinical characteristic"=stringr::str_to_sentence(gsub("_"," ",names(queries))),
+  "HPO queries"=trimws(gsub("^","",sapply(queries, paste, collapse="; <br>"), fixed = TRUE)),
+  "True positive HPO IDs"=sapply(query_hits[names(queries)], data.table::uniqueN)
+)
+knitr::kable(as.data.frame(query_hits_dt))
+```
+
+A true positive rate was calculated as a measure of the recall of the GPT-4 annotations.
 This was achieved by identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the occurrence of certain clinical characteristics, and using all descendants of this HPO branch as true positives.
-For example, all descendants of the terms Abnormal central motor function and abnormality of movement should always or often cause impaired mobility (Table 1).
-
-| Clinical characteristics | True positive HPO branch |
-| ------------------------ | ------------------------ |
-| Intellectual disability  | Intellectual disability (HP:0001249)
-| Impaired mobility.       | Abnormal central motor function (HP:0011442) and Abnormality of movement (HP:0100022)
-| Physical malformations   | Malformation (string search)
-| Blindness                | Blindness (HP:0000618)
-| Sensory impairments      | Abnormality of vision (HP:0000504), Abnormality of the sense of smell (HP:0004408), Abnormality of taste sensation (HP:0000223), Somatic sensory dysfunction (HP:0003474), and Hearing abnormality (HP:0000364)
-| Immunodeficiency         | Immunodeficiency (HP:0002721)
-| Cancer                   | Neoplasm (HP:0002664)
-| Reduced fertility        | Decreased fertility (HP:0000144)
+For example, all descendants of the terms 'Intellectual disability' (HP:0001249) or 'Mental deterioration' (HP:0001268) should be annotated as always or often causing intellectual disability (@tbl-true-positives).
 
 ### Quantifying phenotypic severity
-The GPT-4 generated annotation occurrences were converted into a semi-quantitative scoring system, with ‘always’ corresponding to 3, ‘often’ to 2, ‘rarely’ to 1, and ‘never’ to 0.
-These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided annotations (Table 2).
+
+The GPT-4 generated annotation occurrences were converted into a semi-quantitative scoring system, with 'always' corresponding to 3, 'often' to 2, 'rarely' to 1, and 'never' to 0.
+These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided annotations (@tbl-metric-weights).
 Subsequently, the weighted scores underwent normalisation to yield a final quantitative severity score ranging from 0-100, with 100 signifying the maximum severity score attainable.
 
-| Clinical characteristic     | Always (4) | Often (3) | Rarely (2) | Never (0) |
-| Intellectual_disability (5) |     20     |     15    |    10      |     0     | 
-| Impaired_mobility (5)       |     20     |     15    |    10      |     0     | 
-| Blindness (4)               |     16     |     12    |    8       |     0     | 
-| Congenital onset (4)        |     16     |     12    |    8       |     0     | 
-| Physical_malformations (3)  |     12     |     9     |    6       |     0     | 
-| Sensory_impairments (3)     |     12     |     9     |    6       |     0     | 
-| Immunodeficiency (3)        |     12     |     9     |    6       |     0     | 
-| Cancer (3)                  |     12     |     9     |    6       |     0     | 
-| Reduced_fertility (1)       |     4      |     3     |    2       |     0     |
+Let us denote:
 
-## Supplementary figures
+-   $p$ : a phenotype in the HPO.
 
-```{r supp_figure_1}
-dat2 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
-dat2 <- add_ancestor(dat2,
-                     keep_descendants = keep_descendants)
-dat2[,mean_severity_score_gpt:=mean(severity_score_gpt, na.rm=TRUE),
-     by="ancestor_name"] |>
-  data.table::setorderv("mean_severity_score_gpt", -1, na.last = TRUE)
-dat2[,ancestor_name:=factor(ancestor_name,
-                            levels = unique(dat2$ancestor_name),
-                            ordered = TRUE)]
+-   $j$ : the identity of a given annotation metric (i.e. clinical characteristic, such as 'intellectual disability' or 'congenital onset').
 
-ggplot2::ggplot(dat2, ggplot2::aes(x=severity_score_gpt
-                                   # fill=factor(congenital_onset)
-)) +
-  ggplot2::geom_histogram(bins = 50, fill="slateblue") +
-  ggplot2::geom_vline(ggplot2::aes(xintercept=mean_severity_score_gpt),
-                      color="red") +
-  ggplot2::geom_label(data = unique(
-    dat2[,list(mean_severity_score_gpt), by="ancestor_name"]
-  ),
-  ggplot2::aes(x=mean_severity_score_gpt,
-               y=Inf,
-               label=round(mean_severity_score_gpt,2)),
-  color="red", size=3, hjust = 0, vjust = 1.5, alpha=.65) +
-  ggplot2::facet_wrap(facets = "ancestor_name~.", scales = "free_y",
-                      ncol = 3) +
-  ggplot2::labs(y="Frequency", x="Severity score") +
-  ggplot2::theme_bw() +
-  ggplot2::theme(strip.background = ggplot2::element_rect(
-    fill = "transparent"), 
-    text = element_text(size=16))
+-   $W_j$: the assigned weight of metric $j$.
+
+-   $F_j$: the maximum possible value for metric $j$ (equivalent across all $j$).
+
+-   $F_{pj}$ : the numerically encoded value of annotation metric $j$ for phenotype $p$.
+
+-   $NSS_p$: the final composite severity score for phenotype $p$ after applying normalisation to align values to a 0-100 scale and ensure equivalent meaning regardless of which other phenotypes are being analysed in addition to $p$.
+This allows for direct comparability of severity scores across studies with different sets of phenotypes.
+
+::: {#eq-gpt .content-hidden unless-format="html"}
+![](figures/eq5.png){height="300px"}
+
+Computing normalised severity score from encoded GPT-4 annotations.
+:::
+
+\
+\
+
+::: {.content-visible unless-format="html"}
+```{=tex}
+\begin{equation*}
+  \eqnmarkbox[Brown4]{nss}{NSS_p}
+  =
+  \frac{ 
+    \eqnmarkbox[Goldenrod]{nss2}{\sum_{j=1}^{m}} 
+    (
+      \eqnmarkbox[Goldenrod4]{nss3}{F_{pj}}
+      \times 
+      \eqnmarkbox[IndianRed4]{nss4}{W_j}
+    )
+    }{
+    \eqnmarkbox[Tan]{nss5}{\sum_{j=1}^{m}(\max\{F_j\} \times W_j)} 
+  } \times 100
+\end{equation*}
+\annotate[yshift=1em]{left}{nss}{Normalised Severity Score \\for each phenotype}
+\annotate[yshift=3em]{left}{nss2}{Sum of weighted annotation values \\across all metrics}
+\annotate[yshift=3em]{right}{nss3}{Numerically encoded annotation value \\of metric $j$ for phenotype $p$}
+\annotate[yshift=1em]{right}{nss4}{Weight for metric $j$} 
+\annotate[yshift=-1em]{below,right}{nss5}{Theoretical maximum severity score}
 ```
+:::
+
+\
+
+```{r tbl-metric-weights}
+#| label: tbl-metric-weights
+#| tbl-cap: Weighted scores for each clinical characteristic and GPT-4 response category.
+
+## Get weights from HPOExplorer::gpt_annot_codify function defaults
+args <- formals(HPOExplorer::gpt_annot_codify)
+code_dict <- eval(args$code_dict)|>sort(decreasing = TRUE)
+names(code_dict) <- paste0(stringr::str_to_sentence(names(code_dict)),
+                           " (",code_dict,")")
+tiers_dict <- eval(args$tiers_dict)|>unlist()|>sort(decreasing = TRUE)
+names(tiers_dict) <- paste0(stringr::str_to_sentence(gsub("_"," ",names(tiers_dict))),
+                           " (",tiers_dict,")")
+## Create the outer product of the two weight vectors
+tbl_metric_weights <- as.data.table(outer(tiers_dict,code_dict, "*"),
+                                    keep.rownames = "Clinical characteristic")
+
+knitr::kable(as.data.frame(tbl_metric_weights))
+```
+
+## Data and code availability statement
+
+All code and data used in this study are available on GitHub at:
+
+<https://github.com/neurogenomics/gpt_hpo_annotations>
+
+The GPT-4 annotations for all HPO phenotypes are made available through the R function `HPOExplorer::gpt_annot_read` or in CSV format at:
+
+<https://github.com/neurogenomics/gpt_hpo_annotations/tree/master/data>
+
+A fully reproducible version of this Quarto manuscript can be found at:
+
+<https://github.com/neurogenomics/gpt_hpo_annotations/blob/master/manuscript.qmd>
+
+## Acknowledgements
+
+We would like to thank members of the Monarch Initiative for their insight and feedback throughout this project. In particular, Peter Robinson.
+
+### Funding
+
+This work was supported by a UK Dementia Research Institute (UK DRI) Future Leaders Fellowship \[MR/T04327X/1\] and the UK DRI which receives its funding from UK DRI Ltd, funded by the UK Medical Research Council, Alzheimer’s Society and Alzheimer's Research UK.
 
@@
 
-:::{#refs}
+::: {#refs}
+:::
 
-:::
+{{< pagebreak >}}
+
+## Supplementary Materials
+
+### Supplementary Figures
+
+```{r fig-severity-histo}
+#| label: fig-severity-histo
+#| fig-cap: Distribution of the composite GPT-4 severity score of the severity scores for all HPO terms.
+#| fig-height: 8
+#| fig-width: 10
+
+gpt_annot_plot_out$gp3 +
+  labs(x="Severity score") 
+```
+
+```{r fig-severity-boxplot}
+#| label: fig-severity-boxplot
+#| fig-cap: Boxplot showing the relationship between composite severity score (y-axis) and the frequency response categories within each annotation type.
+#| fig-height: 5
+#| fig-width: 8
+
+gpt_annot_plot_out$gp2 +
+  scale_fill_brewer(palette = "GnBu", direction = -1) +
+  labs(y="Severity score") 
+```
+
+{{< pagebreak >}}
+
+### Supplementary Tables
+
+```{r tbl-fas}
+#| label: tbl-fas
+#| tbl-cap: Severity nnotations generated for GPT-4 annotations for the HPO phenotype 'Fetal akinesia sequence' (HP:000198).
+
+annot_cols <- setdiff(grep("justification",names(gpt_annot), 
+                       value = TRUE, invert = TRUE),
+                       c("hpo_name","hpo_id","pheno_count"))
+tbl_fas <- gpt_annot[hpo_name=='Fetal akinesia sequence',]|>
+  data.table::melt.data.table(id.vars = c("hpo_name","hpo_id"), 
+                              measure.vars = list(
+                                Annotation=annot_cols,
+                                Justification=grep("justification",names(gpt_annot), 
+                                                   value = TRUE)
+                                )
+                              )
+tbl_fas[,"Clinical characteristic":=stringr::str_to_sentence(gsub("_"," ",annot_cols))]
+knitr::kable(
+  as.data.frame(tbl_fas[,c("Clinical characteristic","Annotation","Justification")])
+  )
+```

### 2024-04-19 code/prompt_gen.R

@@ -1,62 +1,59 @@
-## HPO phenotype annotations - code to generate prompts for input into chat GPT ##
-library(HPOExplorer)
-library(data.table)
-
-# get all phenotypes from the HPO 
-all_res <- get_hpo()
-
-
-# get phenotype names
-phenos <- data.frame(phenotype = all_res@elementMetadata[["name"]])
-
-# 2 phenotypes per prompt
-batch_size = 2
-
-n_terms <- nrow(phenos)
-batches <- split(seq_len(n_terms), 
-                 ceiling(seq_along(seq_len(n_terms))/batch_size))
-
-
-# prep prompt 
-effects <- "intellectual disability, death, impaired mobility, 
-physical malformations, blindness, sensory impairments, 
-immunodeficiency, cancer, reduced fertility?"
-
-# define the columns of the output table 
-table_columns <- "phenotype, intellectual_disability, death, impaired_mobility, 
-physical_malformations, blindness, sensory_impairments, immunodeficiency, cancer, 
-reduced_fertility, congenital_onset."
-
-# function to generate prompts
-prompt_gen <- function(x){ 
-  
-  terms <- paste(
-    x,
-    collapse="; "
-  ) 
-  
-  question <- paste("I need to annotate phenotypes as to whether they typically cause:", 
-                   effects, 
-                   "Do they have congenital onset?",
-                   "To answer, use a severity scale of: never, rarely, often, always. Do not consider indirect effects.",
-                   "You must provide the output in python code as a data frame called df with columns:",
-                   table_columns, 
-                   "Also add a separate justification column for each outcome, e.g. death, death_justification.",
-                   "These are the phenotypes:", 
-                   paste(terms,".", sep=""), 
-                   "Placeholders are not acceptable.")
-  question <- gsub("\n", "", question)
-  
-  return(question)
+#' Generate a table of prompts
+#' 
+#' HPO phenotype annotations - code to generate prompts for input into chat GPT. 
+#' @param effects Clinical characteristics to generate annotations for.
+#' @param table_columns Define the columns of the output table.
+#' @param batch_size The number of phenotypes per prompt.
+#' @param hpo Get all phenotypes from the HPO.
+#' @export 
+prompt_gen_table <- function(effects = c("intellectual disability",
+                                          "death",
+                                          "impaired mobility",
+                                          "physical malformations",
+                                          "blindness",
+                                          "sensory impairments",
[…]

### 2024-04-19 docs/_tex/manuscript.tex
change output folder to docs for GH pages site

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{agujournal2019}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\usepackage{annotate-equations}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology},
  pdfauthor={Kitty B Murphy; Brian M Schilder; Nathan G Skene},
  pdfkeywords={Ontology, Human Phenotype Ontology, Large Language
Model, GPT-4, Artificial intelligence, Generative AI, Rare
disease, Automation, Medicine},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{medRxiv}

\draftfalse

\begin{document}
\title{Harnessing AI to annotate the severity of all phenotypic
abnormalities within the Human Phenotype Ontology}

\authors{Kitty B Murphy\affil{1,2}, Brian M Schilder\affil{1,2}, Nathan
G Skene\affil{1,2}}
\affiliation{1}{Department of Brain Sciences, Imperial College London,
UK, }\affiliation{2}{UK Dementia Research Institute at Imperial College
London, UK, }
\correspondingauthor{Kitty B
Murphy}{kitty.murphy@ukdri.ac.uk}\correspondingauthor{Nathan G
Skene}{n.skene@imperial.ac.uk}

\subsection{Abstract}\label{abstract}

The Human Phenotype Ontology (HPO) has played a crucial role in
defining, diagnosing, prognosing, and treating human diseases by
providing a standardised framework for describing phenotypic
abnormalities. However, documentation on the severity of each phenotype
is currently very sparse, making systematic analyses and prioritisation
of results challenging. With \textgreater17500 phenotypic abnormalities
across \textgreater8600 rare diseases, manual curation of such
phenotypic annotations by experts would be exceedingly labour-intensive
and time-consuming. Leveraging advances in artificial intelligence, we
employed the OpenAI GPT-4 large language model (LLM) to systematically
annotate the severity of all phenotypic abnormalities in the HPO. First,
we benchmarked the generative LLM annotations against ground-truth
labels within the HPO (e.g.~phenotypes in the `Cancer' HPO branch were
annotating as causing cancer by GPT-4). True positive recall rates
across different annotations ranged from 89-100\% (mean=96\%), clearly
demonstrating the ability of GPT-4 to automate the curation process with
a high degree of fidelity. Using a novel approach, we developed a
severity scoring system that incorporates both the nature of the
phenotype outcome and the frequency of its occurrence. These severity
metrics will enable efforts to systematically prioritise which human
phenotypes are most detrimental to human health, and best targets for
therapeutic intervention.

\subsection{Introduction}\label{introduction}

Comprehensive annotation of phenotypic abnormalities is invaluable for
defining, diagnosing, prognosing, and treating human disease. Since
2008, the Human Phenotype Ontology (HPO) has been instrumental to this,
by providing a standardised database for the description and analysis of
human phenotypes (Gargano et al., 2024; Köhler et al., 2021). Through
developing open community resources, the depth and breadth of the HPO
has continued to expand and there are now \textgreater17500 phenotypic
abnormalities across \textgreater8600 diseases, described. In recent
years, the HPO has expanded its disease annotations so that each HPO
term can have metadata including typical age of onset and frequency. In
addition, there are the \emph{Clinical modifier} and \emph{Clinical
course} subontologies, which contains terms to describe factors
including severity, triggers, mortality and progression, respectively.
Describing the severity-related attributes of a disease is crucial for
both research and clinical care of individuals with rare diseases. This
includes enhancing diagnostic capabilities, as well as prioritising and
guiding gene therapy trials.

To date, the HPO has largely relied on manual curation by experts
including clinicians, clinical geneticists and researchers. Although
this approach ensures the quality and accuracy of the ontology, it is
time-consuming and labour-intensive. As a result, less than 1\% of terms
within the HPO have metadata pertaining to their features such as time
course and severity. Artificial intelligence (AI) capabilities have
advanced considerably in recent years, presenting new opportunities to
integrate natural language processing technologies into assisting in the
curation process. Specifically, there have recently been considerable
advances in large language model (LLM) and their application to
biomedical problems, in some cases performing as well or better than
human clinicians on standardised medical exams and patient diagnosis
tasks (Bolton et al., n.d.; Cheng et al., 2023; Gu et al., 2021; Labrak
et al., n.d.; Luo et al., 2022; McDuff et al., n.d.; O'Neil et al.,
n.d.; Shin et al., 2020; Singhal et al., n.d., 2023, 2023; Van Veen et
al., 2024; Zhang et al., n.d.). Recent work has demonstrated that the
Generative Pre-trained Transformer 4 (GPT-4) foundation model (OpenAI et
al., n.d.), when combined with strategic prompt engineering, can
outperform even specialist LLMs that are explicitly fine-tuned for
biomedical tasks (Nori et al., n.d.). In a landmark achievement, GPT-4
was the first LLM to surpass a score of 90\% in the United States
Medical Licensing Examination (USML) (Nori et al., n.d.).

Here, we have used GPT-4 to systematically annotate the severity of
17502 / 17548 (99.7\%) phenotypic abnormalities within the HPO. Our
severity annotation framework was adapted from previously defined
criteria developed through consultation with clinicians (Lazarin et al.,
2014). The authors consulted 192 healthcare professionals for their
opinions on the relative severity of various clinical characteristics:
they used this to create a system for categorising the severity of
diseases. Briefly, each healthcare professional was sent a survey asking
them to first rate how important a disease characteristic was for
determining disease severity, and then to rate the severity of a set of
given disease. Using the responses, the authors were able to categorise
clinical characteristics into 4 `severity tiers'. While characteristics
such as shortened lifespan in infancy and intellectual disability were
identified as highly severe and placed into tier 1, sensory impairment
and reduced lifespan were categorised as less severe and placed into
tier 4. Standardised metrics of severity allow clinicians to quickly
assess the urgency of treating a given phenotype, as well as prognosing
what outcomes might be expected.

793 phenotypes were annotated multiple times to evaluate the consistency
of responses generated by GPT-4. For a subset of phenotypes with known
expected annotations, true positive rates were calculated to assess
recall. Additionally, based on the clinical characteristics and their
occurrence, we have quantified the severity of each phenotype, providing
an example of how these clinical characteristic annotations can be used
to guide prioritisation of gene therapy trials. Ultimately, we hope that
our resource will be of utility to those working in rare diseases, as
well as the wider healthcare community.

\subsection{Results}\label{results}

\subsubsection{Annotating the HPO using
GPT-4}\label{annotating-the-hpo-using-gpt-4}

\phantomsection\label{cell-fig-occurrence}
\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-pdf/fig-occurrence-1.pdf}

}

\caption{\label{fig-occurrence}GPT-4 was able to annotate all human
phenotypes based on whether they are always/often/rarely/never
associated with different clinical characteristics. \textbf{a}, An
example of the prompt input given to to GPT-4. The phenotypes listed in
the second to last sentence (\emph{italicised}) were changed to allow
all HPO phenotypes to be annotated. \textbf{b}, Stacked bar plot showing
the proportion of the occurrence of each clinical characteristic across
all annotated HPO phenotypes. The terms shown on the x-axis are the
clinical characteristics for which GPT-4 was asked to determine whether
each phenotype caused them.}

\end{figure}%

We employed the OpenAI GPT-4 model with Python to annotate 17502 terms
within the HPO (v2024-02-08) (Gargano et al., 2024; Köhler et al.,
2021). Our annotation framework was developed based on previously
defined criteria for classifying disease severity (Lazarin et al.,
2014). We sought to evaluate the impact of phenotypes on factors
including intellectual disability, death, impaired mobility, physical
malformations, blindness, sensory impairments, immunodeficiency, cancer,
reduced fertility, and congenital onset. Through prompt design we found
that the performance of GPT-4 improved when we incorporated a scale
associated with each effect and required a justification for each
response. For each effect, we asked about the likelihood of its
occurrence - whether it never, rarely, often, or always occurred.

Phenotype outcome occurrence varied across annotation categories.
\textgreater50\% of phenotypes never caused blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility or intellectual
disability. Only a minority of phenotypes (21.7\%) never had a
congenital onset, which is expected as rare disorders tend to be early
onset genetic conditions Fig.~\ref{fig-occurrence}.

Less than 1\% of phenotype always directly resulted in death, such as
`Stillbirth', `Anencephaly' and `Bilateral lung agenesis'. Only 71
phenotypes were annotated as always causing death, while 9707 phenotypes
were annotated as often or rarely causing death. 7880 phenotypes were
annotated as never causing death. This included 34 unique forms of
syndactyly, a non-lethal condition that causes fused or webbed fingers
(occurring 1 in 1,200--15,000 live births). While not life-threatening
itself, syndactyly is a symptom of genetic disorders that can cause
life-threatening cardiovascular and neurodevelopmental defects, such as
Apert Syndrome (Garagnani \& Smith, 2013). This example highlights the
ability of GPT-4 to successfully distinguish between phenotypes that
directly cause lethality, and those that are often associated with
diseases that cause lethality.

\subsubsection{Annotation consistency and
recall}\label{annotation-consistency-and-recall}

To assess annotation consistency, 793 phenotypes underwent duplicate
annotations. We employed two different metrics to determine the
\emph{consistency rate}. The first, less stringent metric, defined
consistency as the duplicate annotations being either `always' and
`often', or `never' and `rarely'. The second, more stringent metric,
required exact agreement in annotation occurrences, e.g.~`always' and
`always'. For the less stringent metric, duplicated phenotypes were
annotated consistently at a rate of at least 80\%, and for the more
stringent metric, the lowest consistency rate was 57\% for congenital
onset. An example of where annotations were inconsistent was for the HPO
term `Acute leukaemia'. One time, GPT-4 annotated it as often causing
impaired mobility, giving the justification that `weakness and fatigue
from leukaemia and its treatment can impair mobility'. The other time,
GPT-4 annotated it as rarely causing impaired mobility, giving the
justification that `acute leukaemia rarely impairs mobility directly'.
Despite specifying in the prompt for GPT-4 not to take into
consideration indirect effects, here is an example of where it failed to
do so.

\phantomsection\label{cell-fig-checks}
\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-pdf/fig-checks-1.pdf}

}

\caption{\label{fig-checks}GPT-4 annotations are consistent and accurate
across annotations. \textbf{a}, Barplot showing the annotation
consistency within phenotypes that were annotated more than once. In the
lenient metric, annotations were collapsed into two groups
(`always'/`often' and `never'/`rarely'). For a given clinical
characteristic within a given phenotype, if an annotation was always
within the same group it was considered consistent. In the stringent
metric, all four annotation categories were considered to be different
from one another. Thus, annotations were only defined as consistent if
they were all identical. The blue dashed line indicates the probability
of two annotations being consistent by chance in the lenient metric
(\textasciitilde1/2). The gold dashed line is the probability of two
annotations being consistent by chance in the stringent metric
(\textasciitilde1/4). \textbf{b}, Bar plot of the true positive rate for
each annotation. The labels above each bar indicate the number of
phenotypes tested.}

\end{figure}%

In order to evaluate the validity of the annotations, we calculated a
true positive rate. This involved identifying specific branches within
the HPO that would contain phenotypes that would reliably indicate the
presence of certain conditions. For instance, the phenotypes `Decreased
fertility in females' and `Decreased fertility in males' should often or
always cause reduced fertility. We observed an encouraging true positive
rate exceeding 88\% across in every clinical characteristic and
achieving perfect recall (100\%) in 4/8 characteristics.

The lowest true positive rate was observed for physical malformations,
with 88.5\% recall across 87 HPO phenotypes. Some cases in which the
GPT-4 annotations disagreed with the HPO ground truth included: `Angioma
serpentinum', `Nevus anemicus', `Pulmonary arteriovenous fistulas'. In
the case of `Angioma serpentinum' it provided the justification that `No
known association with physical malformations'. In another instance,
GPT-4 noted that `Nevus anemicus' is `Limited to hypopigmented skin
patch; no other malformations.'. This indicates that while technically
incorrect according to our predefined benchmarks, a case could in fact
be made that mild skin conditions do not rise to the level of physical
malformations.

This high level of recall underscores the robustness of our annotations
and the reliability of the HPO framework in capturing clinically
relevant phenotypic information. However, we acknowledge that the number
of testable true positive phenotypes for some of these categories are
low, especially `blindness' for which there is only 1 phenotype in the
HPO (after excluding terms pertaining to colour or night blindness).
Furthermore, some of the true positive phenotypes are lexically similar
to the name of the clinical characteristic itself. In these cases,
annotating `Severe intellectual disability' as always causing
intellectual disability is a relatively trivial task. Nevertheless, even
these scenarios provide a clear and interpretable benchmark for the
model's performance. In addition, were numerous phenotypes with
lexically non-obvious relationships to the clinical characteristic that
were annotated correctly by GPT-4. For example, `Molar tooth sign on
MRI' was correctly annotated as causing intellectual disability.

\subsubsection{Quantifying phenotypic
severity}\label{quantifying-phenotypic-severity}

Quantifying the severity of phenotypes can have important implications
for diagnosis, prognosis, and treatment, for instance, guiding
prioritisation of gene therapy trials for the most severe phenotypes.
First, we created a dictionary to map each phenotype outcome
(e.g.~blindness) and its response (always, often, rarely, never) to
numeric values from 0-3. Then, the phenotype outcome values were
multiplied by their associated response values. Importantly, the values
reflected the severity of each outcome based on both the outcome itself
and the frequency of the response. For instance, a phenotype always
causing death would have a higher multiplied value than a phenotype
rarely causing reduced fertility (see Table~\ref{tbl-metric-weights}).
Next, we computed an average score for each phenotype by aggregating the
multiplied values across all phenotype outcomes and then calculating the
mean. This was then normalised by the theoretical maximum severity
score, so that all phenotypes were on a 0-100 severity scale (where 100
is the most severe phenotype possible). This average normalised score
represents the overall severity of the phenotype based on the severity
of its individual outcomes.

Based on these scores we evaluated the top 50 severe phenotypes. One of
the most severe phenotype was `Anencephaly' (HP:0002323) with a
composite severity score of 52. Anencephaly is a birth defect where the
baby is born without a portion of its brain and skull, often these
babies are stillborn. In fact, many of the most severe phenotypes were
related to developmental brain and neural tube defects. Comparison of
the severity scores for each response, across the phenotype outcomes
annotated, revealed consistent trends: as the response of the phenotype
outcome increased (from never to always), the severity score also
increased (Supplementary Fig.~\ref{fig-severity-boxplot}). We also
evaluated the severity score distribution by HPO branch and calculated
the mean severity score using all phenotypes within each major HPO
branch (Fig.~\ref{fig-severity-histo}). The HPO branch with the greatest
mean severity score was `Abnormal cellular phenotype' (mean=19.6),
followed by `Neoplasm' (mean=17.2), which would include the highly
ranked phenotypes seen in Figure~\ref{fig-top-phenos}.

\phantomsection\label{cell-fig-top-phenos}
\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-pdf/fig-top-phenos-1.pdf}

}

\caption{\label{fig-top-phenos}Quantifying the severity of HPO phenotype
annotations highlights the most impactful conditions. Heatmap of the top
(n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based
on severity score. Severity scores were calculated by multiplying the
numeric values assigned to each clinical characteristic according to
Table~\ref{tbl-metric-weights}. The average normalised score,
representing overall phenotype severity on a 0-100 scale, was calculated
by aggregating the multiplied values and normalising by the theoretical
maximum severity score. The x-axis is scaled by the severity of the
clinical characterisation.}

\end{figure}%

\subsection{Discussion}\label{discussion}

The creation and annotation of biomedical knowledge has traditionally
relied on manual or semi-manual curation by human experts (Gargano et
al., 2024; Köhler et al., 2021; Mungall et al., 2017; Ochoa et al.,
2021; Putman et al., 2024). Performing such manual curation and review
tasks at scale is often infeasible for human biomedical experts given
limited time and resources. LLMs have the capacity to effectively
encode, retrieve, and synthesise vast amounts of diverse information in
a highly scalable manner (OpenAI et al., n.d.; Singhal et al., 2023; Van
Veen et al., 2024). This makes them powerful tools that can be applied
in a rapidly expanding variety of scenarios, including medical practice,
research and data curation (Caufield et al., n.d.; O'Neil et al., n.d.;
Pan et al., n.d.; Singhal et al., 2023; Toro et al., n.d.).

Here, we introduce a novel framework to leverage the current
best-in-class LLM, GPT-4 (OpenAI et al., n.d.), to systematically
annotate the severity of 17502 phenotypic abnormalities within the HPO.
By employing advanced AI capabilities, we have demonstrated the
feasibility of automating this process, significantly enhancing
efficiency without substantially compromising accuracy. Our validation
approach yielded a high true positive rate exceeding 88\% across the
phenotypes tested. Furthermore, our approach can be readily adapted and
scaled to accommodate the growing volume of phenotypic data.

Throughout this study, we observed that GPT-4 was capable of reliably
recovering deep semantic relationships from the medical domain, far
beyond making superficial inferences based on lexical similarities. An
excellent example of this is the phenotype `Molar tooth sign on MRI'
(HP:0002419; severity score=30), which GPT-4 annotated as causing
intellectual disability. At first glance, we ourselves assumed this was
a false positive as the term appeared to be related to dentition.
However, upon further inspection we realised that molar tooth sign is in
fact a pattern of abnormal brain morphology that happens to bear some
resemblance to molar dentition when observed in radiological scans. This
phenotype is a known sign of neurodevelopmental defects that can indeed
cause severe intellectual disability (Gleeson et al., 2004).

In addition to rapidly synthesising and summarising vast amounts of
information, LLMs can also be steered to provide justifications for each
particular response. This makes LLMs amenable to direct interrogation as
a means of recovering explainability, especially when designed to retain
information about previous requests and interactions as they use these
to iteratively improve and update their predictions (Janik, n.d.). This
represents a categorical advance over traditional natural language
processing models based on more shallow forms of statistical or machine
learning (e.g.~Term Frequency-Inverse Document Frequency (Jones, 1972),
Word2vec (Mikolov et al., n.d.)) which lack the ability to provide
chains of causal reasoning to justify their predictions. This highlights
the fundamental trade-off between simpler models with high
explainability (the ability humans to understand the inner workings of
the model) but low interpretability (the ability of humans to trace the
decision process of the model, analogous to human `reasoning'), and
deeper more complex models with low explainability but high
interpretability (Marcinkevičs \& Vogt, n.d.).

A key contribution of our study is the introduction of a quantitative
severity scoring system that integrates both the nature of the phenotype
outcome and the frequency of its occurrence. By encoding the concept of
severity in this way, we are able to prioritise phenotypes based on
their impact on patients. The methodology allowed us to transition from
low-throughput qualitative assessments of severity (e.g. Lazarin et al.
(2014)) to high-throughput quantitative assessments of severity. One of
the most severe phenotypes in the HPO is `Fetal akinesia sequence' (FAS;
HP:0001989, severity score= 50.7), and extremely rare condition that is
almost always lethal. FAS is a complex, multi-system phenotype that can
be caused by at least 24 different genetic disorders. Despite the
complex and heterogenous aetiology of this phenotype, GPT-4 was able to
provide accurate annotations alongside explainable justifications for
those annotations (see Table~\ref{tbl-fas}). For example, this phenotype
almost always results in death, either \emph{in utero} or shortly after
birth. Not only did GPT-4 correctly provide the annotation death as
`always', when asked whether FAS causes sensory impairments it provided
the response `always' with the justification `Fetal akinesia sequence
typically results in severe sensory impairment due to neurodevelopmental
disruption.' Neurodevelopmental disruption is indeed a hallmark
component of FAS (e.g.~hydrocephalus, cerebellar hypoplasia) that causes
severe impairments across multiple sensory systems (Chen, 2012). This
demonstrates that GPT-4 was able to recover the correct chain of
causality from phenotype to clinical outcome.

Our findings highlight the potential of this next generation of natural
language processing technologies in significantly contributing to the
automation and refinement of data curation in biomedical research. These
results have a large number of useful real-world applications, such as
prioritising gene therapy candidates (Murphy et al., n.d.) and guiding
clinical decision-making in rare diseases. As the HPO continues to grow
over time, this approach can easily be repeated to keep pace with the
rapidly evolving biomedical landscape. Furthermore, this approach can be
extended to produce different sets of annotations or be used with any
other ontology. Additional use cases include gathering data on the
prevalence of each phenotype to approximate their social and financial
costs.

One key limitation of our study is the fact that we did not explicitly
interrogate GPT-4 to assess how the availability of treatments affected
the annotations it produced. For example, there are some very severe
conditions for which highly effective treatments and early detection
screens are widely available (e.g.~syphilis, some forms of melanoma),
thus rendering them fully treatable or even curable provided access to
modern healthcare. It would therefore be useful to further interrogate
GPT-4 to uncover how the availability of treatments influences its
responses. Many of our findings here seem to indicate that GPT-4 does
take into account quality of care to the extent that health services
increase the likelihood of desired outcomes. For example, many of the
cancer phenotypes are justified as always or often causing death unless
detected and treated early in the disease course. On the other hand,
some cancers are justified as rarely causing death if appropriate
treatment is provided, which may not always be the case for individuals
or populations with access to less access to quality healthcare
services. Future efforts could more explicitly ask GPT-4 whether the
phenotype would cause death with no or suboptimal treatment.

In addition to these technical challenges, there are multiple factors
that need to be considered when trying to prioritise phenotypes for
their suitability for gene therapy development. First, while we have
attempted to formalise severity here, this is an inherently subjective
concept that may vary considerably across different individuals and
contexts. For instance, one could ask whether a condition that always
causes death is worse than a condition that causes a lifetime of severe
disability (e.g.~paralysis, blindness, intellectual disability). Metrics
such as quality-adjusted life years (QALYs) have been proposed in the
past to address these dilemmas by defining health as a function of both
the length and quality of life (Prieto \& Sacristán, 2003). With regards
to the financial burden of diseases, in some situations phenotypes which
require many years of expensive medical care may be prioritised over
those that result in extremely early onset lethality and little
opportunity for therapeutic intervention. Another factor that affects
the viability of a therapeutic program is the speed, cost and other
practical considerations of a clinical trial. For instance, measuring
risk of ageing-related respiratory failure over a ten-year period may be
impractical in some cases. However, testing for total reversal of an
existing severe phenotype could potentially yield faster and more
immediately impactful results. If performed in close collaboration with
medical ethicists, governmental organisations, advocacy groups and
patient families, such cost/benefit assessments could be aided by LLMs
through the scalable gathering of relevant data. As AI capabilities
continue to advance, the range of applications in which they can be used
effectively will continue to grow.

While our study demonstrates the feasibility and utility of AI-driven
phenotypic annotation, several limitations must be acknowledged. The
reliance on computational algorithms may introduce biases or
inaccuracies inherent to the training data, necessitating ongoing
validation and refinement of our approach. Additionally, our severity
scoring system, while comprehensive, may not capture the full spectrum
of phenotypic variability or account for complex gene-environment
interactions. Future research should focus on further optimising
AI-driven annotation methodologies, incorporating additional data
modalities such as genomic and clinical data to enhance accuracy.

In conclusion, our study represents a significant step towards
harnessing the power of AI to advance phenotypic annotation and severity
assessment across all rare diseases. This resource aims to provide
researchers and clinicians with actionable insights that can inform rare
disease research and improve the lives of individuals affected by rare
diseases.

\subsection{Methods}\label{methods}

\subsubsection{Annotating the HPO using OpenAI
GPT-4}\label{annotating-the-hpo-using-openai-gpt-4}

We wrote a Python script to iteratively query GPT-4 via the OpenAI
application programming interface (API). The ultimately yielded
consistently formatted annotations for 17502 terms within the HPO. Our
annotation framework was developed based on previously defined criteria
for classifying disease severity (Lazarin et al., 2014). We sought to
evaluate whether each phenotype directly caused a given severity-related
clinical characteristic, including: intellectual disability, death,
impaired mobility, physical malformations, blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility, and/or had a
congenital onset. Through prompt engineering we found that the
performance of GPT-4 improved when we incorporated a scale associated
with each effect and required a justification for each response. We
asked how frequently the given phenotype directly causes each clinical
characteristic - whether it never, rarely, often, or always occurred.
Furthermore, our prompt design revealed that the optimal trade-off
between the number of phenotypes and performance (in terms of producing
the desired annotations, and adhering to the formatting requirements)
was achieved when inputting no more than two or three phenotypes per
prompt. An example prompt can be seen in Figure~\ref{fig-occurrence}.
Only 2 phenotypes were included per prompt in order to 1) avoid
exceeding per-query token limits, and 2) prevent the breakdown of GPT-4
performance due to long-form text input, which is presently a known
limitation common to many LLMs including GPT-4 (Wei et al., n.d.).

\subsubsection{Calculating the true positive
rate}\label{calculating-the-true-positive-rate}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1194}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7711}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1095}}@{}}

\caption{\label{tbl-true-positives}For each clinical characteristic, any
terms that matched}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Clinical characteristic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
HPO queries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
True positive HPO IDs
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intellectual disability & Intellectual disability; Mental deterioration
& 19 \\
Impaired mobility & Gait disturbance; Diminished movement; mobility &
319 \\
Physical malformations & malformation & 78 \\
Blindness & blindness & 1 \\
Sensory impairments & Abnormality of vision; Abnormality of the sense of
smell; Abnormality of taste sensation; Somatic sensory dysfunction;
Hearing abnormality & 252 \\
Immunodeficiency & Immunodeficiency; Impaired antigen-specific response
& 29 \\
Cancer & Cancer; malignant; carcinoma & 56 \\
Reduced fertility & Decreased fertility; Hypogonadism & 9 \\

\end{longtable}

A true positive rate was calculated as a measure of the recall of the
GPT-4 annotations. This was achieved by identifying specific branches
within the HPO that would contain phenotypes that would reliably
indicate the occurrence of certain clinical characteristics, and using
all descendants of this HPO branch as true positives. For example, all
descendants of the terms `Intellectual disability' (HP:0001249) or
`Mental deterioration' (HP:0001268) should be annotated as always or
often causing intellectual disability (Table~\ref{tbl-true-positives}).

\subsubsection{Quantifying phenotypic
severity}\label{quantifying-phenotypic-severity-1}

The GPT-4 generated annotation occurrences were converted into a
semi-quantitative scoring system, with `always' corresponding to 3,
`often' to 2, `rarely' to 1, and `never' to 0. These scores were then
weighted by a severity metric on a scale of 1-5, with 5 representing the
highest severity, as determined by the provided annotations
(Table~\ref{tbl-metric-weights}). Subsequently, the weighted scores
underwent normalisation to yield a final quantitative severity score
ranging from 0-100, with 100 signifying the maximum severity score
attainable.

Let us denote:

\begin{itemize}
\item
  \(p\) : a phenotype in the HPO.
\item
  \(j\) : the identity of a given annotation metric (i.e.~clinical
  characteristic, such as `intellectual disability' or `congenital
  onset').
\item
  \(W_j\): the assigned weight of metric \(j\).
\item
  \(F_j\): the maximum possible value for metric \(j\) (equivalent
  across all \(j\)).
\item
  \(F_{pj}\) : the numerically encoded value of annotation metric \(j\)
  for phenotype \(p\).
\item
  \(NSS_p\): The final composite severity score for phenotype \(p\)
  after applying normalisation to align values to a 0-100 scale and
  ensure equivalent meaning regardless of which other phenotypes are
  being analysed in addition to \(p\). This allows for direct
  comparability of severity scores across studies with different sets of
  phenotypes.
\end{itemize}

\hfill\break
\hfill\break

\begin{equation*}
  \eqnmarkbox[Brown4]{nss}{NSS_p}
  =
  \frac{ 
    \eqnmarkbox[Goldenrod]{nss2}{\sum_{j=1}^{m}} 
    (
      \eqnmarkbox[Goldenrod4]{nss3}{F_{pj}}
      \times 
      \eqnmarkbox[IndianRed4]{nss4}{W_j}
    )
    }{
    \eqnmarkbox[Tan]{nss5}{\sum_{j=1}^{m}(\max\{F_j\} \times W_j)} 
  } \times 100
\end{equation*}
\annotate[yshift=1em]{left}{nss}{Normalised Severity Score \\for each phenotype}
\annotate[yshift=3em]{left}{nss2}{Sum of weighted annotation values \\across all metrics}
\annotate[yshift=3em]{right}{nss3}{Numerically encoded annotation value \\of metric $j$ for phenotype $p$}
\annotate[yshift=1em]{right}{nss4}{Weight for metric $j$} 
\annotate[yshift=-1em]{below,right}{nss5}{Theoretical maximum severity score}

\hfill\break

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4000}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1571}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1429}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1571}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1429}}@{}}

\caption{\label{tbl-metric-weights}Weighted scores for each clinical
characteristic and GPT-4 response category.}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Clinical characteristic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Always (3)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Often (2)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Rarely (1)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Never (0)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intellectual disability (5) & 15 & 10 & 5 & 0 \\
Death (5) & 15 & 10 & 5 & 0 \\
Impaired mobility (4) & 12 & 8 & 4 & 0 \\
Blindness (4) & 12 & 8 & 4 & 0 \\
Physical malformations (3) & 9 & 6 & 3 & 0 \\
Sensory impairments (3) & 9 & 6 & 3 & 0 \\
Immunodeficiency (3) & 9 & 6 & 3 & 0 \\
Cancer (3) & 9 & 6 & 3 & 0 \\
Reduced fertility (1) & 3 & 2 & 1 & 0 \\
Congenital onset (1) & 3 & 2 & 1 & 0 \\

\end{longtable}

\subsection{Data and code availability
statement}\label{data-and-code-availability-statement}

All code and data used in this study are available on GitHub at:

\url{https://github.com/neurogenomics/gpt_hpo_annotations}

The GPT-4 annotations for all HPO phenotypes are made available through
the R function \texttt{HPOExplorer::gpt\_annot\_read} or in CSV format
at:

\url{https://github.com/neurogenomics/gpt_hpo_annotations/tree/master/data}

A fully reproducible version of this Quarto manuscript can be found at:

\url{https://github.com/neurogenomics/gpt_hpo_annotations/blob/master/manuscript.qmd}

\subsection{Acknowledgements}\label{acknowledgements}

We would like to thank members of the Monarch Initiative for their
insight and feedback throughout this project. In particular, Peter
Robinson.

\subsubsection{Funding}\label{funding}

This work was supported by a UK Dementia Research Institute (UK DRI)
Future Leaders Fellowship {[}MR/T04327X/1{]} and the UK DRI which
receives its funding from UK DRI Ltd, funded by the UK Medical Research
Council, Alzheimer's Society and Alzheimer's Research UK.

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-boltonBioMedLM7BParameter2024}
Bolton, E., Venigalla, A., Yasunaga, M., Hall, D., Xiong, B., Lee, T.,
et al. (n.d.). BioMedLM: A 2.7B parameter language model trained on
biomedical text. \url{https://doi.org/10.48550/arXiv.2403.18421}

\bibitem[\citeproctext]{ref-caufieldStructuredPromptInterrogation2023}
Caufield, J. H., Hegde, H., Emonet, V., Harris, N. L., Joachimiak, M.
P., Matentzoglu, N., et al. (n.d.). Structured prompt interrogation and
recursive extraction of semantics (SPIRES): A method for populating
knowledge bases using zero-shot learning.
\url{https://doi.org/10.48550/arXiv.2304.02711}

\bibitem[\citeproctext]{ref-chenPrenatalDiagnosisGenetic2012}
Chen, C.-P. (2012). Prenatal diagnosis and genetic analysis of fetal
akinesia deformation sequence and multiple pterygium syndrome associated
with neuromuscular junction disorders: A review. \emph{Taiwanese Journal
of Obstetrics and Gynecology}, \emph{51}(1), 12--17.
\url{https://doi.org/10.1016/j.tjog.2012.01.004}

\bibitem[\citeproctext]{ref-chengExploringPotentialGPT42023}
Cheng, K., Guo, Q., He, Y., Lu, Y., Gu, S., \& Wu, H. (2023). Exploring
the potential of GPT-4 in biomedical engineering: The dawn of a new era.
\emph{Annals of Biomedical Engineering}, \emph{51}(8), 1645--1653.
\url{https://doi.org/10.1007/s10439-023-03221-1}

\bibitem[\citeproctext]{ref-garagnaniSyndromesAssociatedSyndactyly2013}
Garagnani, L., \& Smith, G. D. (2013). Syndromes associated with
syndactyly. In J. M. Abzug, S. Kozin, \& D. A. Zlotolow (Eds.),
\emph{The pediatric upper extremity} (pp. 1--31). Springer.
\url{https://doi.org/10.1007/978-1-4614-8758-6_14-1}

\bibitem[\citeproctext]{ref-garganoHumanPhenotypeOntology2024}
Gargano, M. A., Matentzoglu, N., Coleman, B., Addo-Lartey, E. B.,
Anagnostopoulos, A. V., Anderton, J., et al. (2024). The human phenotype
ontology in 2024: Phenotypes around the world. \emph{Nucleic Acids
Research}, \emph{52}(D1), D1333--D1346.
\url{https://doi.org/10.1093/nar/gkad1005}

\bibitem[\citeproctext]{ref-gleesonMolarToothSign2004}
Gleeson, J. G., Keeler, L. C., Parisi, M. A., Marsh, S. E., Chance, P.
F., Glass, I. A., et al. (2004). Molar tooth sign of the
midbrain--hindbrain junction: Occurrence in multiple distinct syndromes.
\emph{American Journal of Medical Genetics Part A}, \emph{125A}(2),
125--134. \url{https://doi.org/10.1002/ajmg.a.20437}

\bibitem[\citeproctext]{ref-guDomainSpecificLanguageModel2021}
Gu, Y., Tinn, R., Cheng, H., Lucas, M., Usuyama, N., Liu, X., et al.
(2021). Domain-specific language model pretraining for biomedical
natural language processing. \emph{ACM Transactions on Computing for
Healthcare}, \emph{3}(1), 2:1--2:23.
\url{https://doi.org/10.1145/3458754}

\bibitem[\citeproctext]{ref-janikAspectsHumanMemory2024}
Janik, R. A. (n.d.). Aspects of human memory and large language models.
\url{https://doi.org/10.48550/arXiv.2311.03839}

\bibitem[\citeproctext]{ref-jonesStatisticalInterpretationTerm1972}
Jones, K. S. (1972). A statistical interpretation of term specificity
and its application in retrieval. \emph{Journal of Documentation},
\emph{28}(1), 11--21. \url{https://doi.org/10.1108/eb026526}

\bibitem[\citeproctext]{ref-kohlerHumanPhenotypeOntology2021}
Köhler, S., Gargano, M., Matentzoglu, N., Carmody, L. C., Lewis-Smith,
D., Vasilevsky, N. A., et al. (2021). The human phenotype ontology in
2021. \emph{Nucleic Acids Research}, \emph{49}(D1), D1207--D1217.
\url{https://doi.org/10.1093/nar/gkaa1043}

\bibitem[\citeproctext]{ref-labrakBioMistralCollectionOpenSource2024}
Labrak, Y., Bazoge, A., Morin, E., Gourraud, P.-A., Rouvier, M., \&
Dufour, R. (n.d.). BioMistral: A collection of open-source pretrained
large language models for medical domains.
\url{https://doi.org/10.48550/arXiv.2402.10373}

\bibitem[\citeproctext]{ref-Lazarin2014-rz}
Lazarin, G. A., Hawthorne, F., Collins, N. S., Platt, E. A., Evans, E.
A., \& Haque, I. S. (2014). Systematic classification of disease
severity for evaluation of expanded carrier screening panels. \emph{PLoS
One}, \emph{9}(12), e114391.

\bibitem[\citeproctext]{ref-luoBioGPTGenerativePretrained2022}
Luo, R., Sun, L., Xia, Y., Qin, T., Zhang, S., Poon, H., \& Liu, T.-Y.
(2022). BioGPT: Generative pre-trained transformer for biomedical text
generation and mining. \emph{Briefings in Bioinformatics}, \emph{23}(6),
bbac409. \url{https://doi.org/10.1093/bib/bbac409}

\bibitem[\citeproctext]{ref-marcinkevicsInterpretabilityExplainabilityMachine2023}
Marcinkevičs, R., \& Vogt, J. E. (n.d.). Interpretability and
explainability: A machine learning zoo mini-tour.
\url{https://doi.org/10.48550/arXiv.2012.01805}

\bibitem[\citeproctext]{ref-mcduffAccurateDifferentialDiagnosis2023}
McDuff, D., Schaekermann, M., Tu, T., Palepu, A., Wang, A., Garrison,
J., et al. (n.d.). Towards accurate differential diagnosis with large
language models. \url{https://doi.org/10.48550/arXiv.2312.00164}

\bibitem[\citeproctext]{ref-mikolovEfficientEstimationWord2013}
Mikolov, T., Chen, K., Corrado, G., \& Dean, J. (n.d.). Efficient
estimation of word representations in vector space.
\url{https://doi.org/10.48550/arXiv.1301.3781}

\bibitem[\citeproctext]{ref-mungallMonarchInitiativeIntegrative2017a}
Mungall, C. J., McMurry, J. A., Köhler, S., Balhoff, J. P., Borromeo,
C., Brush, M., et al. (2017). The monarch initiative: An integrative
data and analytic platform connecting phenotypes to genotypes across
species. \emph{Nucleic Acids Research}, \emph{45}(D1), D712--D722.
\url{https://doi.org/10.1093/nar/gkw1128}

\bibitem[\citeproctext]{ref-murphyIdentificationCellTypespecific2023}
Murphy, K. B., Gordon-Smith, R., Chapman, J., Otani, M., Schilder, B.
M., \& Skene, N. G. (n.d.). Identification of cell type-specific gene
targets underlying thousands of rare diseases and subtraits.
\url{https://doi.org/10.1101/2023.02.13.23285820}

\bibitem[\citeproctext]{ref-noriCanGeneralistFoundation2023}
Nori, H., Lee, Y. T., Zhang, S., Carignan, D., Edgar, R., Fusi, N., et
al. (n.d.). Can generalist foundation models outcompete special-purpose
tuning? Case study in medicine.
\url{https://doi.org/10.48550/arXiv.2311.16452}

\bibitem[\citeproctext]{ref-oneilPhenomicsAssistantInterface2024}
O'Neil, S. T., Schaper, K., Elsarboukh, G., Reese, J. T., Moxon, S. A.
T., Harris, N. L., et al. (n.d.). Phenomics assistant: An interface for
LLM-based biomedical knowledge graph exploration.
\url{https://doi.org/10.1101/2024.01.31.578275}

\bibitem[\citeproctext]{ref-ochoaOpenTargetsPlatform2021}
Ochoa, D., Hercules, A., Carmona, M., Suveges, D., Gonzalez-Uriarte, A.,
Malangone, C., et al. (2021). Open targets platform: Supporting
systematic drug--target identification and prioritisation. \emph{Nucleic
Acids Research}, \emph{49}(D1), D1302--D1310.
\url{https://doi.org/10.1093/nar/gkaa1027}

\bibitem[\citeproctext]{ref-openaiGPT4TechnicalReport2024}
OpenAI, Achiam, J., Adler, S., Agarwal, S., Ahmad, L., Akkaya, I., et
al. (n.d.). GPT-4 technical report.
\url{https://doi.org/10.48550/arXiv.2303.08774}

\bibitem[\citeproctext]{ref-panLargeLanguageModels2023}
Pan, J. Z., Razniewski, S., Kalo, J.-C., Singhania, S., Chen, J.,
Dietze, S., et al. (n.d.). Large language models and knowledge graphs:
Opportunities and challenges.
\url{https://doi.org/10.48550/arXiv.2308.06374}

\bibitem[\citeproctext]{ref-prietoProblemsSolutionsCalculating2003}
Prieto, L., \& Sacristán, J. A. (2003). Problems and solutions in
calculating quality-adjusted life years (QALYs). \emph{Health and
Quality of Life Outcomes}, \emph{1}, 80.
\url{https://doi.org/10.1186/1477-7525-1-80}

\bibitem[\citeproctext]{ref-putmanMonarchInitiative20242024}
Putman, T. E., Schaper, K., Matentzoglu, N., Rubinetti, V. P.,
Alquaddoomi, F. S., Cox, C., et al. (2024). The monarch initiative in
2024: An analytic platform integrating phenotypes, genes~and diseases
across species. \emph{Nucleic Acids Research}, \emph{52}(D1),
D938--D949. \url{https://doi.org/10.1093/nar/gkad1082}

\bibitem[\citeproctext]{ref-shinBioMegatronLargerBiomedical2020}
Shin, H.-C., Zhang, Y., Bakhturina, E., Puri, R., Patwary, M., Shoeybi,
M., \& Mani, R. (2020). BioMegatron: Larger biomedical domain language
model. In B. Webber, T. Cohn, Y. He, \& Y. Liu (Eds.), \emph{Proceedings
of the 2020 conference on empirical methods in natural language
processing (EMNLP)} (pp. 4700--4706). Association for Computational
Linguistics. \url{https://doi.org/10.18653/v1/2020.emnlp-main.379}

\bibitem[\citeproctext]{ref-singhalExpertLevelMedicalQuestion2023}
Singhal, K., Tu, T., Gottweis, J., Sayres, R., Wulczyn, E., Hou, L., et
al. (n.d.). Towards expert-level medical question answering with large
language models. \url{https://doi.org/10.48550/arXiv.2305.09617}

\bibitem[\citeproctext]{ref-singhalLargeLanguageModels2023}
Singhal, K., Azizi, S., Tu, T., Mahdavi, S. S., Wei, J., Chung, H. W.,
et al. (2023). Large language models encode clinical knowledge.
\emph{Nature}, 1--9. \url{https://doi.org/10.1038/s41586-023-06291-2}

\bibitem[\citeproctext]{ref-toroDynamicRetrievalAugmented2023}
Toro, S., Anagnostopoulos, A. V., Bello, S., Blumberg, K., Cameron, R.,
Carmody, L., et al. (n.d.). Dynamic retrieval augmented generation of
ontologies using artificial intelligence (DRAGON-AI).
\url{https://doi.org/10.48550/arXiv.2312.10904}

\bibitem[\citeproctext]{ref-vanveenAdaptedLargeLanguage2024}
Van Veen, D., Van Uden, C., Blankemeier, L., Delbrouck, J.-B., Aali, A.,
Bluethgen, C., et al. (2024). Adapted large language models can
outperform medical experts in clinical text summarization. \emph{Nature
Medicine}, 1--9. \url{https://doi.org/10.1038/s41591-024-02855-5}

\bibitem[\citeproctext]{ref-weiLongformFactualityLarge2024}
Wei, J., Yang, C., Song, X., Lu, Y., Hu, N., Huang, J., et al. (n.d.).
Long-form factuality in large language models. Retrieved from
\url{http://arxiv.org/abs/2403.18802}

\bibitem[\citeproctext]{ref-zhangBiomedGPTUnifiedGeneralist2023}
Zhang, K., Yu, J., Yan, Z., Liu, Y., Adhikarla, E., Fu, S., et al.
(n.d.). BiomedGPT: A unified and generalist biomedical generative
pre-trained transformer for vision, language, and multimodal tasks.
\url{https://doi.org/10.48550/arXiv.2305.17100}

\end{CSLReferences}

\newpage{}

\subsection{Supplementary Materials}\label{supplementary-materials}

\subsubsection{Supplementary Figures}\label{supplementary-figures}

\phantomsection\label{cell-fig-severity-histo}
\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-pdf/fig-severity-histo-1.pdf}

}

\caption{\label{fig-severity-histo}Distribution of the composite GPT-4
severity score of the severity scores for all HPO terms.}

\end{figure}%

\phantomsection\label{cell-fig-severity-boxplot}
\begin{figure}[H]

\centering{

\includegraphics{manuscript_files/figure-pdf/fig-severity-boxplot-1.pdf}

}

\caption{\label{fig-severity-boxplot}Boxplot showing the relationship
between composite severity score (y-axis) and the frequency response
categories within each annotation type.}

\end{figure}%

\newpage{}

\subsubsection{Supplementary Tables}\label{supplementary-tables}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1379}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0632}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7989}}@{}}

\caption{\label{tbl-fas}Severity nnotations generated for GPT-4
annotations for the HPO phenotype `Fetal akinesia sequence'
(HP:000198).}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Clinical characteristic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Annotation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Justification
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intellectual disability & always & Fetal akinesia sequence typically
results in severe neurodevelopmental impairment, including intellectual
disability. \\
Death & always & Fetal akinesia sequence is typically fatal in utero or
shortly after birth. \\
Impaired mobility & always & Fetal akinesia sequence results in severe
physical impairment, including impaired mobility. \\
Physical malformations & always & Fetal akinesia sequence is associated
with multiple physical malformations. \\
Blindness & often & Visual impairment is common in surviving individuals
with fetal akinesia sequence due to neurodevelopmental impairment. \\
Sensory impairments & always & Fetal akinesia sequence typically results
in severe sensory impairment due to neurodevelopmental disruption. \\
Immunodeficiency & rarely & While not a direct feature, some individuals
with fetal akinesia sequence may have associated immune
abnormalities. \\
Cancer & never & Fetal akinesia sequence does not cause cancer. \\
Reduced fertility & often & Given the severe physical impairments
associated with fetal akinesia sequence, fertility is likely to be
reduced in surviving individuals. \\
Congenital onset & always & Fetal akinesia sequence is a congenital
disorder. \\

\end{longtable}

\end{document}

### 2024-05-14 docs/_tex/manuscript.tex
Update manuscript to address issues: #1 #2 #3 #4 #5 #6 #7 #8 #9

@@
 human clinicians on standardised medical exams and patient diagnosis
-tasks (Bolton et al., n.d.; Cheng et al., 2023; Gu et al., 2021; Labrak
-et al., n.d.; Luo et al., 2022; McDuff et al., n.d.; O'Neil et al.,
-n.d.; Shin et al., 2020; Singhal et al., n.d., 2023, 2023; Van Veen et
-al., 2024; Zhang et al., n.d.). Recent work has demonstrated that the
-Generative Pre-trained Transformer 4 (GPT-4) foundation model (OpenAI et
-al., n.d.), when combined with strategic prompt engineering, can
-outperform even specialist LLMs that are explicitly fine-tuned for
-biomedical tasks (Nori et al., n.d.). In a landmark achievement, GPT-4
-was the first LLM to surpass a score of 90\% in the United States
-Medical Licensing Examination (USML) (Nori et al., n.d.).
+tasks (Bolton et al., 2024; Cheng et al., 2023; Gu et al., 2021; Labrak
+et al., 2024; Luo et al., 2022; McDuff et al., 2023; O'Neil et al.,
+2024; Shin et al., 2020; Singhal, Azizi, et al., 2023, 2023; Singhal,
+Tu, et al., 2023; Van Veen et al., 2024; Zhang et al., 2023). Recent
+work has demonstrated that the Generative Pre-trained Transformer 4
+(GPT-4) foundation model (OpenAI et al., 2024), when combined with
+strategic prompt engineering, can outperform even specialist LLMs that
+are explicitly fine-tuned for biomedical tasks (Nori et al., 2023). In a
+landmark achievement, GPT-4 was the first LLM to surpass a score of 90\%
+in the United States Medical Licensing Examination (USML) (Nori et al.,
+2023).
 
@@
 
-To assess annotation consistency, 793 phenotypes underwent duplicate
-annotations. We employed two different metrics to determine the
-\emph{consistency rate}. The first, less stringent metric, defined
-consistency as the duplicate annotations being either `always' and
-`often', or `never' and `rarely'. The second, more stringent metric,
+To assess annotation consistency, 793 we queried GPT-4 with a subset of
+the HPO phenotypes multiple times. We employed two different metrics to
+determine the \emph{consistency rate}. The first, less stringent metric,
+defined consistency as the duplicate annotations being either `always'
+and `often', or `never' and `rarely'. The second, more stringent metric,
 required exact agreement in annotation occurrences, e.g.~`always' and
@@
 Despite specifying in the prompt for GPT-4 not to take into
-consideration indirect effects, here is an example of where it failed to
+consideration indirect effects, this is an example of where it failed to
 do so.
+
+We also reasoned that GPT-4 would be better able to give consistent
+answers for more specific phenotypes lower in the ontology, as they are
+more likely to have a single cause. We found that the stringent
+consistency rate did indeed significantly improve with greater HPO
+ontology depth (\(X_{Pearson}^2\)=22.17, \(\hat{V}_{Cramer}\)=0.03,
+p=0.05). See Figure~\ref{fig-consist-vs-ontLvl} for a visual
+representation of this relationship.
 
@@
 
+\subsubsection{Severity classes}\label{severity-classes}
+
+While the continuous severity score is a helpful metric, there may be
+some use cases where a categorical classification of severity is more
+immediately useful. In work by Lazarin et al. (2014), the authors
+defined severity classed using a simple decision tree based on the
+individual severity annotations. We approximated this approach using our
+GPT-4 annotations. This categorical approach showed a strong degree of
+positive correspondence with the continuous severity score
+(\(\hat{\omega_{p}^2}\)=0.87, p\textless2.2e-308). In other words,
+severity score increased with severity class level (mild \textless{}
+moderate \textless{} severe \textless{} profound) as expected. The
+distribution of severity classes is shown in
+Figure~\ref{fig-severity-class}.
+
+\subsubsection{Correlations between severity
+metrics}\label{correlations-between-severity-metrics}
+
+We found that some severity metrics were correlated with one another,
+with a mean Pearson correlation of 0.2 across all individual metrics
+(see Figure~\ref{fig-metric-corplot}). In particular, blindness and
+sensory impairment were highly correlated with one another (r=0.62,
+p=0). Some metrics drove the composite severity score more than other,
+which is a reflection of both our per-metric weighting scheme, response
+type frequencies, and the correlation structure between metrics.
+Overall, impaired mobility seemed to be the strongest driver of the
+composite severity score with a Pearson correlation of 0.6138757,
+followed by intellectual disability (r=0.6) and sensory impairments
+(r=0.54).
+
 \subsection{Discussion}\label{discussion}
@@
 encode, retrieve, and synthesise vast amounts of diverse information in
-a highly scalable manner (OpenAI et al., n.d.; Singhal et al., 2023; Van
-Veen et al., 2024). This makes them powerful tools that can be applied
-in a rapidly expanding variety of scenarios, including medical practice,
-research and data curation (Caufield et al., n.d.; O'Neil et al., n.d.;
-Pan et al., n.d.; Singhal et al., 2023; Toro et al., n.d.).
+a highly scalable manner (OpenAI et al., 2024; Singhal, Azizi, et al.,
+2023; Van Veen et al., 2024). This makes them powerful tools that can be
+applied in a rapidly expanding variety of scenarios, including medical
+practice, research and data curation (Caufield et al., 2023; O'Neil et
+al., 2024; Pan et al., 2023; Singhal, Azizi, et al., 2023; Toro et al.,
+2023).
 
 Here, we introduce a novel framework to leverage the current
-best-in-class LLM, GPT-4 (OpenAI et al., n.d.), to systematically
+best-in-class LLM, GPT-4 (OpenAI et al., 2024), to systematically
 annotate the severity of 17502 phenotypic abnormalities within the HPO.
@@
 information about previous requests and interactions as they use these
-to iteratively improve and update their predictions (Janik, n.d.). This
+to iteratively improve and update their predictions (Janik, 2024). This
 represents a categorical advance over traditional natural language
@@
 learning (e.g.~Term Frequency-Inverse Document Frequency (Jones, 1972),
-Word2vec (Mikolov et al., n.d.)) which lack the ability to provide
+Word2vec (Mikolov et al., 2013)) which lack the ability to provide
 chains of causal reasoning to justify their predictions. This highlights
@@
 deeper more complex models with low explainability but high
-interpretability (Marcinkevičs \& Vogt, n.d.).
+interpretability (Marcinkevičs \& Vogt, 2023).
 
@@
 be caused by at least 24 different genetic disorders. Despite the
-complex and heterogenous aetiology of this phenotype, GPT-4 was able to
+complex and heterogeneous aetiology of this phenotype, GPT-4 was able to
 provide accurate annotations alongside explainable justifications for
@@
 results have a large number of useful real-world applications, such as
-prioritising gene therapy candidates (Murphy et al., n.d.) and guiding
+prioritising gene therapy candidates (Murphy et al., 2023) and guiding
 clinical decision-making in rare diseases. As the HPO continues to grow
@@
 performance due to long-form text input, which is presently a known
-limitation common to many LLMs including GPT-4 (Wei et al., n.d.).
+limitation common to many LLMs including GPT-4 (Wei et al., 2024).
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1194}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7711}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1095}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4000}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}@{}}
 
@@
 \endlastfoot
-Intellectual disability & Intellectual disability; Mental deterioration
-& 19 \\
-Impaired mobility & Gait disturbance; Diminished movement; mobility &
-319 \\
-Physical malformations & malformation & 78 \\
-Blindness & blindness & 1 \\
-Sensory impairments & Abnormality of vision; Abnormality of the sense of
-smell; Abnormality of taste sensation; Somatic sensory dysfunction;
-Hearing abnormality & 252 \\
-Immunodeficiency & Immunodeficiency; Impaired antigen-specific response
-& 29 \\
-Cancer & Cancer; malignant; carcinoma & 56 \\
-Reduced fertility & Decreased fertility; Hypogonadism & 9 \\
+Intellectual disability & `Intellectual disability'; `Mental
+deterioration' & 19 \\
+Impaired mobility & `Gait disturbance'; `Diminished movement'; '
+mobility' & 319 \\
+Physical malformations & `malformation' & 78 \\
+Blindness & `blindness' & 1 \\
+Sensory impairments & `Abnormality of vision'; `Abnormality of the sense
+of smell'; `Abnormality of taste sensation'; `Somatic sensory
+dysfunction'; `Hearing abnormality' & 252 \\
+Immunodeficiency & `Immunodeficiency'; `Impaired antigen-specific
+response' & 29 \\
+Cancer & `Cancer'; `malignant'; `carcinoma' & 56 \\
+Reduced fertility & `Decreased fertility'; `Hypogonadism' & 9 \\
 
@@
 \item
-  \(NSS_p\): The final composite severity score for phenotype \(p\)
+  \(NSS_p\): the final composite severity score for phenotype \(p\)
   after applying normalisation to align values to a 0-100 scale and
@@
 
+\subsubsection{Severity classes}\label{severity-classes-1}
+
+The decision tree algorithm used in Lazarin et al. (2014) was adapted
+here for use with the GPT-4 annotations. This algorithm first assigned
+each annotation metric to a tier, where Tier 1 indicated the most severe
+annotation metrics and Tier 4 indicated the least severe annotation
+metrics (`death'=1, `intellectual disability'=1, `impaired mobility'=2,
+`physical malformations'=2, `blindness'=3, `sensory impairments'=3,
+`immunodeficiency'=3, `cancer'=3, `reduced fertility'=4). If a phenotype
+often or always caused more than one Tier 1 annotation, it was assigned
+a severity class of ``Profound''. If the phenotype often or always
+caused only one Tier 1 annotation, it was assigned a severity class of
+``Severe''. A ``Severe'' class assignment was also assigned if the
+phenotype often or always caused three or more Tier 2 and Tier3
+annotations. If the phenotype often or always caused at least one Tier 2
+annotation, it was assigned a severity class of ``Moderate''. All
+remaining phenotypes were was assigned a severity class of ``Mild''. In
+cases where the phenotype mapped to more than one class, only the most
+severe class was used. This procedure is implemented within the function
+\texttt{HPOExplorer::gpt\_annot\_class}.
+
+\subsubsection{Correlations between severity
+metrics}\label{correlations-between-severity-metrics-1}
+
+To assess the correlation structure between each severity metric, as
+well as between the composite severity score and each metric, we
+computed Pearson correlation coefficients for all pairwise combinations
+of these variables using the numerically encoded metric values. The
+correlation matrix was visualised using a heatmap, with the colour
+intensity representing the strength of the correlation
+(Figure~\ref{fig-metric-corplot}).
+
 \subsection{Data and code availability
@@
 Bolton, E., Venigalla, A., Yasunaga, M., Hall, D., Xiong, B., Lee, T.,
-et al. (n.d.). BioMedLM: A 2.7B parameter language model trained on
-biomedical text. \url{https://doi.org/10.48550/arXiv.2403.18421}
+et al. (2024). BioMedLM: A 2.7B parameter language model trained on
+biomedical text. \emph{arXiv}.
+\url{https://doi.org/10.48550/arXiv.2403.18421}
 
@@
 Caufield, J. H., Hegde, H., Emonet, V., Harris, N. L., Joachimiak, M.
-P., Matentzoglu, N., et al. (n.d.). Structured prompt interrogation and
+P., Matentzoglu, N., et al. (2023). Structured prompt interrogation and
 recursive extraction of semantics (SPIRES): A method for populating
-knowledge bases using zero-shot learning.
+knowledge bases using zero-shot learning. \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.2304.02711}
@@
 \bibitem[\citeproctext]{ref-janikAspectsHumanMemory2024}
-Janik, R. A. (n.d.). Aspects of human memory and large language models.
-\url{https://doi.org/10.48550/arXiv.2311.03839}
+Janik, R. A. (2024). Aspects of human memory and large language models.
+\emph{arXiv}. \url{https://doi.org/10.48550/arXiv.2311.03839}
 
@@
 Labrak, Y., Bazoge, A., Morin, E., Gourraud, P.-A., Rouvier, M., \&
-Dufour, R. (n.d.). BioMistral: A collection of open-source pretrained
-large language models for medical domains.
+Dufour, R. (2024). BioMistral: A collection of open-source pretrained
+large language models for medical domains. \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.2402.10373}
@@
 \bibitem[\citeproctext]{ref-marcinkevicsInterpretabilityExplainabilityMachine2023}
-Marcinkevičs, R., \& Vogt, J. E. (n.d.). Interpretability and
-explainability: A machine learning zoo mini-tour.
+Marcinkevičs, R., \& Vogt, J. E. (2023). Interpretability and
+explainability: A machine learning zoo mini-tour. \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.2012.01805}
@@
 McDuff, D., Schaekermann, M., Tu, T., Palepu, A., Wang, A., Garrison,
-J., et al. (n.d.). Towards accurate differential diagnosis with large
-language models. \url{https://doi.org/10.48550/arXiv.2312.00164}
+J., et al. (2023). Towards accurate differential diagnosis with large
+language models. \emph{arXiv}.
+\url{https://doi.org/10.48550/arXiv.2312.00164}
 
 \bibitem[\citeproctext]{ref-mikolovEfficientEstimationWord2013}
-Mikolov, T., Chen, K., Corrado, G., \& Dean, J. (n.d.). Efficient
-estimation of word representations in vector space.
+Mikolov, T., Chen, K., Corrado, G., \& Dean, J. (2013). Efficient
+estimation of word representations in vector space. \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.1301.3781}
@@
 Murphy, K. B., Gordon-Smith, R., Chapman, J., Otani, M., Schilder, B.
-M., \& Skene, N. G. (n.d.). Identification of cell type-specific gene
+M., \& Skene, N. G. (2023). Identification of cell type-specific gene
 targets underlying thousands of rare diseases and subtraits.
-\url{https://doi.org/10.1101/2023.02.13.23285820}
+\emph{medRxiv}. \url{https://doi.org/10.1101/2023.02.13.23285820}
 
@@
 Nori, H., Lee, Y. T., Zhang, S., Carignan, D., Edgar, R., Fusi, N., et
-al. (n.d.). Can generalist foundation models outcompete special-purpose
-tuning? Case study in medicine.
+al. (2023). Can generalist foundation models outcompete special-purpose
+tuning? Case study in medicine. \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.2311.16452}
@@
 O'Neil, S. T., Schaper, K., Elsarboukh, G., Reese, J. T., Moxon, S. A.
-T., Harris, N. L., et al. (n.d.). Phenomics assistant: An interface for
-LLM-based biomedical knowledge graph exploration.
+T., Harris, N. L., et al. (2024). Phenomics assistant: An interface for
+LLM-based biomedical knowledge graph exploration. \emph{bioRxiv}.
 \url{https://doi.org/10.1101/2024.01.31.578275}
@@
 OpenAI, Achiam, J., Adler, S., Agarwal, S., Ahmad, L., Akkaya, I., et
-al. (n.d.). GPT-4 technical report.
+al. (2024). GPT-4 technical report. \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.2303.08774}
@@
 Pan, J. Z., Razniewski, S., Kalo, J.-C., Singhania, S., Chen, J.,
-Dietze, S., et al. (n.d.). Large language models and knowledge graphs:
-Opportunities and challenges.
+Dietze, S., et al. (2023). Large language models and knowledge graphs:
+Opportunities and challenges. \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.2308.06374}
@@
 
-\bibitem[\citeproctext]{ref-singhalExpertLevelMedicalQuestion2023}
-Singhal, K., Tu, T., Gottweis, J., Sayres, R., Wulczyn, E., Hou, L., et
-al. (n.d.). Towards expert-level medical question answering with large
-language models. \url{https://doi.org/10.48550/arXiv.2305.09617}
-
 \bibitem[\citeproctext]{ref-singhalLargeLanguageModels2023}
@@
 
+\bibitem[\citeproctext]{ref-singhalExpertLevelMedicalQuestion2023}
+Singhal, K., Tu, T., Gottweis, J., Sayres, R., Wulczyn, E., Hou, L., et
+al. (2023). Towards expert-level medical question answering with large
+language models. \emph{arXiv}.
+\url{https://doi.org/10.48550/arXiv.2305.09617}
+
 \bibitem[\citeproctext]{ref-toroDynamicRetrievalAugmented2023}
 Toro, S., Anagnostopoulos, A. V., Bello, S., Blumberg, K., Cameron, R.,
-Carmody, L., et al. (n.d.). Dynamic retrieval augmented generation of
-ontologies using artificial intelligence (DRAGON-AI).
+Carmody, L., et al. (2023). Dynamic retrieval augmented generation of
+ontologies using artificial intelligence (DRAGON-AI). \emph{arXiv}.
 \url{https://doi.org/10.48550/arXiv.2312.10904}
@@
 \bibitem[\citeproctext]{ref-weiLongformFactualityLarge2024}
-Wei, J., Yang, C., Song, X., Lu, Y., Hu, N., Huang, J., et al. (n.d.).
-Long-form factuality in large language models. Retrieved from
-\url{http://arxiv.org/abs/2403.18802}
+Wei, J., Yang, C., Song, X., Lu, Y., Hu, N., Huang, J., et al. (2024).
+Long-form factuality in large language models. \emph{arXiv}. Retrieved
+from \url{http://arxiv.org/abs/2403.18802}
 
@@
 Zhang, K., Yu, J., Yan, Z., Liu, Y., Adhikarla, E., Fu, S., et al.
-(n.d.). BiomedGPT: A unified and generalist biomedical generative
+(2023). BiomedGPT: A unified and generalist biomedical generative
 pre-trained transformer for vision, language, and multimodal tasks.
-\url{https://doi.org/10.48550/arXiv.2305.17100}
+\emph{arXiv}. \url{https://doi.org/10.48550/arXiv.2305.17100}
 
@@
 \subsubsection{Supplementary Figures}\label{supplementary-figures}
+
+\phantomsection\label{cell-fig-consist-vs-ontLvl}
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{manuscript_files/figure-pdf/fig-consist-vs-ontLvl-1.pdf}
+
+}
+
+\caption{\label{fig-consist-vs-ontLvl}Relationship between the
+consistency of GPT-4 annotations (using the stringent criterion) and the
+level of each phenotype within the HPO ontology (with the number of
+phenotypes in parentheses). Greater ontology levels (x-axis) indicate
+more specific phenotypes. The subtitle indicates summary statistics for
+the overall relationship between HPO level and the proportion of
+phenotypes that were annotated consistently. The p-values above each bar
+indicate whether the distribution of consistent/inconsistent
+annotations, within a given HPO level, significantly deviate from the
+expected null distribution.}
+
+\end{figure}%
 
@@
 
+\phantomsection\label{cell-fig-metric-corplot}
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{manuscript_files/figure-pdf/fig-metric-corplot-1.pdf}
+
+}
+
+\caption{\label{fig-metric-corplot}Pearson correlations between each
+individual severity metric and the composite severity score
+(`severity\_score\_gpt').}
+
+\end{figure}%
+
+\phantomsection\label{cell-fig-severity-class}
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{manuscript_files/figure-pdf/fig-severity-class-1.pdf}
+
+}
+
+\caption{\label{fig-severity-class}Distribution of the composite GPT-4
+severity score introduced in this paper (y-axis) by an approximation of
+the severity class system introduced Lazarin et al. (2014) (x-axis).
+While these are different schemes for ranking phenotype severity, there
+is a strong correspondence between them (see summary statistics in
+subtitle). The sample size (number of phenotypes) is shown in
+parentheses along the x-axis.}
+
+\end{figure}%
+
 \newpage{}
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1379}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0632}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7989}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.6000}}@{}}
 

### 2024-05-14 manuscript.qmd
Update manuscript to address issues: #1 #2 #3 #4 #5 #6 #7 #8 #9

@@
 keywords:
-  - HPO
-  - chatGPT
-  - rare disease 
+  - Ontology
+  - Human Phenotype Ontology
+  - Large Language Model
+  - GPT-4
+  - Artificial intelligence
+  - Generative AI
+  - Rare disease 
+  - Automation
+  - Medicine
 bibliography: references.bib
+date: last-modified
+citation:
+  container-title: medRxiv
 ---
 
-```{r setup}
+```{r setup, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
 library(dplyr)
 library(ggplot2)
+library(ggtext)
 library(patchwork)
 library(HPOExplorer)
+library(KGExplorer)
+library(data.table)
+library(Hmisc)
+library(corrplot)
+library(ggstatsplot)
+options("knitr.table.format"="pipe")
+```
+
+```{r import-data}
+#### Load HPO data
+tag <- "v2024-02-08"
+save_dir <- "data"
+keep_descendants <- "Phenotypic abnormality"
+p2g <- HPOExplorer::load_phenotype_to_genes(tag = tag,
+                                            save_dir=save_dir)
+hpo <- HPOExplorer::get_hpo(method="github",
+                            tag=tag, 
+                            save_dir=save_dir)
+hpo <- KGExplorer::filter_ontology(ont = hpo, 
+                                   keep_descendants = keep_descendants)
+#### Read HPO GPT annotations 
+gpt_annot <- HPOExplorer::gpt_annot_read(hpo = hpo)
+#### Compute IDs counts
+ids_matched <- intersect(c(hpo@terms,names(hpo@alternative_terms)), 
+                         gpt_annot$hpo_id)
+names_matched <- intersect(hpo@elementMetadata$name,
+                           trimws(gsub("obsolete","",gpt_annot$hpo_name, ignore.case = TRUE)))
+
+n_ids_hpo <- hpo@n_terms
+n_ids_gpt <- length(unique(data.table::fcoalesce(gpt_annot$hpo_name,gpt_annot$hpo_id)))
+```
+
+```{r gpt_annot_check}
+query_hits <- HPOExplorer::search_hpo(search_cols = c("name"),
+                                      hpo = hpo)
+#### Run annotation checks
+checks <- HPOExplorer::gpt_annot_check(annot = gpt_annot, 
+                                       query_hits = query_hits)
+# checks$true_pos_rate
+```
+
+```{r gpt_annot_codify}
+res_coded <- HPOExplorer::gpt_annot_codify(annot = gpt_annot)
+annot_melt <- HPOExplorer::gpt_annot_melt(res_coded = res_coded)
+```
+
+```{r gpt_annot_class, results='hide'}
+res_class <- HPOExplorer::gpt_annot_class(annot_coded = res_coded$annot_coded)
+res_class <- merge(res_class,
+                   res_coded$annot_weighted[,c("hpo_name","severity_score_gpt")])
+fig_severity_class <- ggstatsplot::ggbetweenstats(res_class,
+                            x="severity_class", y="severity_score_gpt")
+fig_severity_class_dt <- MSTExplorer:::get_ggstatsplot_stats(fig_severity_class)
+```
+
+```{r gpt_annot_plot}
+gpt_annot_plot_out <- HPOExplorer::gpt_annot_plot(
+  annot = gpt_annot[hpo_name!=keep_descendants],
+  keep_descendants = keep_descendants, 
+  width = 10)
 ```
@@
 
-### Annotating the HPO using OpenAI GPT-4
-We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO).
-Our annotation framework was developed based on previously defined criteria for classifying disease severity @Lazarin2014-rz.
-We sought to evaluate the impact of phenotypes on outcomes including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
-Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
-Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance was achieved when inputting no more than two or three phenotypes per prompt.
-An example prompt can be seen in Figure 1.
-The phenotype outcome occurrence varied by category.
-While > 50% of annotated phenotypes never caused blindness, cancer, immunodeficiency, intellectual disability, reduced fertility, and sensory impairment, a similar proportion often or always had congenital onset (Fig.
-1).
+### Annotating the HPO using GPT-4
 
-```{r occurence_plot}
-#| echo: false
-#| warning: false
-
-# read in hpo gpt annotations 
-hpo_annot <- HPOExplorer::gpt_annot_read()
+```{r fig-occurrence}
+#| label: fig-occurrence
+#| fig-cap: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics.
+**a**, An example of the prompt input given to to GPT-4.
+The phenotypes listed in the second to last sentence (*italicised*) were changed to allow all HPO phenotypes to be annotated.
+**b**, Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all annotated HPO phenotypes.
+The terms shown on the x-axis are the clinical characteristics for which GPT-4 was asked to determine whether each phenotype caused them.
+#| fig-height: 6
+#| fig-width: 10
 
 # subset to clinical characteristics 
-occurr_df <- hpo_annot[,c("hpo_name", "intellectual_disability", "death", "impaired_mobility", "blindness",
-                          "sensory_impairments", "immunodeficiency", "cancer", "reduced_fertility",
-                          "congenital_onset")]
-
+annotation_order <- c("death", "intellectual disability", "impaired mobility", "blindness", 
+                      "congenital onset", "physical malformations", "sensory impairments", 
+                      "immunodeficiency", "cancer", "reduced fertility")
+occurr_df <- gpt_annot[,c("hpo_name",gsub(" ", "_", annotation_order)),with=FALSE]
 cols_of_interest <- names(occurr_df)[-1]
-
 # Create an empty dataframe to store the counts
@@
 # set annotation order for plots 
-annotation_order <- c("death", "intellectual disability", "impaired mobility", "blindness", 
-                      "congenital onset", "physical malformations", "sensory impairments", 
-                      "immunodeficiency", "cancer", "reduced fertility")
+count_df_long$category <- factor(count_df_long$category, 
+                                 levels = annotation_order)
 
-count_df_long$category <- factor(count_df_long$category, levels = annotation_order)
+# Load the prompt generation function 
+source("code/prompt_gen.R")
+prompts <- prompt_gen_table(hpo = hpo)
+txt_prompt <- prompts$prompt[1688]
+for(x in prompts$terms[[1688]]){
+  txt_prompt <- sub(x,paste0("*",x,"*"),txt_prompt)
+}
+for(x in c(
+  eval(formals(prompt_gen_table)$effects),
+  eval(formals(prompt_gen_table)$responses),
+  "justification")){
+  txt_prompt <- sub(x,paste0("**",x,"**"),txt_prompt)
+}
+## Subplot a: Example prompt
+f1a <- ggplot() + 
+  labs(title="Example GPT-4 prompt") +
+  ggtext::geom_textbox(aes(x=.5,y=.7),
+                       box.padding = margin(20, 10, 20, 10),
+                       box.margin = margin(0, 0, 0, 0),
+                       box.r = grid::unit(30, "pt"),
+                       width = unit(.8, "npc"),
+                       size=3,
+                       halign=.5,
+                       fill="grey90",
+                       label=txt_prompt) +
+  ylim(c(0,1)) +
+  theme_void() +
+  theme(plot.title = element_text(hjust=0.5))
 
-# Stacked bar plot of clinical characteristic occurrence
-ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
+## Subplot b: Stacked bar plot of clinical characteristic occurrence.
+f1b <- ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
   geom_bar(stat = "identity") +
   labs(title = "Clinical characteristic occurrence",
-       x = "", y = "HPO phenotypes (n)", fill = NULL) +
+       x = NULL, y = "HPO phenotypes (n)", fill = NULL) +
   theme_minimal() +
-  theme(legend.position = "right", plot.title = element_text(hjust=0.5), 
-        text = element_text(size=16),
-        axis.text.x = element_text(angle=45, size=16, hjust=1), 
-        title = element_text(size=15)) +
+  theme(legend.position = "right", 
+        plot.title = element_text(hjust=0.5, vjust=-1),  
+        axis.text.x = element_text(angle=45, hjust=1),  
+        ) +
   scale_fill_brewer(palette = "GnBu", direction = -1)
 
+(
+  f1a| 
+    (f1b/plot_spacer() ) + plot_layout(heights = c(1, .1))
+) + 
+  patchwork::plot_layout(widths = c(1,1)) +
+  patchwork::plot_annotation(tag_levels = letters) 
 ```
 
-![Figure 1: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics.
-a An example of the prompt input given to to GPT-4.
-The phenotypes listed in the last sentence were changed to allow ~17,000 phenotypes to be annotated.
-b Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all ~17,000 phenotypes annotated.The terms shown on the X-axis are the clinical characteristics, for which GPT-4 was asked to determine whether each phenotype caused them.
-For example, it annotated ‘Respiratory arrest’ as always causing death.
-Only 75 phenotypes were annotated as always causing death, despite 9,544 phenotypes often or rarely causing it.](figures/clinical_characteristic_occurrence.png)
-### Annotation consistency and accuracy
-To assess annotation consistency, 793 phenotypes underwent duplicate annotations, with two metrics employed to determine the 'consistency rate'.
+We employed the OpenAI GPT-4 model with Python to annotate `r length(unique(gpt_annot$hpo_name))` terms within the HPO (`r tag`) [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
+Our annotation framework was developed based on previously defined criteria for classifying disease severity [@Lazarin2014-rz].
+We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
+Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
+For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
+
+```{r always_death}
+never_thresh <- 50
+never_dt <- subset(count_df_long, condition=="never" & percentage>=never_thresh)
+always_death <- gpt_annot[,list(always_death=all(death=="always"),
+                                hpo_name=hpo_name[1]),
+                          by="hpo_id"][always_death==TRUE,]
+```
+
+Phenotype outcome occurrence varied across annotation categories.
+\>`r never_thresh`% of phenotypes never caused `r paste(never_dt$category[-1], collapse=", ")` or `r never_dt$category[1]`.
+Only a minority of phenotypes (`r round(sum(subset(count_df_long, condition %in% c("never") & category=="congenital onset")$percentage),1)`%) never had a congenital onset, which is expected as rare disorders tend to be early onset genetic conditions [Fig.
+@fig-occurrence].
+
+Less than `r ceiling(subset(count_df_long, condition=="always" & category=="death")$percentage)`% of phenotype always directly resulted in death, such as 'Stillbirth', 'Anencephaly' and 'Bilateral lung agenesis'.
+Only `r uniqueN(gpt_annot[death=="always"]$hpo_name)` phenotypes were annotated as always causing death, while `r uniqueN(gpt_annot[death %in% c("often","rarely")]$hpo_name)` phenotypes were annotated as often or rarely causing death.
+`r uniqueN(gpt_annot[death %in% c("never")]$hpo_name)` phenotypes were annotated as never causing death.
+This included `r sum(grepl("syndactyly",gpt_annot[death %in% c("never")]$hpo_name, ignore.case = TRUE))` unique forms of syndactyly, a non-lethal condition that causes fused or webbed fingers (occurring 1 in 1,200–15,000 live births).
+While not life-threatening itself, syndactyly is a symptom of genetic disorders that can cause life-threatening cardiovascular and neurodevelopmental defects, such as Apert Syndrome [@garagnaniSyndromesAssociatedSyndactyly2013].
+This example highlights the ability of GPT-4 to successfully distinguish between phenotypes that directly cause lethality, and those that are often associated with diseases that cause lethality.
+
+### Annotation consistency and recall
+
+```{r ontLvl-vs-consistency}
+consist <- checks$annot_stringent_mean
+consist <- HPOExplorer::add_hpo_id(consist, hpo = hpo)
+consist <- HPOExplorer::add_ont_lvl(consist, hpo = hpo)
+# consist_vs_lvl <- Hmisc::rcorr(as.matrix(consist[!is.na(ontLvl)][,-c("hpo_name","hpo_id")]))
+# corrplot::corrplot(consist_vs_lvl$r)
+
+consist_vs_ontLvl <- consist|>
+  data.table::melt.data.table(id.vars = c("hpo_name","hpo_id","ontLvl"), 
+                              variable.name = "metric", 
+                              value.name = "Consistent\n(stringent)")|>
+  # ggstatsplot::ggscatterstats(x = "consistency",
+  #                             y = "ontLvl")
+  # ggstatsplot::ggbetweenstats(x = "consistency", 
+  #                             y = "ontLvl", type = "parametric")
+  ggstatsplot::ggbarstats(x = "Consistent\n(stringent)", 
+                          y = "ontLvl") +
+  ggplot2::labs(title = "Consistency of GPT-4 annotations by HPO level")
+consist_vs_ontLvl_dt <- MSTExplorer:::get_ggstatsplot_stats(consist_vs_ontLvl)
+```
+
+To assess annotation consistency, `r checks$consistency_count[[1]]` we queried GPT-4 with a subset of the HPO phenotypes multiple times.
+We employed two different metrics to determine the *consistency rate*.
 The first, less stringent metric, defined consistency as the duplicate annotations being either 'always' and 'often', or 'never' and 'rarely'.
@@
 'always' and 'always'.
-For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least 80%, and for the more stringent metric, the lowest consistency rate was 57%, for congenital onset.
-An example of where annotations were inconsistent was for the HPO term ‘Acute leukaemia’.
-One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that ‘weakness and fatigue from leukaemia and its treatment can impair mobility’.
-The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that ‘acute leukaemia rarely impairs mobility directly’.
-Despite specifying in the prompt for GPT-4 not to take into consideration indirect effects, here is an example of where it failed to do so.
+For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least `r floor(min(checks$consistency_rate)*100)`%, and for the more stringent metric, the lowest consistency rate was `r round(min(checks$consistency_stringent_rate)*100)`% for congenital onset.
+An example of where annotations were inconsistent was for the HPO term 'Acute leukaemia'.
+One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that 'weakness and fatigue from leukaemia and its treatment can impair mobility'.
+The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that 'acute leukaemia rarely impairs mobility directly'.
+Despite specifying in the prompt for GPT-4 not to take into consideration indirect effects, this is an example of where it failed to do so.
+
+We also reasoned that GPT-4 would be better able to give consistent answers for more specific phenotypes lower in the ontology, as they are more likely to have a single cause.
+We found that the stringent consistency rate did indeed significantly improve with greater HPO ontology depth ($X_{Pearson}^2$=`r round(consist_vs_ontLvl_dt$summary_data$statistic,2)`, $\hat{V}_{Cramer}$=`r round(consist_vs_ontLvl_dt$summary_data$estimate,2)`, p=`r format(consist_vs_ontLvl_dt$summary_data$p.value, digits = 1)`).
+See @fig-consist-vs-ontLvl for a visual representation of this relationship.
 
 ```{r annot_checks}
-# Annotation checks
-checks <- HPOExplorer::gpt_annot_check()
-
 # Define items to be checked
@@
 Based on these scores we evaluated the top 50 severe phenotypes.
-The most severe phenotype, with a score of 58, was ‘Anencephaly’ (HP:0002323).
+One of the most severe phenotype was `r shQuote(anenceph_dt$hpo_name)` (`r anenceph_dt$hpo_id`) with a composite severity score of `r round(anenceph_dt$severity_score_gpt, 1)`.
 Anencephaly is a birth defect where the baby is born without a portion of its brain and skull, often these babies are stillborn.
 In fact, many of the most severe phenotypes were related to developmental brain and neural tube defects.
-Comparison of the severity scores for each response, across the phenotype outcomes annotated, revealed consistent trends: as the response of the phenotype outcome increased (from never to always), the severity score also increased (Supplementary Fig.
-1).
-We also evaluated the severity score distribution by HPO branch and calculated the mean severity score using all phenotypes within a HPO branch (Fig.
-4).
-After the term ‘Phenotypic abnormality’, the highest severity score was for the HPO branch ‘Abnormal cellular phenotype’, followed by ‘Abnormality of prenatal development or birth’, which would include the high severity scoring phenotypes seen in Fig.
-3.
+Comparison of the severity scores for each response, across the phenotype outcomes annotated, revealed consistent trends: as the response of the phenotype outcome increased (from never to always), the severity score also increased ([Supplementary Fig.
+@fig-severity-boxplot]).
+We also evaluated the severity score distribution by HPO branch and calculated the mean severity score using all phenotypes within each major HPO branch ([Fig.
+@fig-severity-histo]).
+The HPO branch with the greatest mean severity score was `r shQuote(unique(gpt_annot_plot_out$data$dat2$ancestor_name)[1])` (mean=`r round(unique(gpt_annot_plot_out$data$dat2$mean_severity_score_gpt)[1],1)`), followed by `r shQuote(unique(gpt_annot_plot_out$data$dat2$ancestor_name)[2])` (mean=`r round(unique(gpt_annot_plot_out$data$dat2$mean_severity_score_gpt)[2],1)`), which would include the highly ranked phenotypes seen in @fig-top-phenos.
 
 ```{r severity_scoring}
+## code adapted from HPOExplorer::gpt_annot_plot()
 # get all descendants of the term phenotypic abnormality 
-keep_descendants="Phenotypic abnormality"
 keep_ont_levels=seq(3,17)
+#### Filter out ont levels ####
+annot_melt <- HPOExplorer::add_ancestor(annot_melt,
+                                        keep_descendants = keep_descendants)
+data.table::setorderv(annot_melt,"severity_score_gpt",-1)
 
-ancestor_name <- variable <- hpo_id <- hpo_name <-
-  value <- severity_score_gpt <- mean_severity_score_gpt <- NULL;
-
-#### Prepare annotation results ####
-gpt_annot <- HPOExplorer::gpt_annot_read()
-res_coded <- HPOExplorer::gpt_annot_codify(annot = gpt_annot)
-dat1 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
-
-#### Filter out ont levels ####
-dat1 <- add_ancestor(dat1,
-                     keep_descendants = keep_descendants)
-data.table::setorderv(dat1,"severity_score_gpt",-1)
-
-dat1 <- HPOExplorer::add_ont_lvl(dat1, keep_ont_levels = keep_ont_levels)
-dat1$variable <- gsub("_", " ", dat1$variable)
+annot_melt <- HPOExplorer::add_ont_lvl(annot_melt, keep_ont_levels = keep_ont_levels)
+annot_melt$variable <- gsub("_", " ", annot_melt$variable)
 
 # get all unique phenotypes
-total_unique <- length(unique(dat1$hpo_id))
+total_unique <- length(unique(annot_melt$hpo_id))
 
 # get the top N phenotypes
-top_n=15
-dat_top <- dat1[hpo_id %in% unique(dat1$hpo_id)[seq(top_n)]]
+top_n <- 15
+dat_top <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(top_n)]]
 
@@
 
-print(p3)
+methods::show(p3)
+```
 
+### Severity classes
+
+While the continuous severity score is a helpful metric, there may be some use cases where a categorical classification of severity is more immediately useful.
+In work by @Lazarin2014-rz, the authors defined severity classed using a simple decision tree based on the individual severity annotations.
+We approximated this approach using our GPT-4 annotations.
+This categorical approach showed a strong degree of positive correspondence with the continuous severity score ($\hat{\omega_{p}^2}$=`r round(fig_severity_class_dt$summary_data$estimate,2)`, p<`r format(.Machine$double.xmin, digits = 2)`).
+In other words, severity score increased with severity class level (mild < moderate < severe < profound) as expected.
+The distribution of severity classes is shown in @fig-severity-class.
+
+### Correlations between severity metrics
+
+```{r cor-metrics}
+cor_metrics <- Hmisc::rcorr(as.matrix(res_coded$annot_weighted[,-c("hpo_id","hpo_name")]))
+cor_composite <- sort(cor_metrics$r["severity_score_gpt",], decreasing = TRUE)[-1]
+```
+
+We found that some severity metrics were correlated with one another, with a mean Pearson correlation of `r round(mean(cor_metrics$r[-nrow(cor_metrics$r),-ncol(cor_metrics$r)]),2)` across all individual metrics (see @fig-metric-corplot).
+In particular, blindness and sensory impairment were highly correlated with one another (r=`r round(cor_metrics$r["blindness","sensory_impairments"],2)`, p=`r round(cor_metrics$P["blindness","sensory_impairments"],4)`).
+Some metrics drove the composite severity score more than other, which is a reflection of both our per-metric weighting scheme, response type frequencies, and the correlation structure between metrics.
+Overall, `r gsub("_"," ",names(cor_composite[1]))` seemed to be the strongest driver of the composite severity score with a Pearson correlation of `r mean(cor_composite[1],2)`, followed by `r gsub("_"," ",names(cor_composite[2]))` (r=`r round(cor_composite[2],2)`) and `r gsub("_"," ",names(cor_composite[3]))` (r=`r round(cor_composite[3],2)`).
+
+```{r, eval=FALSE}
+### Correlation with QALY
+ihme <- data.table::fread("~/Downloads/IHME-GBD_2019_DATA-e678e00e-1/IHME-GBD_2019_DATA-e678e00e-1.csv")
+ihme[,nm:=tolower(cause_name)]
+
+res_coded$annot_weighted[,nm:=tolower(hpo_name)]
+ihme_merged <- merge(data.table::dcast.data.table(ihme, 
+                             formula = "nm~measure_name+metric_name", 
+                             value.var = "val",
+                             fun.aggregate = mean, na.rm=TRUE), 
+                     res_coded$annot_weighted)
+metric_cols <- list(
+  IHME=grep(paste(paste0("_",
+                         "Percent",
+                         # unique(ihme$metric_name)[1],
+                         "$"),collapse = "|"),names(ihme_merged), value = TRUE),
+  GPT=setdiff(names(res_coded$annot_weighted),c("hpo_id","hpo_name","nm"))
+)
+
+Xcor <- cor(ihme_merged[,unlist(metric_cols),with=FALSE], use="complete.obs")
+heatmaply::heatmaply(Xcor[metric_cols$IHME,metric_cols$GPT])
 ```
@@
 
-Here, we present a novel approach leveraging the OpenAI GPT-4 model, to systematically annotate the severity of 17,000 phenotypic abnormalities within the HPO.
-Our findings highlight the potential of natural language processing technologies in significantly contributing to the automation and refinement of the curation process in genomics and rare disease research.
-Manual curation of ontologies has traditionally relied on expert input, a process that is both time-consuming and labor-intensive.
-By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without compromising accuracy.
-Our validation approach yielded a high true positive rate exceeding 73% across the phenotypes tested.
+The creation and annotation of biomedical knowledge has traditionally relied on manual or semi-manual curation by human experts [@putmanMonarchInitiative20242024; @ochoaOpenTargetsPlatform2021; @mungallMonarchInitiativeIntegrative2017a; @kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
+Performing such manual curation and review tasks at scale is often infeasible for human biomedical experts given limited time and resources.
+LLMs have the capacity to effectively encode, retrieve, and synthesise vast amounts of diverse information in a highly scalable manner [@vanveenAdaptedLargeLanguage2024; @singhalLargeLanguageModels2023; @openaiGPT4TechnicalReport2024].
+This makes them powerful tools that can be applied in a rapidly expanding variety of scenarios, including medical practice, research and data curation [@singhalLargeLanguageModels2023; @toroDynamicRetrievalAugmented2023; @panLargeLanguageModels2023; @oneilPhenomicsAssistantInterface2024; @caufieldStructuredPromptInterrogation2023].
+
+Here, we introduce a novel framework to leverage the current best-in-class LLM, GPT-4 [@openaiGPT4TechnicalReport2024], to systematically annotate the severity of `r n_ids_gpt` phenotypic abnormalities within the HPO.
+By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without substantially compromising accuracy.
+Our validation approach yielded a high true positive rate exceeding `r floor(min(checks$true_pos_rate)*100)`% across the phenotypes tested.
 Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.
 
-A key contribution of our study is the development of a severity scoring system that integrates both the nature of the phenotype outcome and the frequency of its occurrence.
-By quantifying the phenotypic severity this way, we highlight the most impactful conditions and provide a comprehensive framework for prioritising gene therapy trials and guiding clinical decision-making in rare diseases.
+Throughout this study, we observed that GPT-4 was capable of reliably recovering deep semantic relationships from the medical domain, far beyond making superficial inferences based on lexical similarities.
+An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt`), which GPT-4 annotated as causing intellectual disability.
+At first glance, we ourselves assumed this was a false positive as the term appeared to be related to dentition.
+However, upon further inspection we realised that molar tooth sign is in fact a pattern of abnormal brain morphology that happens to bear some resemblance to molar dentition when observed in radiological scans.
+This phenotype is a known sign of neurodevelopmental defects that can indeed cause severe intellectual disability [@gleesonMolarToothSign2004].
 
-** Brian to pick examples from severity scoring **
+In addition to rapidly synthesising and summarising vast amounts of information, LLMs can also be steered to provide justifications for each particular response.
+This makes LLMs amenable to direct interrogation as a means of recovering explainability, especially when designed to retain information about previous requests and interactions as they use these to iteratively improve and update their predictions [@janikAspectsHumanMemory2024].
+This represents a categorical advance over traditional natural language processing models based on more shallow forms of statistical or machine learning (e.g.
+Term Frequency-Inverse Document Frequency [@jonesStatisticalInterpretationTerm1972], Word2vec [@mikolovEfficientEstimationWord2013]) which lack the ability to provide chains of causal reasoning to justify their predictions.
+This highlights the fundamental trade-off between simpler models with high explainability (the ability humans to understand the inner workings of the model) but low interpretability (the ability of humans to trace the decision process of the model, analogous to human 'reasoning'), and deeper more complex models with low explainability but high interpretability [@marcinkevicsInterpretabilityExplainabilityMachine2023].
 
-Given its scalability it can be easily applied as more phenotypes are added to the HPO, but can also be extended to include different annotations and to be used with other ontologies.
-For instance, we could have also tried to use GPT-4 to annotate how prevalent a phenotype is in the human population, or what the cost to the health system is of a phenotype.
-Currently there are no systematic annotations of these on a phenotypic level.
-However, it’s unclear whether GPT-4 could currently be applied for such annotations, or whether they require much more extensive data collection.
+A key contribution of our study is the introduction of a quantitative severity scoring system that integrates both the nature of the phenotype outcome and the frequency of its occurrence.
+By encoding the concept of severity in this way, we are able to prioritise phenotypes based on their impact on patients.
+The methodology allowed us to transition from low-throughput qualitative assessments of severity (e.g.
+@Lazarin2014-rz) to high-throughput quantitative assessments of severity.
+One of the most severe phenotypes in the HPO is 'Fetal akinesia sequence' (FAS; `r annot_melt[hpo_name=='Fetal akinesia sequence']$hpo_id[1]`, severity score= `r round(annot_melt[hpo_name=='Fetal akinesia sequence']$severity_score_gpt[1],1)`), and extremely rare condition that is almost always lethal.
+FAS is a complex, multi-system phenotype that can be caused by at least `r uniqueN(p2g[hpo_name=='Fetal akinesia sequence']$disease_id)` different genetic disorders.
+Despite the complex and heterogeneous aetiology of this phenotype, GPT-4 was able to provide accurate annotations alongside explainable justifications for those annotations (see @tbl-fas).
+For example, this phenotype almost always results in death, either *in utero* or shortly after birth.
+Not only did GPT-4 correctly provide the annotation death as `r shQuote(gpt_annot[hpo_name=='Fetal akinesia sequence']$death)`, when asked whether FAS causes sensory impairments it provided the response `r shQuote(gpt_annot[hpo_name=='Fetal akinesia sequence']$sensory_impairments)` with the justification `r shQuote(gpt_annot[hpo_name=='Fetal akinesia sequence']$sensory_impairments_justification)` Neurodevelopmental disruption is indeed a hallmark component of FAS (e.g.
+hydrocephalus, cerebellar hypoplasia) that causes severe impairments across multiple sensory systems [@chenPrenatalDiagnosisGenetic2012].
+This demonstrates that GPT-4 was able to recover the correct chain of causality from phenotype to clinical outcome.
 
-Importantly, there are additional factors that need to be considered when trying to prioritise phenotypes for their suitability for gene therapy development.
-First, our scoring is in some sense subjective, for instance, one could ask is something that always causes death, always worse than something which causes a lifetime of other severe clinical characteristics.
-In terms of viability for a clinical trial, a phenotype which causes a lifetime of expensive burden on the medical system may be a better target than a phenotype which causes a quick death with little or no other clinical characteristics before.
-It would be worth exploring whether GPT-4 could estimate such costs.
-Another factor that affects the viability of developing a treatment for a therapy is how quickly a clinical trial can be run.
-For instance, measuring a high risk of respiratory failure over ten years would require a long trial.
-However, testing for total reversal of an existing severe phenotype would be much faster.
-As AI capabilities continue to advance, perhaps this is something GPT-4 and other models of the like could test for.
+Our findings highlight the potential of this next generation of natural language processing technologies in significantly contributing to the automation and refinement of data curation in biomedical research.
+These results have a large number of useful real-world applications, such as prioritising gene therapy candidates [@murphyIdentificationCellTypespecific2023] and guiding clinical decision-making in rare diseases.
+As the HPO continues to grow over time, this approach can easily be repeated to keep pace with the rapidly evolving biomedical landscape.
+Furthermore, this approach can be extended to produce different sets of annotations or be used with any other ontology.
+Additional use cases include gathering data on the prevalence of each phenotype to approximate their social and financial costs.
 
-An additional annotation that we didn’t take into consideration here is that GPT-4 might be aware of whether there are existing treatments for these phenotypes.
-Therefore, some of the top-ranking severe phenotype we have identified may already have treatment options available.
-It would therefore be useful to know whether a phenotype, although highly severe, is currently treatable, and also whether there exist known clinical endpoints for clinical trial testing.
-GPT-4 does seem to take into account quality of care, i.e.
-the degree to which health services increase the likelihood of desired outcomes (World Health Organization).
-For example, many of the cancer phenotypes are justified as always or often causing death unless treated or caught early.
-On the other hand, some cancers are justified as rarely causing death if appropriate treatment is provided, which may not always be the case for individuals or populations with access to less advanced medical systems.
-In future efforts, we might want to ask GPT-4 more explicitly to distinguish between whether a phenotype would cause death without treatment, or with best available standard of care.
+One key limitation of our study is the fact that we did not explicitly interrogate GPT-4 to assess how the availability of treatments affected the annotations it produced.
+For example, there are some very severe conditions for which highly effective treatments and early detection screens are widely available (e.g.
+syphilis, some forms of melanoma), thus rendering them fully treatable or even curable provided access to modern healthcare.
+It would therefore be useful to further interrogate GPT-4 to uncover how the availability of treatments influences its responses.
+Many of our findings here seem to indicate that GPT-4 does take into account quality of care to the extent that health services increase the likelihood of desired outcomes.
+For example, many of the cancer phenotypes are justified as always or often causing death unless detected and treated early in the disease course.
+On the other hand, some cancers are justified as rarely causing death if appropriate treatment is provided, which may not always be the case for individuals or populations with access to less access to quality healthcare services.
+Future efforts could more explicitly ask GPT-4 whether the phenotype would cause death with no or suboptimal treatment.
+
+In addition to these technical challenges, there are multiple factors that need to be considered when trying to prioritise phenotypes for their suitability for gene therapy development.
+First, while we have attempted to formalise severity here, this is an inherently subjective concept that may vary considerably across different individuals and contexts.
+For instance, one could ask whether a condition that always causes death is worse than a condition that causes a lifetime of severe disability (e.g.
+paralysis, blindness, intellectual disability).
+Metrics such as quality-adjusted life years (QALYs) have been proposed in the past to address these dilemmas by defining health as a function of both the length and quality of life [@prietoProblemsSolutionsCalculating2003].
+With regards to the financial burden of diseases, in some situations phenotypes which require many years of expensive medical care may be prioritised over those that result in extremely early onset lethality and little opportunity for therapeutic intervention.
+Another factor that affects the viability of a therapeutic program is the speed, cost and other practical considerations of a clinical trial.
+For instance, measuring risk of ageing-related respiratory failure over a ten-year period may be impractical in some cases.
+However, testing for total reversal of an existing severe phenotype could potentially yield faster and more immediately impactful results.
+If performed in close collaboration with medical ethicists, governmental organisations, advocacy groups and patient families, such cost/benefit assessments could be aided by LLMs through the scalable gathering of relevant data.
+As AI capabilities continue to advance, the range of applications in which they can be used effectively will continue to grow.
 
@@
 ### Annotating the HPO using OpenAI GPT-4
-We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO).
-Our annotation framework was developed based on previously defined criteria for classifying disease severity (Lazarin et al., 2014).
-We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
-Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
+
+We wrote a Python script to iteratively query GPT-4 via the OpenAI application programming interface (API).
+The ultimately yielded consistently formatted annotations for `r n_ids_gpt` terms within the HPO.
+Our annotation framework was developed based on previously defined criteria for classifying disease severity [@Lazarin2014-rz].
+We sought to evaluate whether each phenotype directly caused a given severity-related clinical characteristic, including: intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and/or had a congenital onset.
+Through prompt engineering we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
+We asked how frequently the given phenotype directly causes each clinical characteristic - whether it never, rarely, often, or always occurred.
+Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance (in terms of producing the desired annotations, and adhering to the formatting requirements) was achieved when inputting no more than two or three phenotypes per prompt.
+An example prompt can be seen in @fig-occurrence.
+Only 2 phenotypes were included per prompt in order to 1) avoid exceeding per-query token limits, and 2) prevent the breakdown of GPT-4 performance due to long-form text input, which is presently a known limitation common to many LLMs including GPT-4 [@weiLongformFactualityLarge2024].
 
 ### Calculating the true positive rate
-A true positive rate was calculated as a measure of the accuracy of the GPT-4 annotations.
+
+```{r tbl-true-positives}
+#| label: tbl-true-positives
+#| tbl-cap: For each clinical characteristic, any terms that matched 
+#| tbl-colwidths: [40,40,20]
+
+queries <- eval(formals(HPOExplorer::search_hpo)$queries)
+query_hits_dt <- data.table(
+  "Clinical characteristic"=stringr::str_to_sentence(gsub("_"," ",names(queries))),
+  "HPO queries"=trimws(gsub("^","",
+                            sapply(queries, function(x){paste(shQuote(x),collapse="; ")}), 
+                            fixed = TRUE)),
+  "True positive HPO IDs"=sapply(query_hits[names(queries)], data.table::uniqueN)
+)
+knitr::kable(as.data.frame(query_hits_dt))
+```
+
+A true positive rate was calculated as a measure of the recall of the GPT-4 annotations.
 This was achieved by identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the occurrence of certain clinical characteristics, and using all descendants of this HPO branch as true positives.
-For example, all descendants of the terms Abnormal central motor function and abnormality of movement should always or often cause impaired mobility (Table 1).
-
-| Clinical characteristics | True positive HPO branch |
-| ------------------------ | ------------------------ |
-| Intellectual disability  | Intellectual disability (HP:0001249)
-| Impaired mobility.       | Abnormal central motor function (HP:0011442) and Abnormality of movement (HP:0100022)
-| Physical malformations   | Malformation (string search)
-| Blindness                | Blindness (HP:0000618)
-| Sensory impairments      | Abnormality of vision (HP:0000504), Abnormality of the sense of smell (HP:0004408), Abnormality of taste sensation (HP:0000223), Somatic sensory dysfunction (HP:0003474), and Hearing abnormality (HP:0000364)
-| Immunodeficiency         | Immunodeficiency (HP:0002721)
-| Cancer                   | Neoplasm (HP:0002664)
-| Reduced fertility        | Decreased fertility (HP:0000144)
+For example, all descendants of the terms 'Intellectual disability' (HP:0001249) or 'Mental deterioration' (HP:0001268) should be annotated as always or often causing intellectual disability (@tbl-true-positives).
 
 ### Quantifying phenotypic severity
-The GPT-4 generated annotation occurrences were converted into a semi-quantitative scoring system, with ‘always’ corresponding to 3, ‘often’ to 2, ‘rarely’ to 1, and ‘never’ to 0.
-These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided annotations (Table 2).
+
+The GPT-4 generated annotation occurrences were converted into a semi-quantitative scoring system, with 'always' corresponding to 3, 'often' to 2, 'rarely' to 1, and 'never' to 0.
+These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided annotations (@tbl-metric-weights).
 Subsequently, the weighted scores underwent normalisation to yield a final quantitative severity score ranging from 0-100, with 100 signifying the maximum severity score attainable.
 
-| Clinical characteristic     | Always (4) | Often (3) | Rarely (2) | Never (0) |
-| Intellectual_disability (5) |     20     |     15    |    10      |     0     | 
-| Impaired_mobility (5)       |     20     |     15    |    10      |     0     | 
-| Blindness (4)               |     16     |     12    |    8       |     0     | 
-| Congenital onset (4)        |     16     |     12    |    8       |     0     | 
-| Physical_malformations (3)  |     12     |     9     |    6       |     0     | 
-| Sensory_impairments (3)     |     12     |     9     |    6       |     0     | 
-| Immunodeficiency (3)        |     12     |     9     |    6       |     0     | 
-| Cancer (3)                  |     12     |     9     |    6       |     0     | 
-| Reduced_fertility (1)       |     4      |     3     |    2       |     0     |
+Let us denote:
 
-## Supplementary figures
+-   $p$ : a phenotype in the HPO.
 
-```{r supp_figure_1}
-dat2 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
-dat2 <- add_ancestor(dat2,
-                     keep_descendants = keep_descendants)
-dat2[,mean_severity_score_gpt:=mean(severity_score_gpt, na.rm=TRUE),
-     by="ancestor_name"] |>
-  data.table::setorderv("mean_severity_score_gpt", -1, na.last = TRUE)
-dat2[,ancestor_name:=factor(ancestor_name,
-                            levels = unique(dat2$ancestor_name),
-                            ordered = TRUE)]
+-   $j$ : the identity of a given annotation metric (i.e. clinical characteristic, such as 'intellectual disability' or 'congenital onset').
 
-ggplot2::ggplot(dat2, ggplot2::aes(x=severity_score_gpt
-                                   # fill=factor(congenital_onset)
-)) +
-  ggplot2::geom_histogram(bins = 50, fill="slateblue") +
-  ggplot2::geom_vline(ggplot2::aes(xintercept=mean_severity_score_gpt),
-                      color="red") +
-  ggplot2::geom_label(data = unique(
-    dat2[,list(mean_severity_score_gpt), by="ancestor_name"]
-  ),
-  ggplot2::aes(x=mean_severity_score_gpt,
-               y=Inf,
-               label=round(mean_severity_score_gpt,2)),
-  color="red", size=3, hjust = 0, vjust = 1.5, alpha=.65) +
-  ggplot2::facet_wrap(facets = "ancestor_name~.", scales = "free_y",
-                      ncol = 3) +
-  ggplot2::labs(y="Frequency", x="Severity score") +
-  ggplot2::theme_bw() +
-  ggplot2::theme(strip.background = ggplot2::element_rect(
-    fill = "transparent"), 
-    text = element_text(size=16))
+-   $W_j$: the assigned weight of metric $j$.
+
+-   $F_j$: the maximum possible value for metric $j$ (equivalent across all $j$).
+
+-   $F_{pj}$ : the numerically encoded value of annotation metric $j$ for phenotype $p$.
+
+-   $NSS_p$: the final composite severity score for phenotype $p$ after applying normalisation to align values to a 0-100 scale and ensure equivalent meaning regardless of which other phenotypes are being analysed in addition to $p$.
+This allows for direct comparability of severity scores across studies with different sets of phenotypes.
+
+::: {#eq-gpt .content-hidden unless-format="html"}
+![](figures/eq5.png){height="300px"}
+
+Computing normalised severity score from encoded GPT-4 annotations.
+:::
+
+\
+\
+
+::: {.content-visible unless-format="html"}
+```{=tex}
+\begin{equation*}
+  \eqnmarkbox[Brown4]{nss}{NSS_p}
+  =
+  \frac{ 
+    \eqnmarkbox[Goldenrod]{nss2}{\sum_{j=1}^{m}} 
+    (
+      \eqnmarkbox[Goldenrod4]{nss3}{F_{pj}}
+      \times 
+      \eqnmarkbox[IndianRed4]{nss4}{W_j}
+    )
+    }{
+    \eqnmarkbox[Tan]{nss5}{\sum_{j=1}^{m}(\max\{F_j\} \times W_j)} 
+  } \times 100
+\end{equation*}
+\annotate[yshift=1em]{left}{nss}{Normalised Severity Score \\for each phenotype}
+\annotate[yshift=3em]{left}{nss2}{Sum of weighted annotation values \\across all metrics}
+\annotate[yshift=3em]{right}{nss3}{Numerically encoded annotation value \\of metric $j$ for phenotype $p$}
+\annotate[yshift=1em]{right}{nss4}{Weight for metric $j$} 
+\annotate[yshift=-1em]{below,right}{nss5}{Theoretical maximum severity score}
 ```
+:::
+
+\
+
+```{r tbl-metric-weights}
+#| label: tbl-metric-weights
+#| tbl-cap: Weighted scores for each clinical characteristic and GPT-4 response category.
+
+## Get weights from HPOExplorer::gpt_annot_codify function defaults
+args <- formals(HPOExplorer::gpt_annot_codify)
+code_dict <- eval(args$code_dict)|>sort(decreasing = TRUE)
+names(code_dict) <- paste0(stringr::str_to_sentence(names(code_dict)),
+                           " (",code_dict,")")
+weights_dict <- eval(args$weights_dict)|>unlist()|>sort(decreasing = TRUE)
+names(weights_dict) <- paste0(stringr::str_to_sentence(gsub("_"," ",names(weights_dict))),
+                           " (",weights_dict,")")
+## Create the outer product of the two weight vectors
+tbl_metric_weights <- as.data.table(outer(weights_dict,code_dict, "*"),
+                                    keep.rownames = "Clinical characteristic")
+
+knitr::kable(as.data.frame(tbl_metric_weights))
+```
+
+### Severity classes
+
+```{r tiers_dict}
+tiers_dict <- unlist(eval(formals(HPOExplorer::gpt_annot_class)$tiers_dict))
+```
+
+The decision tree algorithm used in @Lazarin2014-rz was adapted here for use with the GPT-4 annotations.
+This algorithm first assigned each annotation metric to a tier, where Tier 1 indicated the most severe annotation metrics and Tier 4 indicated the least severe annotation metrics (`r paste(paste0(shQuote(gsub("_"," ",names(tiers_dict))),"=",tiers_dict), collapse = ", ")`).
+If a phenotype often or always caused more than one Tier 1 annotation, it was assigned a severity class of "Profound".
+If the phenotype often or always caused only one Tier 1 annotation, it was assigned a severity class of "Severe".
+A "Severe" class assignment was also assigned if the phenotype often or always caused three or more Tier 2 and Tier3 annotations.
+If the phenotype often or always caused at least one Tier 2 annotation, it was assigned a severity class of "Moderate".
+All remaining phenotypes were was assigned a severity class of "Mild".
+In cases where the phenotype mapped to more than one class, only the most severe class was used.
+This procedure is implemented within the function `HPOExplorer::gpt_annot_class`.
+
+### Correlations between severity metrics
+
+To assess the correlation structure between each severity metric, as well as between the composite severity score and each metric, we computed Pearson correlation coefficients for all pairwise combinations of these variables using the numerically encoded metric values.
+The correlation matrix was visualised using a heatmap, with the colour intensity representing the strength of the correlation (@fig-metric-corplot).
+
+## Data and code availability statement
+
+All code and data used in this study are available on GitHub at:
+
+<https://github.com/neurogenomics/gpt_hpo_annotations>
+
+The GPT-4 annotations for all HPO phenotypes are made available through the R function `HPOExplorer::gpt_annot_read` or in CSV format at:
+
+<https://github.com/neurogenomics/gpt_hpo_annotations/tree/master/data>
+
+A fully reproducible version of this Quarto manuscript can be found at:
+
+<https://github.com/neurogenomics/gpt_hpo_annotations/blob/master/manuscript.qmd>
+
+## Acknowledgements
+
+We would like to thank members of the Monarch Initiative for their insight and feedback throughout this project. In particular, Peter Robinson.
+
+### Funding
+
+This work was supported by a UK Dementia Research Institute (UK DRI) Future Leaders Fellowship \[MR/T04327X/1\] and the UK DRI which receives its funding from UK DRI Ltd, funded by the UK Medical Research Council, Alzheimer’s Society and Alzheimer's Research UK.
 
@@
 
-:::{#refs}
+::: {#refs}
+:::
 
-:::
+{{< pagebreak >}}
+
+## Supplementary Materials
+
+### Supplementary Figures
+
+```{r fig-consist-vs-ontLvl}
+#| label: fig-consist-vs-ontLvl
+#| fig-cap: Relationship between the consistency of GPT-4 annotations (using the stringent criterion) and the level of each phenotype within the HPO ontology (with the number of phenotypes in parentheses).
+Greater ontology levels (x-axis) indicate more specific phenotypes.
+The subtitle indicates summary statistics for the overall relationship between HPO level and the proportion of phenotypes that were annotated consistently.
+The p-values above each bar indicate whether the distribution of consistent/inconsistent annotations, within a given HPO level, significantly deviate from the expected null distribution.
+#| fig-height: 6
+#| fig-width: 12
+
+consist_vs_ontLvl$labels$caption <- NULL
+consist_vs_ontLvl
+```
+
+```{r fig-severity-histo}
+#| label: fig-severity-histo
+#| fig-cap: Distribution of the composite GPT-4 severity score of the severity scores for all HPO terms.
+#| fig-height: 8
+#| fig-width: 10
+
+gpt_annot_plot_out$gp3 +
+  labs(x="Severity score") 
+```
+
+```{r fig-severity-boxplot}
+#| label: fig-severity-boxplot
+#| fig-cap: Boxplot showing the relationship between composite severity score (y-axis) and the frequency response categories within each annotation type.
+#| fig-height: 5
+#| fig-width: 8
+
+gpt_annot_plot_out$gp2 +
+  scale_fill_brewer(palette = "GnBu", direction = -1) +
+  labs(y="Severity score") 
+```
+
+```{r fig-metric-corplot}
+#| label: fig-metric-corplot
+#| fig-cap: Pearson correlations between each individual severity metric and the composite severity score ('severity_score_gpt').
+#| fig-height: 8
+#| fig-width: 8
+
+diag(cor_metrics$P) <- 0
+corrplot::corrplot(cor_metrics$r,
+                   method = c("circle", "square", "ellipse", "number", "shade", "color", "pie")[2],
+                   tl.srt = 45,
+                   tl.col = "grey20",
+                    # insig='blank',
+                   addCoef.col ='grey20', number.cex = 0.8,
+                   order="hclust",
+                   addCoefasPercent=TRUE,
+                   p.mat = cor_metrics$P)
+```
+
+```{r fig-severity-class}
+#| label: fig-severity-class
+#| fig-cap: Distribution of the composite GPT-4 severity score introduced in this paper (y-axis) by an approximation of the severity class system introduced @Lazarin2014-rz (x-axis).
+While these are different schemes for ranking phenotype severity, there is a strong correspondence between them (see summary statistics in subtitle).
+The sample size (number of phenotypes) is shown in parentheses along the x-axis.
+#| fig-height: 6
+#| fig-width: 8
+
+fig_severity_class
+```
+
+{{< pagebreak >}}
+
+### Supplementary Tables
+
+```{r tbl-fas}
+#| label: tbl-fas
+#| tbl-cap: Severity nnotations generated for GPT-4 annotations for the HPO phenotype 'Fetal akinesia sequence' (HP:000198).
+#| tbl-colwidths: [20,20,60]
+
+annot_cols <- setdiff(grep("justification",names(gpt_annot), 
+                       value = TRUE, invert = TRUE),
+                       c("hpo_name","hpo_id","pheno_count"))
+tbl_fas <- gpt_annot[hpo_name=='Fetal akinesia sequence',]|>
+  data.table::melt.data.table(id.vars = c("hpo_name","hpo_id"), 
+                              measure.vars = list(
+                                Annotation=annot_cols,
+                                Justification=grep("justification",names(gpt_annot), 
+                                                   value = TRUE)
+                                )
+                              )
+tbl_fas[,"Clinical characteristic":=stringr::str_to_sentence(gsub("_"," ",annot_cols))]
+knitr::kable(
+  as.data.frame(tbl_fas[,c("Clinical characteristic","Annotation","Justification")])
+  )
+```

### 2024-05-17 docs/_tex/manuscript.tex
Address issue #12

@@
 \hypersetup{
-  pdftitle={Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology},
+  pdftitle={Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology},
   pdfauthor={Kitty B Murphy; Brian M Schilder; Nathan G Skene},
@@
 \begin{document}
-\title{Harnessing AI to annotate the severity of all phenotypic
-abnormalities within the Human Phenotype Ontology}
+\title{Harnessing generative AI to annotate the severity of all
+phenotypic abnormalities within the Human Phenotype Ontology}
 
@@
 the most severe phenotype was `Anencephaly' (HP:0002323) with a
-composite severity score of 52. Anencephaly is a birth defect where the
+composite severity score of 45. Anencephaly is a birth defect where the
 baby is born without a portion of its brain and skull, often these
@@
 branch (Fig.~\ref{fig-severity-histo}). The HPO branch with the greatest
-mean severity score was `Abnormal cellular phenotype' (mean=19.6),
-followed by `Neoplasm' (mean=17.2), which would include the highly
+mean severity score was `Abnormal cellular phenotype' (mean=17),
+followed by `Neoplasm' (mean=15.2), which would include the highly
 ranked phenotypes seen in Figure~\ref{fig-top-phenos}.
@@
 \caption{\label{fig-top-phenos}Quantifying the severity of HPO phenotype
-annotations highlights the most impactful conditions. Heatmap of the top
-(n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based
-on severity score. Severity scores were calculated by multiplying the
-numeric values assigned to each clinical characteristic according to
+annotations highlights the most impactful conditions. Heatmap of 10
+represetantive phenotypes from each severity class (Profound, Severe,
+Moderate, Mild) stratified by whether the phenotypes are often/always
+congenital (\textbf{a}-\textbf{b}) or rarely/never congenital
+(\textbf{c}-\textbf{d}). Continuous severity scores are shown as bars
+(\textbf{b},\textbf{d}) and were calculated by multiplying the numeric
+values assigned to each clinical characteristic according to
 Table~\ref{tbl-metric-weights}. The average normalised score,
@@
 by aggregating the multiplied values and normalising by the theoretical
-maximum severity score. The x-axis is scaled by the severity of the
-clinical characterisation.}
+maximum severity score. The x-axes show each of the clinical
+characteristics.}
 
@@
 positive correspondence with the continuous severity score
-(\(\hat{\omega_{p}^2}\)=0.87, p\textless2.2e-308). In other words,
+(\(\hat{\omega_{p}^2}\)=0.88, p\textless2.2e-308). In other words,
 severity score increased with severity class level (mild \textless{}
@@
 Overall, impaired mobility seemed to be the strongest driver of the
-composite severity score with a Pearson correlation of 0.6138757,
-followed by intellectual disability (r=0.6) and sensory impairments
-(r=0.54).
+composite severity score with a Pearson correlation of 0.6001824,
+followed by intellectual disability (r=0.59) and death (r=0.56).
 
@@
 excellent example of this is the phenotype `Molar tooth sign on MRI'
-(HP:0002419; severity score=30), which GPT-4 annotated as causing
+(HP:0002419; severity score=25.56), which GPT-4 annotated as causing
 intellectual disability. At first glance, we ourselves assumed this was
@@
 the most severe phenotypes in the HPO is `Fetal akinesia sequence' (FAS;
-HP:0001989, severity score= 50.7), and extremely rare condition that is
+HP:0001989, severity score= 43.9), and extremely rare condition that is
 almost always lethal. FAS is a complex, multi-system phenotype that can
@@
 \endlastfoot
+Death (6) & 18 & 12 & 6 & 0 \\
 Intellectual disability (5) & 15 & 10 & 5 & 0 \\
-Death (5) & 15 & 10 & 5 & 0 \\
 Impaired mobility (4) & 12 & 8 & 4 & 0 \\

### 2024-05-17 docs/manuscript.qmd
Address issue #12

--- 2024-05-14 manuscript.qmd
@@
 ---
-title: Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
+title: Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
 author:
@@
 ```{r gpt_annot_class, results='hide'}
-res_class <- HPOExplorer::gpt_annot_class(annot_coded = res_coded$annot_coded)
-res_class <- merge(res_class,
-                   res_coded$annot_weighted[,c("hpo_name","severity_score_gpt")])
+res_class <- HPOExplorer::gpt_annot_class(res_coded = res_coded)
+
 fig_severity_class <- ggstatsplot::ggbetweenstats(res_class,
@@
   width = 10)
+```
+
+```{r weights_dict}
+weights_dict <- eval(formals(gpt_annot_codify)$weights_dict)
+annotation_order <- gsub("_"," ",names(sort(unlist(weights_dict), decreasing = TRUE)))
 ```
@@
 # subset to clinical characteristics   
-annotation_order <- c("death", "intellectual disability", "impaired mobility", "blindness", 
-                      "congenital onset", "physical malformations", "sensory impairments", 
-                      "immunodeficiency", "cancer", "reduced fertility")
 occurr_df <- gpt_annot[,c("hpo_name",gsub(" ", "_", annotation_order)),with=FALSE]
@@
 
-```{r severity_scoring}
-## code adapted from HPOExplorer::gpt_annot_plot()
-# get all descendants of the term phenotypic abnormality 
-keep_ont_levels=seq(3,17)
-#### Filter out ont levels ####
-annot_melt <- HPOExplorer::add_ancestor(annot_melt,
-                                        keep_descendants = keep_descendants)
-data.table::setorderv(annot_melt,"severity_score_gpt",-1)
-
-annot_melt <- HPOExplorer::add_ont_lvl(annot_melt, keep_ont_levels = keep_ont_levels)
-annot_melt$variable <- gsub("_", " ", annot_melt$variable)
-
-# get all unique phenotypes
-total_unique <- length(unique(annot_melt$hpo_id))
-
-# get the top N phenotypes
-top_n <- 15
-dat_top <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(top_n)]]
-
-# Get the bottom N phenotypes
-bottom_n <- 15
-dat_bottom <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(total_unique - bottom_n + 1, total_unique)]]
-
-# Define the range for selecting phenotypes from the middle
-start_index <- ceiling(total_unique / 2)  
-end_index <- start_index + 12
-
-# Get phenotypes from the middle
-dat_middle <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(start_index, end_index)]]
-
-# define labels
-dat_top$position <- "Top"
-dat_middle$position <- "Middle"
-dat_bottom$position <- "Bottom"
-
-dat_all <- rbind(dat_top, dat_middle, dat_bottom)
-dat_all$position <- factor(dat_all$position, levels=unique(dat_all$position))
-dat_all$variable <- factor(dat_all$variable, levels = annotation_order)
-```
-
 ```{r fig-top-phenos}
@@
 #| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions.
-Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score.
-Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights.
+Heatmap of 10 represetantive phenotypes from each severity class (Profound, Severe, Moderate, Mild) stratified by whether the phenotypes are often/always congenital (**a**-**b**) or rarely/never congenital (**c**-**d**).
+Continuous severity scores are shown as bars (**b**,**d**) and were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights.
 The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score.
-The x-axis is scaled by the severity of the clinical characterisation.
-#| fig-height: 7.5
-#| fig-width: 8
+The x-axes show each of the clinical characteristics.
+#| fig-height: 10
+#| fig-width: 13
 
-p1 <- ggplot2::ggplot(data = dat_all,
-                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
-  ggplot2::geom_tile(colour = "white", lwd = 0.5, linetype =1) +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_brewer(palette = "GnBu", direction = -1) +
-  ggplot2::theme_classic() +
-  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right') +
-  labs(x = "Clinical characteristic",
-       y = "HPO phenotype",
-       fill = "Occurrence") +
-  ggplot2::facet_grid(position ~ ., scales = "free_y")
-
-p2 <- ggplot2::ggplot(data = dat_all,
-                      ggplot2::aes(x="severity_score_gpt", y=hpo_name,
-                                   fill=severity_score_gpt)) +
-  ggplot2::geom_tile() +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = "reverse") +
-  ggplot2::labs(x=NULL, y=NULL, fill = "Severity\nscore") +
-  ggplot2::theme_void() +
-  guides(fill = guide_colorbar(reverse = TRUE)) +
-  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
-                 axis.ticks.y = ggplot2::element_blank(),
-                 legend.position = 'right')
-
-p3 <- patchwork::wrap_plots(p1, p2, ncol = 2,
-                            widths = c(1,.2),
-                            guides = "collect")
-
-methods::show(p3)
+plot_top_phenos_out <- HPOExplorer::plot_top_phenos(res_class = res_class)
+plot_top_phenos_out$plot
 ```
@@
 Throughout this study, we observed that GPT-4 was capable of reliably recovering deep semantic relationships from the medical domain, far beyond making superficial inferences based on lexical similarities.
-An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt`), which GPT-4 annotated as causing intellectual disability.
+An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r round(res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt,2)`), which GPT-4 annotated as causing intellectual disability.
 At first glance, we ourselves assumed this was a false positive as the term appeared to be related to dentition.

### 2024-05-22 docs/_tex/manuscript.tex
Add fig-congenital-branches

@@
 
+\subsubsection{Congenital onset by HPO
+branch}\label{congenital-onset-by-hpo-branch}
+
+\phantomsection\label{cell-fig-congenital-branches}
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{manuscript_files/figure-pdf/fig-congenital-branches-1.pdf}
+
+}
+
+\caption{\label{fig-congenital-branches}Distribution of congenital onset
+across HPO branches. The y-axis shows the proportion of phenotypes that
+are always/often/rarely/never congenital. The x-axis shows the HPO
+branch, orderered from highest to lowest proportion of always congenital
+phenotypes.}
+
+\end{figure}%
+
+Next, we assessed the distribution of congenital onset across HPO
+branches (Fig.~\ref{fig-congenital-branches}). We found that the
+Abnormality of prenatal development or birth branch contained the
+greatest proportion of phenotypes that were always congenital (70.15\%),
+followed by Abnormality of the musculoskeletal system (45.34\%) and
+Growth abnormality (37.62\%). This is concordant with the expectation
+that these phenotypes should largely be congenital. The HPO branches
+with the least commonly congenital phenotypes were Constitutional
+symptom (0\%), Abnormality of the thoracic cavity (0\%), and Phenotypic
+abnormality (0\%).
+\href{https://hpo.jax.org/app/browse/term/HP:0025142}{`Constitutional
+symptom'} is a fairly broad term defined as \emph{`A symptom or
+manifestation indicating a systemic or general effect of a disease and
+that may affect the general well-being or status of an individual.'}
+Examples include `Fatigue' `Exercise intolerance', `Hot flashes' and
+`Sneeze'.
+
 \subsection{Discussion}\label{discussion}

### 2024-05-22 docs/manuscript.qmd
Add fig-congenital-branches

@@
 
+### Congenital onset by HPO branch
+
+```{r gpt_annot_plot_branches}
+gpt_annot_plot_branches_out <- HPOExplorer::gpt_annot_plot_branches(
+  gpt_annot = gpt_annot, 
+  hpo = hpo, 
+  metric = "congenital_onset", 
+  fill_lab = "Congenital\nonset", 
+  show_plot = FALSE)
+top_congenital <- unique(gpt_annot_plot_branches_out$data[!is.na(congenital_onset),c("ancestor_name","proportion_always")])
+```
+
+```{r fig-congenital-branches}
+#| label: fig-congenital-branches
+#| fig-cap: Distribution of congenital onset across HPO branches.
+The y-axis shows the proportion of phenotypes that are always/often/rarely/never congenital.
+The x-axis shows the HPO branch, orderered from highest to lowest proportion of always congenital phenotypes.
+#| fig-height: 6
+#| fig-width: 6
+
+gpt_annot_plot_branches_out$plot
+```
+
+Next, we assessed the distribution of congenital onset across HPO branches ([Fig.
+@fig-congenital-branches]).
+We found that the `r top_congenital$ancestor_name[1]` branch contained the greatest proportion of phenotypes that were always congenital (`r round(top_congenital$proportion_always[1]*100,2)`%), followed by `r top_congenital$ancestor_name[2]` (`r round(top_congenital$proportion_always[2]*100,2)`%) and `r top_congenital$ancestor_name[3]` (`r round(top_congenital$proportion_always[3]*100,2)`%).
+This is concordant with the expectation that these phenotypes should largely be congenital.
+The HPO branches with the least commonly congenital phenotypes were `r rev(top_congenital$ancestor_name)[1]` (`r round(rev(top_congenital$proportion_always)[1]*100,2)`%), `r rev(top_congenital$ancestor_name)[2]` (`r round(rev(top_congenital$proportion_always)[2]*100,2)`%), and `r rev(top_congenital$ancestor_name)[3]` (`r round(rev(top_congenital$proportion_always)[3]*100,2)`%).
+['Constitutional symptom'](https://hpo.jax.org/app/browse/term/HP:0025142) is a fairly broad term defined as *'A symptom or manifestation indicating a systemic or general effect of a disease and that may affect the general well-being or status of an individual.'* Examples include 'Fatigue' 'Exercise intolerance', 'Hot flashes' and 'Sneeze'.
+
 ## Discussion

### 2024-05-23 docs/_tex/manuscript.tex
address NGS hypothesis comments

@@
 
-The Human Phenotype Ontology (HPO) has played a crucial role in
-defining, diagnosing, prognosing, and treating human diseases by
-providing a standardised framework for describing phenotypic
-abnormalities. However, documentation on the severity of each phenotype
-is currently very sparse, making systematic analyses and prioritisation
-of results challenging. With \textgreater17500 phenotypic abnormalities
-across \textgreater8600 rare diseases, manual curation of such
-phenotypic annotations by experts would be exceedingly labour-intensive
-and time-consuming. Leveraging advances in artificial intelligence, we
-employed the OpenAI GPT-4 large language model (LLM) to systematically
-annotate the severity of all phenotypic abnormalities in the HPO. First,
-we benchmarked the generative LLM annotations against ground-truth
-labels within the HPO (e.g.~phenotypes in the `Cancer' HPO branch were
-annotating as causing cancer by GPT-4). True positive recall rates
-across different annotations ranged from 89-100\% (mean=96\%), clearly
-demonstrating the ability of GPT-4 to automate the curation process with
-a high degree of fidelity. Using a novel approach, we developed a
-severity scoring system that incorporates both the nature of the
-phenotype outcome and the frequency of its occurrence. These severity
-metrics will enable efforts to systematically prioritise which human
-phenotypes are most detrimental to human health, and best targets for
-therapeutic intervention.
+There are thousands of human phenotypes which are linked to genetic
+variation. These range from the benign (white eyelashes) to the deadly
+(respiratory failure). The Human Phenotype Ontology has categorised all
+human phenotypic variation into an unified framework that defines the
+relationships between them (e.g.~missing arms and missing legs are both
+abnormalities of the limb). This has made it possible to perform
+phenome-wide analyses, e.g.~to prioritise which make the best candidates
+for gene therapies. However, there is currently limited metadata
+describing the effects / severity of these phenotypes. With
+\textgreater17500 phenotypic abnormalities across \textgreater8600 rare
+diseases, manual curation of such phenotypic annotations by experts
+would be exceedingly labour-intensive and time-consuming. Leveraging
+advances in artificial intelligence, we employed the OpenAI GPT-4 large
+language model (LLM) to systematically annotate the severity of all
+phenotypic abnormalities in the HPO. First, we benchmarked the
+generative LLM annotations against ground-truth labels within the HPO
+(e.g.~phenotypes in the `Cancer' HPO branch were annotating as causing
+cancer by GPT-4). True positive recall rates across different
+annotations ranged from 89-100\% (mean=96\%), clearly demonstrating the
+ability of GPT-4 to automate the curation process with a high degree of
+fidelity. Using a novel approach, we developed a severity scoring system
+that incorporates both the nature of the phenotype outcome and the
+frequency of its occurrence. These severity metrics will enable efforts
+to systematically prioritise which human phenotypes are most detrimental
+to human health, and best targets for therapeutic intervention.
 
@@
 
-Comprehensive annotation of phenotypic abnormalities is invaluable for
+Ontologies provide a common language with which to communicate concepts.
+In medicine, ontologies for phenotypic abnormalities are invaluable for
 defining, diagnosing, prognosing, and treating human disease. Since
-2008, the Human Phenotype Ontology (HPO) has been instrumental to this,
-by providing a standardised database for the description and analysis of
-human phenotypes (Gargano et al., 2024; Köhler et al., 2021). Through
-developing open community resources, the depth and breadth of the HPO
-has continued to expand and there are now \textgreater17500 phenotypic
-abnormalities across \textgreater8600 diseases, described. In recent
-years, the HPO has expanded its disease annotations so that each HPO
-term can have metadata including typical age of onset and frequency. In
-addition, there are the \emph{Clinical modifier} and \emph{Clinical
-course} subontologies, which contains terms to describe factors
-including severity, triggers, mortality and progression, respectively.
+2008, the Human Phenotype Ontology (HPO) has been instrumental in
+healthcare and biomedical research by providing a framework for
+comprehensively describing human phenotypes and the relationships
+between them (Gargano et al., 2024; Köhler et al., 2021). By expanding
+its depth and breadth over time, the HPO now contains \textgreater17500
+phenotypic abnormalities across \textgreater8600 diseases. Some HPO
+phenotypes also contain metadata annotations such typical age of onset,
+frequency, triggers, time course, mortality rate and typical severity.
 Describing the severity-related attributes of a disease is crucial for
-both research and clinical care of individuals with rare diseases. This
-includes enhancing diagnostic capabilities, as well as prioritising and
-guiding gene therapy trials.
-
-To date, the HPO has largely relied on manual curation by experts
-including clinicians, clinical geneticists and researchers. Although
-this approach ensures the quality and accuracy of the ontology, it is
-time-consuming and labour-intensive. As a result, less than 1\% of terms
-within the HPO have metadata pertaining to their features such as time
-course and severity. Artificial intelligence (AI) capabilities have
-advanced considerably in recent years, presenting new opportunities to
-integrate natural language processing technologies into assisting in the
-curation process. Specifically, there have recently been considerable
-advances in large language model (LLM) and their application to
-biomedical problems, in some cases performing as well or better than
-human clinicians on standardised medical exams and patient diagnosis
-tasks (Bolton et al., 2024; Cheng et al., 2023; Gu et al., 2021; Labrak
-et al., 2024; Luo et al., 2022; McDuff et al., 2023; O'Neil et al.,
-2024; Shin et al., 2020; Singhal, Azizi, et al., 2023, 2023; Singhal,
-Tu, et al., 2023; Van Veen et al., 2024; Zhang et al., 2023). Recent
-work has demonstrated that the Generative Pre-trained Transformer 4
-(GPT-4) foundation model (OpenAI et al., 2024), when combined with
-strategic prompt engineering, can outperform even specialist LLMs that
-are explicitly fine-tuned for biomedical tasks (Nori et al., 2023). In a
-landmark achievement, GPT-4 was the first LLM to surpass a score of 90\%
-in the United States Medical Licensing Examination (USML) (Nori et al.,
-2023).
+both research and clinical care of individuals with rare diseases. When
+researchers or clinicians are presented with phenotypes that fall
+outside of their expertise, resources to quickly and reliably retrieve
+summaries with additional relevant information about these phenotypes
+are essential. In the clinic, this can help in reaching a differential
+diagnosis or prioritising the treatment of some phenotypes over others.
+In research, this information is useful for prioritising targets for
+causal disease mechanisms, performing large-scale analyses of phenotypic
+data, and guiding funding agencies when assessing the potential impact
+and need for research in a given disease area. To date, the HPO has
+largely relied on manual curation by domain experts. While this approach
+can improve annotation quality and accuracy, it is both time-consuming
+and labour-intensive. As a result, less than 1\% of terms within the HPO
+contain metadata such as time course and severity.
+
+Artificial intelligence (AI) capabilities have advanced considerably in
+recent years, presenting new opportunities to integrate natural language
+processing technologies into assisting in the curation process.
+Specifically, there have recently been considerable advances in large
+language model (LLM) and their application to biomedical problems, in
+some cases performing as well or better than human clinicians on
+standardised medical exams and patient diagnosis tasks (Bolton et al.,
+2024; Cheng et al., 2023; Gu et al., 2021; Labrak et al., 2024; Luo et
+al., 2022; McDuff et al., 2023; O'Neil et al., 2024; Shin et al., 2020;
+Singhal, Azizi, et al., 2023, 2023; Singhal, Tu, et al., 2023; Van Veen
+et al., 2024; Zhang et al., 2023). Recent work has demonstrated that the
+Generative Pre-trained Transformer 4 (GPT-4) foundation model (OpenAI et
+al., 2024), when combined with strategic prompt engineering, can
+outperform even specialist LLMs that are explicitly fine-tuned for
+biomedical tasks (Nori et al., 2023). In a landmark achievement, GPT-4
+was the first LLM to surpass a score of 90\% in the United States
+Medical Licensing Examination (USML) (Nori et al., 2023).
 
@@
 
-793 phenotypes were annotated multiple times to evaluate the consistency
-of responses generated by GPT-4. For a subset of phenotypes with known
-expected annotations, true positive rates were calculated to assess
-recall. Additionally, based on the clinical characteristics and their
-occurrence, we have quantified the severity of each phenotype, providing
-an example of how these clinical characteristic annotations can be used
-to guide prioritisation of gene therapy trials. Ultimately, we hope that
-our resource will be of utility to those working in rare diseases, as
-well as the wider healthcare community.
+To evaluate the consistency of responses generated by GPT-4 793
+phenotypes were annotated multiple times. For a subset of phenotypes
+with known expected annotations, true positive rates were calculated to
+assess recall. Additionally, based on the clinical characteristics and
+their occurrence, we have quantified the severity of each phenotype,
+providing an example of how these clinical characteristic annotations
+can be used to guide prioritisation of gene therapy trials. Ultimately,
+we hope that our resource will be of utility to those working in rare
+diseases, as well as the wider healthcare community.
 
@@
 associated with each effect and required a justification for each
-response. For each effect, we asked about the likelihood of its
+response. For each effect, we asked about the frequency of its
 occurrence - whether it never, rarely, often, or always occurred.
+Framing the queries in this way served two purposes. First, this helped
+to constrain the responses of GPT-4 to a specific range of values,
+making answers more consistent and amenable to downstream data analysis.
+Second, it served to overcome one of the main limitations noted by
+Lazarin et al. (2014) as they did not collect information on how the
+frequency of each disease affected their decision making when generating
+severity annotations.
 
@@
 
-Less than 1\% of phenotype always directly resulted in death, such as
-`Stillbirth', `Anencephaly' and `Bilateral lung agenesis'. Only 71
-phenotypes were annotated as always causing death, while 9707 phenotypes
-were annotated as often or rarely causing death. 7880 phenotypes were
-annotated as never causing death. This included 34 unique forms of
+Less than 1\% of phenotypes always directly resulted in death (n=71),
+such as `Stillbirth', `Anencephaly' and `Bilateral lung agenesis'.
+Meanwhile, 9707 phenotypes were annotated as often or rarely causing
+death. 7880 phenotypes were annotated as never causing death. Examples
+of phenotypes that never cause death included 34 unique forms of
 syndactyly, a non-lethal condition that causes fused or webbed fingers
@@
 
-To assess annotation consistency, 793 we queried GPT-4 with a subset of
-the HPO phenotypes multiple times. We employed two different metrics to
-determine the \emph{consistency rate}. The first, less stringent metric,
-defined consistency as the duplicate annotations being either `always'
-and `often', or `never' and `rarely'. The second, more stringent metric,
-required exact agreement in annotation occurrences, e.g.~`always' and
-`always'. For the less stringent metric, duplicated phenotypes were
-annotated consistently at a rate of at least 80\%, and for the more
-stringent metric, the lowest consistency rate was 57\% for congenital
-onset. An example of where annotations were inconsistent was for the HPO
-term `Acute leukaemia'. One time, GPT-4 annotated it as often causing
-impaired mobility, giving the justification that `weakness and fatigue
-from leukaemia and its treatment can impair mobility'. The other time,
-GPT-4 annotated it as rarely causing impaired mobility, giving the
-justification that `acute leukaemia rarely impairs mobility directly'.
-Despite specifying in the prompt for GPT-4 not to take into
-consideration indirect effects, this is an example of where it failed to
-do so.
+To assess annotation consistency, we queried GPT-4 with a subset of the
+HPO phenotypes multiple times (n=793 unique phenotypes). We employed two
+different metrics to determine the \emph{consistency rate}. The first,
+less stringent metric, defined consistency as the duplicate annotations
+being either `always' and `often', or `never' and `rarely'. The second,
+more stringent metric, required exact agreement in annotation
+occurrences, e.g.~`always' and `always'. For the less stringent metric,
+duplicated phenotypes were annotated consistently at a rate of at least
+80\%, and for the more stringent metric, the lowest consistency rate was
+57\% for congenital onset. An example of where annotations were
+inconsistent was for the HPO term `Acute leukaemia'. One time, GPT-4
+annotated it as often causing impaired mobility, giving the
+justification that `weakness and fatigue from leukaemia and its
+treatment can impair mobility'. The other time, GPT-4 annotated it as
+rarely causing impaired mobility, giving the justification that `acute
+leukaemia rarely impairs mobility directly'. Despite specifying in the
+prompt for GPT-4 not to take into consideration indirect effects, this
+is an example of where it failed to do so.
 
@@
 were annotated correctly by GPT-4. For example, `Molar tooth sign on
-MRI' was correctly annotated as causing intellectual disability.
+MRI' (a neurodevelopmental pathology observed in radiological scans) was
+correctly annotated as causing intellectual disability.
 
@@
 
-Quantifying the severity of phenotypes can have important implications
-for diagnosis, prognosis, and treatment, for instance, guiding
-prioritisation of gene therapy trials for the most severe phenotypes.
-First, we created a dictionary to map each phenotype outcome
-(e.g.~blindness) and its response (always, often, rarely, never) to
-numeric values from 0-3. Then, the phenotype outcome values were
-multiplied by their associated response values. Importantly, the values
-reflected the severity of each outcome based on both the outcome itself
-and the frequency of the response. For instance, a phenotype always
-causing death would have a higher multiplied value than a phenotype
-rarely causing reduced fertility (see Table~\ref{tbl-metric-weights}).
-Next, we computed an average score for each phenotype by aggregating the
+While individual annotations are informative, we wanted to be able to
+distil the severity of each phenotype into a single score. Quantifying
+the overall severity of phenotypes can have important implications for
+diagnosis, prognosis, and treatment. It may also guide the
+prioritisation of gene therapy trials for phenotypes with the most
+severe outcomes and thus the most urgent need. Importantly, the values
+reflected the severity of each outcome based on both the type of outcome
+itself and its frequency within a particular phenotype. For instance, a
+phenotype always causing death would have a higher multiplied value than
+a phenotype often causing reduced fertility (see
+Table~\ref{tbl-metric-weights}). First, we created a dictionary to map
+each phenotype outcome (e.g.~blindness) and its frequency (always,
+often, rarely, never) to numeric values from 0-3. Then, the phenotype
+outcome values were multiplied by phenotype outcome weights. Next, we
+computed an average score for each phenotype by aggregating the
 multiplied values across all phenotype outcomes and then calculating the
@@
 maximum severity score. The x-axes show each of the clinical
-characteristics.}
+characteristics. All data for this figure, as well as justifications for
+each annotation, can be found in Table~\ref{tbl-annotations}.}
 
@@
 
+Phenotype severity annotations have utility across a wide variety of
+applications in both the clinic and research. In clinical settings,
+severity annotations can be used to prioritise the treatment of some
+phenotypes over others in patients with complex presentations, avoid
+administering contraindicated drugs, and prognosing potential health
+outcomes. In research settings, severity annotations can be used to
+identify phenotypes that have a large impact on patient outcomes and yet
+are currently understudied. They may also be used to help design new
+experiments and studies, or even provide new insights into the
+underlying aetiology of the disease by making expert-level summaries
+more immediately accessible to the wider research community.
+
 The creation and annotation of biomedical knowledge has traditionally
@@
 phenotypes tested. Furthermore, our approach can be readily adapted and
-scaled to accommodate the growing volume of phenotypic data.
+scaled to accommodate the growing volume of phenotypic data. In total,
+the entire study cost \$296.27 in queries to the OpenAI API. While we do
+not have a direct comparison, this likely represents a extremely small
+fraction of the total costs of such a study if performed manually by
+human experts charging at an hourly rate. Even if all human annotations
+were provided on a volunteer basis, this would still require hundreds if
+not thousands of hours of cumulative manual human labour. Using our
+approach, severity annotations for the entire HPO can be generated
+automatically in a matter of several hours.
 
@@
 prioritising gene therapy candidates (Murphy et al., 2023) and guiding
-clinical decision-making in rare diseases. As the HPO continues to grow
-over time, this approach can easily be repeated to keep pace with the
-rapidly evolving biomedical landscape. Furthermore, this approach can be
-extended to produce different sets of annotations or be used with any
-other ontology. Additional use cases include gathering data on the
-prevalence of each phenotype to approximate their social and financial
-costs.
+clinical decision-making in rare diseases. It may also be used as tool
+to help inform policy decisions and funding allocation by healthcare or
+governmental institutions. This of course would need to be in
+consultation with subject matter medical experts, patients, advocates
+and biomedical ethicists before reaching a final decision. Nevertheless,
+access to succinct, interpretable, and semi-quantitative severity
+annotations may encourage key decision makers with limited time to
+review individual proposals to pay heed to phenotypes and diseases that
+would otherwise be overlooked. As the HPO and the broader literature
+continue to grow over time, our automated AI-based approach can easily
+be repeated to keep pace with the rapidly evolving biomedical landscape.
+Furthermore, it can be extended to produce different sets of annotations
+or be used with any other ontology. Additional use cases include
+gathering data on the prevalence of each phenotype to approximate their
+social and financial costs.
 
@@
 phenotype would cause death with no or suboptimal treatment.
+
+Another limitation with the present dataset is that phenotypes
+themselves can manifest with different degrees of severity, in the sense
+that they are more pronounced or intense. For example, sensitivity to
+light could range from a mild inconvenience to a severe disability that
+prevents the individual from leaving their home during the day. The
+effect of onset (beyond congenital vs.~non-congenital) and time course
+(acute, slowly progression, relapse-remitting) were also not explicitly
+considered. Finally, we did not ask GPT-4 to consider phenotypes as they
+present within particular diseases. For example, while the phenotype
+`Hypertension' may be mild to moderate in the general population and not
+present until middle-age, it can also present early in life as very
+severe in the context of a rare genetic disorder such as Liddle
+syndrome. Future work could explore these nuances in more detail.
 
@@
 characteristic - whether it never, rarely, often, or always occurred.
+This design helps to constrain the potential responses of GPT-4 and thus
+make it more amenable to machine-readable post-processing. It also
+serves to address one of its key limitations from the Lazarin et al.
+(2014) survey, namely the lack information on how clinical
+characteristic frequency affected the clinicians' severity annotations.
+Here, we can instead use the frequency values to generate more precise
+annotations and downstream severity ranking scores.
+
 Furthermore, our prompt design revealed that the optimal trade-off
@@
 prompt. An example prompt can be seen in Figure~\ref{fig-occurrence}.
-Only 2 phenotypes were included per prompt in order to 1) avoid
+Thus, only two phenotypes were included per prompt in order to 1) avoid
 exceeding per-query token limits, and 2) prevent the breakdown of GPT-4
@@
 \subsubsection{Supplementary Tables}\label{supplementary-tables}
+
+\begin{table}
+
+\caption{\label{tbl-annotations}Table of GPT-4 annotations for all Human
+Phenotype Ontology (HPO) phenotypes in Figure~\ref{fig-top-phenos}. For
+each phenotype, this includes the name of the phenotype (`hpo\_name'),
+the ID of the phenotype (`hpo\_id'), the frequency of each annotation
+(always, often, rarely, never), and the justification for each
+annotation ('\ldots\_justification'). These results can also be
+downloaded programmatically using the R function
+\texttt{HPOExplorer::gpt\_annot\_check}.}
+
+\centering{
+
+\href{https://github.com/neurogenomics/gpt_hpo_annotations/raw/master/data/top_phenos_annotations.csv.gz}{\textbf{Top
+phenotype annotations table}}
+
+}
+
+\end{table}%
 

### 2024-05-23 docs/manuscript.qmd
address NGS hypothesis comments

@@
 res_class <- HPOExplorer::gpt_annot_class(res_coded = res_coded)
-
 fig_severity_class <- ggstatsplot::ggbetweenstats(res_class,
@@
 ## Abstract
-
-The Human Phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised framework for describing phenotypic abnormalities.
-However, documentation on the severity of each phenotype is currently very sparse, making systematic analyses and prioritisation of results challenging.
+There are thousands of human phenotypes which are linked to genetic variation.
+These range from the benign (white eyelashes) to the deadly (respiratory failure).
+The Human Phenotype Ontology has categorised all human phenotypic variation into an unified framework that defines the relationships between them (e.g.
+missing arms and missing legs are both abnormalities of the limb).
+This has made it possible to perform phenome-wide analyses, e.g.
+to prioritise which make the best candidates for gene therapies.
+However, there is currently limited metadata describing the effects / severity of these phenotypes.
 With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming.
@@
 
-Comprehensive annotation of phenotypic abnormalities is invaluable for defining, diagnosing, prognosing, and treating human disease.
-Since 2008, the Human Phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
-Through developing open community resources, the depth and breadth of the HPO has continued to expand and there are now \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases, described.
-In recent years, the HPO has expanded its disease annotations so that each HPO term can have metadata including typical age of onset and frequency.
-In addition, there are the *Clinical modifier* and *Clinical course* subontologies, which contains terms to describe factors including severity, triggers, mortality and progression, respectively.
+Ontologies provide a common language with which to communicate concepts.
+In medicine, ontologies for phenotypic abnormalities are invaluable for defining, diagnosing, prognosing, and treating human disease.
+Since 2008, the Human Phenotype Ontology (HPO) has been instrumental in healthcare and biomedical research by providing a framework for comprehensively describing human phenotypes and the relationships between them [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
+By expanding its depth and breadth over time, the HPO now contains \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases.
+Some HPO phenotypes also contain metadata annotations such typical age of onset, frequency, triggers, time course, mortality rate and typical severity.
 Describing the severity-related attributes of a disease is crucial for both research and clinical care of individuals with rare diseases.
-This includes enhancing diagnostic capabilities, as well as prioritising and guiding gene therapy trials.
+When researchers or clinicians are presented with phenotypes that fall outside of their expertise, resources to quickly and reliably retrieve summaries with additional relevant information about these phenotypes are essential.
+In the clinic, this can help in reaching a differential diagnosis or prioritising the treatment of some phenotypes over others.
+In research, this information is useful for prioritising targets for causal disease mechanisms, performing large-scale analyses of phenotypic data, and guiding funding agencies when assessing the potential impact and need for research in a given disease area.
+To date, the HPO has largely relied on manual curation by domain experts.
+While this approach can improve annotation quality and accuracy, it is both time-consuming and labour-intensive.
+As a result, less than 1% of terms within the HPO contain metadata such as time course and severity.
 
-To date, the HPO has largely relied on manual curation by experts including clinicians, clinical geneticists and researchers.
-Although this approach ensures the quality and accuracy of the ontology, it is time-consuming and labour-intensive.
-As a result, less than 1% of terms within the HPO have metadata pertaining to their features such as time course and severity.
 Artificial intelligence (AI) capabilities have advanced considerably in recent years, presenting new opportunities to integrate natural language processing technologies into assisting in the curation process.
@@
 
-`r checks$consistency_count[[1]]` phenotypes were annotated multiple times to evaluate the consistency of responses generated by GPT-4.
+To evaluate the consistency of responses generated by GPT-4 `r checks$consistency_count[[1]]` phenotypes were annotated multiple times.
 For a subset of phenotypes with known expected annotations, true positive rates were calculated to assess recall.
@@
   labs(title="Example GPT-4 prompt") +
-  ggtext::geom_textbox(aes(x=.5,y=.7),
+  ggtext::geom_textbox(aes(x=.5,y=.6),
                        box.padding = margin(20, 10, 20, 10),
@@
                        width = unit(.8, "npc"),
-                       size=3,
+                       size=3.5,
                        halign=.5,
@@
 Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.
+For each effect, we asked about the frequency of its occurrence - whether it never, rarely, often, or always occurred.
+Framing the queries in this way served two purposes.
+First, this helped to constrain the responses of GPT-4 to a specific range of values, making answers more consistent and amenable to downstream data analysis.
+Second, it served to overcome one of the main limitations noted by @Lazarin2014-rz as they did not collect information on how the frequency of each disease affected their decision making when generating severity annotations.
 
@@
 
-Less than `r ceiling(subset(count_df_long, condition=="always" & category=="death")$percentage)`% of phenotype always directly resulted in death, such as 'Stillbirth', 'Anencephaly' and 'Bilateral lung agenesis'.
-Only `r uniqueN(gpt_annot[death=="always"]$hpo_name)` phenotypes were annotated as always causing death, while `r uniqueN(gpt_annot[death %in% c("often","rarely")]$hpo_name)` phenotypes were annotated as often or rarely causing death.
+Less than `r ceiling(subset(count_df_long, condition=="always" & category=="death")$percentage)`% of phenotypes always directly resulted in death (n=`r uniqueN(gpt_annot[death=="always"]$hpo_name)`), such as 'Stillbirth', 'Anencephaly' and 'Bilateral lung agenesis'.
+Meanwhile, `r uniqueN(gpt_annot[death %in% c("often","rarely")]$hpo_name)` phenotypes were annotated as often or rarely causing death.
 `r uniqueN(gpt_annot[death %in% c("never")]$hpo_name)` phenotypes were annotated as never causing death.
-This included `r sum(grepl("syndactyly",gpt_annot[death %in% c("never")]$hpo_name, ignore.case = TRUE))` unique forms of syndactyly, a non-lethal condition that causes fused or webbed fingers (occurring 1 in 1,200–15,000 live births).
+Examples of phenotypes that never cause death included `r sum(grepl("syndactyly",gpt_annot[death %in% c("never")]$hpo_name, ignore.case = TRUE))` unique forms of syndactyly, a non-lethal condition that causes fused or webbed fingers (occurring 1 in 1,200–15,000 live births).
 While not life-threatening itself, syndactyly is a symptom of genetic disorders that can cause life-threatening cardiovascular and neurodevelopmental defects, such as Apert Syndrome [@garagnaniSyndromesAssociatedSyndactyly2013].
@@
                           y = "ontLvl") +
-  ggplot2::labs(title = "Consistency of GPT-4 annotations by HPO level")
+  ggplot2::labs(x="Human Phenotype Ontology level", title = "Consistency of GPT-4 annotations by HPO level")
 consist_vs_ontLvl_dt <- MSTExplorer:::get_ggstatsplot_stats(consist_vs_ontLvl)
@@
 
-To assess annotation consistency, `r checks$consistency_count[[1]]` we queried GPT-4 with a subset of the HPO phenotypes multiple times.
+To assess annotation consistency, we queried GPT-4 with a subset of the HPO phenotypes multiple times (n=`r checks$consistency_count[[1]]` unique phenotypes).
 We employed two different metrics to determine the *consistency rate*.
@@
 In addition, were numerous phenotypes with lexically non-obvious relationships to the clinical characteristic that were annotated correctly by GPT-4.
-For example, 'Molar tooth sign on MRI' was correctly annotated as causing intellectual disability.
+For example, 'Molar tooth sign on MRI' (a neurodevelopmental pathology observed in radiological scans) was correctly annotated as causing intellectual disability.
 
@@
 
-Quantifying the severity of phenotypes can have important implications for diagnosis, prognosis, and treatment, for instance, guiding prioritisation of gene therapy trials for the most severe phenotypes.
+While individual annotations are informative, we wanted to be able to distil the severity of each phenotype into a single score.
+Quantifying the overall severity of phenotypes can have important implications for diagnosis, prognosis, and treatment.
+It may also guide the prioritisation of gene therapy trials for phenotypes with the most severe outcomes and thus the most urgent need.
+Importantly, the values reflected the severity of each outcome based on both the type of outcome itself and its frequency within a particular phenotype.
+For instance, a phenotype always causing death would have a higher multiplied value than a phenotype often causing reduced fertility (see @tbl-metric-weights).
 First, we created a dictionary to map each phenotype outcome (e.g.
-blindness) and its response (always, often, rarely, never) to numeric values from 0-3.
-Then, the phenotype outcome values were multiplied by their associated response values.
-Importantly, the values reflected the severity of each outcome based on both the outcome itself and the frequency of the response.
-For instance, a phenotype always causing death would have a higher multiplied value than a phenotype rarely causing reduced fertility (see @tbl-metric-weights).
+blindness) and its frequency (always, often, rarely, never) to numeric values from 0-3.
+Then, the phenotype outcome values were multiplied by phenotype outcome weights.
 Next, we computed an average score for each phenotype by aggregating the multiplied values across all phenotype outcomes and then calculating the mean.
@@
 The x-axes show each of the clinical characteristics.
+All data for this figure, as well as justifications for each annotation, can be found in @tbl-annotations.
 #| fig-height: 10
@@
 
+Phenotype severity annotations have utility across a wide variety of applications in both the clinic and research.
+In clinical settings, severity annotations can be used to prioritise the treatment of some phenotypes over others in patients with complex presentations, avoid administering contraindicated drugs, and prognosing potential health outcomes.
+In research settings, severity annotations can be used to identify phenotypes that have a large impact on patient outcomes and yet are currently understudied.
+They may also be used to help design new experiments and studies, or even provide new insights into the underlying aetiology of the disease by making expert-level summaries more immediately accessible to the wider research community.
+
 The creation and annotation of biomedical knowledge has traditionally relied on manual or semi-manual curation by human experts [@putmanMonarchInitiative20242024; @ochoaOpenTargetsPlatform2021; @mungallMonarchInitiativeIntegrative2017a; @kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
@@
 Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.
+In total, the entire study cost $296.27 in queries to the OpenAI API.
+While we do not have a direct comparison, this likely represents a extremely small fraction of the total costs of such a study if performed manually by human experts charging at an hourly rate.
+Even if all human annotations were provided on a volunteer basis, this would still require hundreds if not thousands of hours of cumulative manual human labour.
+Using our approach, severity annotations for the entire HPO can be generated automatically in a matter of several hours.
 
@@
 These results have a large number of useful real-world applications, such as prioritising gene therapy candidates [@murphyIdentificationCellTypespecific2023] and guiding clinical decision-making in rare diseases.
-As the HPO continues to grow over time, this approach can easily be repeated to keep pace with the rapidly evolving biomedical landscape.
-Furthermore, this approach can be extended to produce different sets of annotations or be used with any other ontology.
+It may also be used as tool to help inform policy decisions and funding allocation by healthcare or governmental institutions.
+This of course would need to be in consultation with subject matter medical experts, patients, advocates and biomedical ethicists before reaching a final decision.
+Nevertheless, access to succinct, interpretable, and semi-quantitative severity annotations may encourage key decision makers with limited time to review individual proposals to pay heed to phenotypes and diseases that would otherwise be overlooked.
+As the HPO and the broader literature continue to grow over time, our automated AI-based approach can easily be repeated to keep pace with the rapidly evolving biomedical landscape.
+Furthermore, it can be extended to produce different sets of annotations or be used with any other ontology.
 Additional use cases include gathering data on the prevalence of each phenotype to approximate their social and financial costs.
@@
 Future efforts could more explicitly ask GPT-4 whether the phenotype would cause death with no or suboptimal treatment.
+
+Another limitation with the present dataset is that phenotypes themselves can manifest with different degrees of severity, in the sense that they are more pronounced or intense.
+For example, sensitivity to light could range from a mild inconvenience to a severe disability that prevents the individual from leaving their home during the day.
+The effect of onset (beyond congenital vs.
+non-congenital) and time course (acute, slowly progression, relapse-remitting) were also not explicitly considered.
+Finally, we did not ask GPT-4 to consider phenotypes as they present within particular diseases.
+For example, while the phenotype 'Hypertension' may be mild to moderate in the general population and not present until middle-age, it can also present early in life as very severe in the context of a rare genetic disorder such as Liddle syndrome.
+Future work could explore these nuances in more detail.
 
@@
 We asked how frequently the given phenotype directly causes each clinical characteristic - whether it never, rarely, often, or always occurred.
+This design helps to constrain the potential responses of GPT-4 and thus make it more amenable to machine-readable post-processing.
+It also serves to address one of its key limitations from the @Lazarin2014-rz survey, namely the lack information on how clinical characteristic frequency affected the clinicians' severity annotations.
+Here, we can instead use the frequency values to generate more precise annotations and downstream severity ranking scores.
+
 Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance (in terms of producing the desired annotations, and adhering to the formatting requirements) was achieved when inputting no more than two or three phenotypes per prompt.
 An example prompt can be seen in @fig-occurrence.
-Only 2 phenotypes were included per prompt in order to 1) avoid exceeding per-query token limits, and 2) prevent the breakdown of GPT-4 performance due to long-form text input, which is presently a known limitation common to many LLMs including GPT-4 [@weiLongformFactualityLarge2024].
+Thus, only two phenotypes were included per prompt in order to 1) avoid exceeding per-query token limits, and 2) prevent the breakdown of GPT-4 performance due to long-form text input, which is presently a known limitation common to many LLMs including GPT-4 [@weiLongformFactualityLarge2024].
 
@@
 consist_vs_ontLvl$labels$caption <- NULL
-consist_vs_ontLvl
+(
+consist_vs_ontLvl /
+HPOExplorer::plot_arrow(x=0,xend=1, y=1, yend=1, labels_x=c(.8,.2), labels_y=c(1,1), 
+labels=c("Specific\nphenotypes","Broad\nphenotypes"))
+) + patchwork::plot_layout(heights = c(1, 0.1))
+
 ```
@@
 
+```{r top_phenos_path}
+top_phenos_path <- here::here("data","top_phenos_annotations.csv.gz")
+top_phenos <- data.table::rbindlist(plot_top_phenos_out$data)
+data.table::setcolorder(top_phenos, "variable",before="value")
+data.table::setcolorder(top_phenos, "hpo_name",after="hpo_id") 
+top_phenos <- checks$annot[hpo_id %in% top_phenos$hpo_id]
+data.table::fwrite(top_phenos, top_phenos_path)
+```
+
+::: {#tbl-annotations}
+[**Top phenotype annotations table**](https://github.com/neurogenomics/gpt_hpo_annotations/raw/master/data/`r basename(top_phenos_path)`)
+
+Table of GPT-4 annotations for all Human Phenotype Ontology (HPO) phenotypes in @fig-top-phenos.
+For each phenotype, this includes the name of the phenotype ('hpo_name'), the ID of the phenotype ('hpo_id'), the frequency of each annotation (always, often, rarely, never), and the justification for each annotation ('..._justification').
+These results can also be downloaded programmatically using the R function  `HPOExplorer::gpt_annot_check`.
+:::
+
 ```{r tbl-fas}

### 2024-05-28 docs/_tex/manuscript.tex
add query rate

@@
 approach, severity annotations for the entire HPO can be generated
-automatically in a matter of several hours.
+automatically at a rate of \textasciitilde100 phenotypes/hour. Further
+optimisation of the annotation process and increased API rate limits
+could potentially accelerate this even further.
 

### 2024-05-28 manuscript.qmd
add query rate

@@
 ```{r gpt_annot_class, results='hide'}
-res_class <- HPOExplorer::gpt_annot_class(annot_coded = res_coded$annot_coded)
-res_class <- merge(res_class,
-                   res_coded$annot_weighted[,c("hpo_name","severity_score_gpt")])
+res_class <- HPOExplorer::gpt_annot_class(res_coded = res_coded)
 fig_severity_class <- ggstatsplot::ggbetweenstats(res_class,
@@
 
+```{r weights_dict}
+weights_dict <- eval(formals(gpt_annot_codify)$weights_dict)
+annotation_order <- gsub("_"," ",names(sort(unlist(weights_dict), decreasing = TRUE)))
+```
+
 ## Abstract
-
-The Human Phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised framework for describing phenotypic abnormalities.
-However, documentation on the severity of each phenotype is currently very sparse, making systematic analyses and prioritisation of results challenging.
+There are thousands of human phenotypes which are linked to genetic variation.
+These range from the benign (white eyelashes) to the deadly (respiratory failure).
+The Human Phenotype Ontology has categorised all human phenotypic variation into an unified framework that defines the relationships between them (e.g.
+missing arms and missing legs are both abnormalities of the limb).
+This has made it possible to perform phenome-wide analyses, e.g.
+to prioritise which make the best candidates for gene therapies.
+However, there is currently limited metadata describing the effects / severity of these phenotypes.
 With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming.
@@
 #| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions.
-Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score.
-Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights.
+Heatmap of 10 represetantive phenotypes from each severity class (Profound, Severe, Moderate, Mild) stratified by whether the phenotypes are often/always congenital (**a**-**b**) or rarely/never congenital (**c**-**d**).
+Continuous severity scores are shown as bars (**b**,**d**) and were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights.
 The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score.
-The x-axis is scaled by the severity of the clinical characterisation.
-#| fig-height: 7.5
-#| fig-width: 8
+The x-axes show each of the clinical characteristics.
+All data for this figure, as well as justifications for each annotation, can be found in @tbl-annotations.
+#| fig-height: 10
+#| fig-width: 13
 
-p1 <- ggplot2::ggplot(data = dat_all,
-                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
-  ggplot2::geom_tile(colour = "white", lwd = 0.5, linetype =1) +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_brewer(palette = "GnBu", direction = -1) +
-  ggplot2::theme_classic() +
-  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
-                 legend.position = 'right') +
-  labs(x = "Clinical characteristic",
-       y = "HPO phenotype",
-       fill = "Occurrence") +
-  ggplot2::facet_grid(position ~ ., scales = "free_y")
-
-p2 <- ggplot2::ggplot(data = dat_all,
-                      ggplot2::aes(x="severity_score_gpt", y=hpo_name,
-                                   fill=severity_score_gpt)) +
-  ggplot2::geom_tile() +
-  ggplot2::scale_y_discrete(limits=rev) +
-  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = "reverse") +
-  ggplot2::labs(x=NULL, y=NULL, fill = "Severity\nscore") +
-  ggplot2::theme_void() +
-  guides(fill = guide_colorbar(reverse = TRUE)) +
-  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
-                 axis.ticks.y = ggplot2::element_blank(),
-                 legend.position = 'right')
-
-p3 <- patchwork::wrap_plots(p1, p2, ncol = 2,
-                            widths = c(1,.2),
-                            guides = "collect")
-
-methods::show(p3)
+plot_top_phenos_out <- HPOExplorer::plot_top_phenos(res_class = res_class)
+plot_top_phenos_out$plot
 ```
@@
 
+### Congenital onset by HPO branch
+
+```{r gpt_annot_plot_branches}
+gpt_annot_plot_branches_out <- HPOExplorer::gpt_annot_plot_branches(
+  gpt_annot = gpt_annot, 
+  hpo = hpo, 
+  metric = "congenital_onset", 
+  fill_lab = "Congenital\nonset", 
+  show_plot = FALSE)
+top_congenital <- unique(gpt_annot_plot_branches_out$data[!is.na(congenital_onset),c("ancestor_name","proportion_always")])
+```
+
+```{r fig-congenital-branches}
+#| label: fig-congenital-branches
+#| fig-cap: Distribution of congenital onset across HPO branches.
+The y-axis shows the proportion of phenotypes that are always/often/rarely/never congenital.
+The x-axis shows the HPO branch, orderered from highest to lowest proportion of always congenital phenotypes.
+#| fig-height: 6
+#| fig-width: 6
+
+gpt_annot_plot_branches_out$plot
+```
+
+Next, we assessed the distribution of congenital onset across HPO branches ([Fig.
+@fig-congenital-branches]).
+We found that the `r top_congenital$ancestor_name[1]` branch contained the greatest proportion of phenotypes that were always congenital (`r round(top_congenital$proportion_always[1]*100,2)`%), followed by `r top_congenital$ancestor_name[2]` (`r round(top_congenital$proportion_always[2]*100,2)`%) and `r top_congenital$ancestor_name[3]` (`r round(top_congenital$proportion_always[3]*100,2)`%).
+This is concordant with the expectation that these phenotypes should largely be congenital.
+The HPO branches with the least commonly congenital phenotypes were `r rev(top_congenital$ancestor_name)[1]` (`r round(rev(top_congenital$proportion_always)[1]*100,2)`%), `r rev(top_congenital$ancestor_name)[2]` (`r round(rev(top_congenital$proportion_always)[2]*100,2)`%), and `r rev(top_congenital$ancestor_name)[3]` (`r round(rev(top_congenital$proportion_always)[3]*100,2)`%).
+['Constitutional symptom'](https://hpo.jax.org/app/browse/term/HP:0025142) is a fairly broad term defined as *'A symptom or manifestation indicating a systemic or general effect of a disease and that may affect the general well-being or status of an individual.'* Examples include 'Fatigue' 'Exercise intolerance', 'Hot flashes' and 'Sneeze'.
+
 ## Discussion
+
+Phenotype severity annotations have utility across a wide variety of applications in both the clinic and research.
+In clinical settings, severity annotations can be used to prioritise the treatment of some phenotypes over others in patients with complex presentations, avoid administering contraindicated drugs, and prognosing potential health outcomes.
+In research settings, severity annotations can be used to identify phenotypes that have a large impact on patient outcomes and yet are currently understudied.
+They may also be used to help design new experiments and studies, or even provide new insights into the underlying aetiology of the disease by making expert-level summaries more immediately accessible to the wider research community.
 
@@
 Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.
+In total, the entire study cost $296.27 in queries to the OpenAI API.
+While we do not have a direct comparison, this likely represents a extremely small fraction of the total costs of such a study if performed manually by human experts charging at an hourly rate.
+Even if all human annotations were provided on a volunteer basis, this would still require hundreds if not thousands of hours of cumulative manual human labour.
+Using our approach, severity annotations for the entire HPO can be generated automatically at a rate of ~100 phenotypes/hour.
+Further optimisation of the annotation process and increased API rate limits could potentially accelerate this even further.
 
 Throughout this study, we observed that GPT-4 was capable of reliably recovering deep semantic relationships from the medical domain, far beyond making superficial inferences based on lexical similarities.
-An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt`), which GPT-4 annotated as causing intellectual disability.
+An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r round(res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt,2)`), which GPT-4 annotated as causing intellectual disability.
 At first glance, we ourselves assumed this was a false positive as the term appeared to be related to dentition.

### 2024-06-10 docs/_tex/manuscript.tex
KBM manuscript edits

@@
 for gene therapies. However, there is currently limited metadata
-describing the effects / severity of these phenotypes. With
-\textgreater17500 phenotypic abnormalities across \textgreater8600 rare
-diseases, manual curation of such phenotypic annotations by experts
+describing the clinical characteristics / severity of these phenotypes.
+With \textgreater17500 phenotypic abnormalities across \textgreater8600
+rare diseases, manual curation of such phenotypic annotations by experts
 would be exceedingly labour-intensive and time-consuming. Leveraging
@@
 language model (LLM) to systematically annotate the severity of all
-phenotypic abnormalities in the HPO. First, we benchmarked the
-generative LLM annotations against ground-truth labels within the HPO
+phenotypic abnormalities in the HPO. Phenotypic severity was defined
+using a set of clinical characteristics and their frequency of
+occurrence. First, we benchmarked the generative LLM clinical
+characteristic annotations against ground-truth labels within the HPO
 (e.g.~phenotypes in the `Cancer' HPO branch were annotating as causing
-cancer by GPT-4). True positive recall rates across different
-annotations ranged from 89-100\% (mean=96\%), clearly demonstrating the
-ability of GPT-4 to automate the curation process with a high degree of
-fidelity. Using a novel approach, we developed a severity scoring system
-that incorporates both the nature of the phenotype outcome and the
-frequency of its occurrence. These severity metrics will enable efforts
-to systematically prioritise which human phenotypes are most detrimental
-to human health, and best targets for therapeutic intervention.
+cancer by GPT-4). True positive recall rates across different clinical
+characteristics ranged from 89-100\% (mean=96\%), clearly demonstrating
+the ability of GPT-4 to automate the curation process with a high degree
+of fidelity. Using a novel approach, we developed a severity scoring
+system that incorporates both the nature of the clinical characteristic
+and the frequency of its occurrence. These clinical characteristic
+severity metrics will enable efforts to systematically prioritise which
+human phenotypes are most detrimental to human health, and best targets
+for therapeutic intervention.
 
@@
 phenotypes were annotated multiple times. For a subset of phenotypes
-with known expected annotations, true positive rates were calculated to
-assess recall. Additionally, based on the clinical characteristics and
-their occurrence, we have quantified the severity of each phenotype,
-providing an example of how these clinical characteristic annotations
-can be used to guide prioritisation of gene therapy trials. Ultimately,
-we hope that our resource will be of utility to those working in rare
-diseases, as well as the wider healthcare community.
+with known expected clinical characteristics, true positive rates were
+calculated to assess recall. Additionally, based on the clinical
+characteristics and their occurrence, we have quantified the severity of
+each phenotype, providing an example of how these clinical
+characteristic annotations can be used to guide prioritisation of gene
+therapy trials. Ultimately, we hope that our resource will be of utility
+to those working in rare diseases, as well as the wider healthcare
+community.
 
@@
 phenotypes based on whether they are always/often/rarely/never
-associated with different clinical characteristics. \textbf{a}, An
+associated with different clinical characteristics. \textbf{a} An
 example of the prompt input given to to GPT-4. The phenotypes listed in
 the second to last sentence (\emph{italicised}) were changed to allow
-all HPO phenotypes to be annotated. \textbf{b}, Stacked bar plot showing
+all HPO phenotypes to be annotated. \textbf{b} Stacked bar plot showing
 the proportion of the occurrence of each clinical characteristic across
@@
 that the performance of GPT-4 improved when we incorporated a scale
-associated with each effect and required a justification for each
-response. For each effect, we asked about the frequency of its
-occurrence - whether it never, rarely, often, or always occurred.
-Framing the queries in this way served two purposes. First, this helped
-to constrain the responses of GPT-4 to a specific range of values,
-making answers more consistent and amenable to downstream data analysis.
-Second, it served to overcome one of the main limitations noted by
-Lazarin et al. (2014) as they did not collect information on how the
-frequency of each disease affected their decision making when generating
-severity annotations.
-
-Phenotype outcome occurrence varied across annotation categories.
+associated with each clinical characteristic and required a
+justification for each response. For each clinical characteristic, we
+asked about the frequency of its occurrence - whether it never, rarely,
+often, or always occurred. Framing the queries in this way served two
+purposes. First, this helped to constrain the responses of GPT-4 to a
+specific range of values, making answers more consistent and amenable to
+downstream data analysis. Second, it served to overcome one of the main
+limitations noted by Lazarin et al. (2014) as they did not collect
+information on how the frequency of each disease affected their decision
+making when generating severity annotations.
+
+Clinical characteristic occurrence varied across annotation categories.
 \textgreater50\% of phenotypes never caused blindness, sensory
@@
 congenital onset, which is expected as rare disorders tend to be early
-onset genetic conditions Fig.~\ref{fig-occurrence}.
+onset genetic conditions (Fig.~\ref{fig-occurrence}).
 
@@
 \caption{\label{fig-checks}GPT-4 annotations are consistent and accurate
-across annotations. \textbf{a}, Barplot showing the annotation
+across annotations. \textbf{a} Barplot showing the annotation
 consistency within phenotypes that were annotated more than once. In the
@@
 annotations being consistent by chance in the stringent metric
-(\textasciitilde1/4). \textbf{b}, Bar plot of the true positive rate for
+(\textasciitilde1/4). \textbf{b} Bar plot of the true positive rate for
 each annotation. The labels above each bar indicate the number of
@@
 prioritisation of gene therapy trials for phenotypes with the most
-severe outcomes and thus the most urgent need. Importantly, the values
-reflected the severity of each outcome based on both the type of outcome
-itself and its frequency within a particular phenotype. For instance, a
-phenotype always causing death would have a higher multiplied value than
-a phenotype often causing reduced fertility (see
+severe clinical characteristics and thus the most urgent need.
+Importantly, the values reflected the severity of each clinical
+characteristic based on both the type of characteristic itself and its
+frequency within a particular phenotype. For instance, a phenotype
+always causing death would have a higher multiplied value than a
+phenotype often causing reduced fertility (see
 Table~\ref{tbl-metric-weights}). First, we created a dictionary to map
-each phenotype outcome (e.g.~blindness) and its frequency (always,
-often, rarely, never) to numeric values from 0-3. Then, the phenotype
-outcome values were multiplied by phenotype outcome weights. Next, we
-computed an average score for each phenotype by aggregating the
-multiplied values across all phenotype outcomes and then calculating the
-mean. This was then normalised by the theoretical maximum severity
-score, so that all phenotypes were on a 0-100 severity scale (where 100
-is the most severe phenotype possible). This average normalised score
-represents the overall severity of the phenotype based on the severity
-of its individual outcomes.
+each clinical characteristic (e.g.~blindness) and its frequency (always,
+often, rarely, never) to numeric values from 0-3. Then, the clinical
+characteristic values were multiplied by weights. Next, we computed an
+average score for each phenotype by aggregating the multiplied values
+across all clinical characteristics and then calculating the mean. This
+was then normalised by the theoretical maximum severity score, so that
+all phenotypes were on a 0-100 severity scale (where 100 is the most
+severe phenotype possible). This average normalised score represents the
+overall severity of the phenotype based on the severity of its
+individual clinical characteristics.
 
@@
 related to developmental brain and neural tube defects. Comparison of
-the severity scores for each response, across the phenotype outcomes
-annotated, revealed consistent trends: as the response of the phenotype
-outcome increased (from never to always), the severity score also
-increased (Supplementary Fig.~\ref{fig-severity-boxplot}). We also
-evaluated the severity score distribution by HPO branch and calculated
-the mean severity score using all phenotypes within each major HPO
-branch (Fig.~\ref{fig-severity-histo}). The HPO branch with the greatest
-mean severity score was `Abnormal cellular phenotype' (mean=17),
-followed by `Neoplasm' (mean=15.2), which would include the highly
-ranked phenotypes seen in Figure~\ref{fig-top-phenos}.
+the severity scores for each response, across the clinical
+characteristics annotated, revealed consistent trends: as the response
+of the clinical characteristic increased (from never to always), the
+severity score also increased (Supplementary
+Fig.~\ref{fig-severity-boxplot}). We also evaluated the severity score
+distribution by HPO branch and calculated the mean severity score using
+all phenotypes within each major HPO branch
+(Fig.~\ref{fig-severity-histo}). The HPO branch with the greatest mean
+severity score was `Abnormal cellular phenotype' (mean=17), followed by
+`Neoplasm' (mean=16.7), which would include the highly ranked phenotypes
+seen in Figure~\ref{fig-top-phenos}.
 
@@
 
-\subsubsection{Correlations between severity
-metrics}\label{correlations-between-severity-metrics}
-
-We found that some severity metrics were correlated with one another,
-with a mean Pearson correlation of 0.2 across all individual metrics
-(see Figure~\ref{fig-metric-corplot}). In particular, blindness and
-sensory impairment were highly correlated with one another (r=0.62,
-p=0). Some metrics drove the composite severity score more than other,
-which is a reflection of both our per-metric weighting scheme, response
-type frequencies, and the correlation structure between metrics.
-Overall, impaired mobility seemed to be the strongest driver of the
-composite severity score with a Pearson correlation of 0.6001824,
-followed by intellectual disability (r=0.59) and death (r=0.56).
+\subsubsection{Correlations between clinical characteristic severity
+metrics}\label{correlations-between-clinical-characteristic-severity-metrics}
+
+We found that some clinical characteristic severity metrics were
+correlated with one another, with a mean Pearson correlation of 0.2
+across all individual metrics (see Figure~\ref{fig-metric-corplot}). In
+particular, blindness and sensory impairment were highly correlated with
+one another (r=0.62, p=0). Some metrics drove the composite severity
+score more than other, which is a reflection of both our per-metric
+weighting scheme, response type frequencies, and the correlation
+structure between metrics. Overall, impaired mobility seemed to be the
+strongest driver of the composite severity score with a Pearson
+correlation of 0.6001824, followed by intellectual disability (r=0.59)
+and death (r=0.56).
 
@@
 A key contribution of our study is the introduction of a quantitative
-severity scoring system that integrates both the nature of the phenotype
-outcome and the frequency of its occurrence. By encoding the concept of
-severity in this way, we are able to prioritise phenotypes based on
-their impact on patients. The methodology allowed us to transition from
-low-throughput qualitative assessments of severity (e.g. Lazarin et al.
-(2014)) to high-throughput quantitative assessments of severity. One of
-the most severe phenotypes in the HPO is `Fetal akinesia sequence' (FAS;
-HP:0001989, severity score= 43.9), and extremely rare condition that is
-almost always lethal. FAS is a complex, multi-system phenotype that can
-be caused by at least 24 different genetic disorders. Despite the
-complex and heterogeneous aetiology of this phenotype, GPT-4 was able to
-provide accurate annotations alongside explainable justifications for
-those annotations (see Table~\ref{tbl-fas}). For example, this phenotype
-almost always results in death, either \emph{in utero} or shortly after
-birth. Not only did GPT-4 correctly provide the annotation death as
-`always', when asked whether FAS causes sensory impairments it provided
-the response `always' with the justification `Fetal akinesia sequence
-typically results in severe sensory impairment due to neurodevelopmental
-disruption.' Neurodevelopmental disruption is indeed a hallmark
-component of FAS (e.g.~hydrocephalus, cerebellar hypoplasia) that causes
-severe impairments across multiple sensory systems (Chen, 2012). This
+severity scoring system that integrates both the nature of the clinical
+characteristic and the frequency of its occurrence. By encoding the
+concept of severity in this way, we are able to prioritise phenotypes
+based on their impact on patients. The methodology allowed us to
+transition from low-throughput qualitative assessments of severity (e.g.
+Lazarin et al. (2014)) to high-throughput quantitative assessments of
+severity. One of the most severe phenotypes in the HPO is `Fetal
+akinesia sequence' (FAS; HP:0001989, severity score= 43.9), and
+extremely rare condition that is almost always lethal. FAS is a complex,
+multi-system phenotype that can be caused by at least 24 different
+genetic disorders. Despite the complex and heterogeneous aetiology of
+this phenotype, GPT-4 was able to provide accurate annotations alongside
+explainable justifications for those annotations (see
+Table~\ref{tbl-fas}). For example, this phenotype almost always results
+in death, either \emph{in utero} or shortly after birth. Not only did
+GPT-4 correctly provide the annotation death as `always', when asked
+whether FAS causes sensory impairments it provided the response `always'
+with the justification `Fetal akinesia sequence typically results in
+severe sensory impairment due to neurodevelopmental disruption.'
+Neurodevelopmental disruption is indeed a hallmark component of FAS
+(e.g.~hydrocephalus, cerebellar hypoplasia) that causes severe
+impairments across multiple sensory systems (Chen, 2012). This
 demonstrates that GPT-4 was able to recover the correct chain of
-causality from phenotype to clinical outcome.
+causality from phenotype to clinical characteristic.
 
@@
 performance of GPT-4 improved when we incorporated a scale associated
-with each effect and required a justification for each response. We
-asked how frequently the given phenotype directly causes each clinical
-characteristic - whether it never, rarely, often, or always occurred.
-This design helps to constrain the potential responses of GPT-4 and thus
-make it more amenable to machine-readable post-processing. It also
-serves to address one of its key limitations from the Lazarin et al.
-(2014) survey, namely the lack information on how clinical
-characteristic frequency affected the clinicians' severity annotations.
-Here, we can instead use the frequency values to generate more precise
-annotations and downstream severity ranking scores.
+with each clinical characteristic and required a justification for each
+response. We asked how frequently the given phenotype directly causes
+each clinical characteristic - whether it never, rarely, often, or
+always occurred. This design helps to constrain the potential responses
+of GPT-4 and thus make it more amenable to machine-readable
+post-processing. It also serves to address one of its key limitations
+from the Lazarin et al. (2014) survey, namely the lack information on
+how clinical characteristic frequency affected the clinicians' severity
+annotations. Here, we can instead use the frequency values to generate
+more precise annotations and downstream severity ranking scores.
 
@@
 
-\caption{\label{tbl-true-positives}For each clinical characteristic, any
-terms that matched}
+\caption{\label{tbl-true-positives}The HPO branches and their
+descendants used as true positives for each clinical characteristic.}
 
@@
 
-The GPT-4 generated annotation occurrences were converted into a
-semi-quantitative scoring system, with `always' corresponding to 3,
-`often' to 2, `rarely' to 1, and `never' to 0. These scores were then
+The GPT-4 generated clinical characteristic occurrences were converted
+into a semi-quantitative scoring system, with `always' corresponding to
+3, `often' to 2, `rarely' to 1, and `never' to 0. These scores were then
 weighted by a severity metric on a scale of 1-5, with 5 representing the
-highest severity, as determined by the provided annotations
+highest severity, as determined by the provided clinical characteristics
 (Table~\ref{tbl-metric-weights}). Subsequently, the weighted scores
@@
 \end{itemize}
+
+\vspace{2em}
 
@@
 The decision tree algorithm used in Lazarin et al. (2014) was adapted
-here for use with the GPT-4 annotations. This algorithm first assigned
-each annotation metric to a tier, where Tier 1 indicated the most severe
-annotation metrics and Tier 4 indicated the least severe annotation
-metrics (`death'=1, `intellectual disability'=1, `impaired mobility'=2,
-`physical malformations'=2, `blindness'=3, `sensory impairments'=3,
+here for use with the GPT-4 clinical characteristic annotations. This
+algorithm first assigned each clinical chacteristic to a tier, where
+Tier 1 indicated the most severe clinical characteristics and Tier 4
+indicated the least severe clinical characteristics (`death'=1,
+`intellectual disability'=1, `impaired mobility'=2, `physical
+malformations'=2, `blindness'=3, `sensory impairments'=3,
 `immunodeficiency'=3, `cancer'=3, `reduced fertility'=4). If a phenotype
-often or always caused more than one Tier 1 annotation, it was assigned
-a severity class of ``Profound''. If the phenotype often or always
-caused only one Tier 1 annotation, it was assigned a severity class of
-``Severe''. A ``Severe'' class assignment was also assigned if the
-phenotype often or always caused three or more Tier 2 and Tier3
-annotations. If the phenotype often or always caused at least one Tier 2
-annotation, it was assigned a severity class of ``Moderate''. All
-remaining phenotypes were was assigned a severity class of ``Mild''. In
-cases where the phenotype mapped to more than one class, only the most
-severe class was used. This procedure is implemented within the function
+often or always caused more than one Tier 1 clinical characteristic, it
+was assigned a severity class of ``Profound''. If the phenotype often or
+always caused only one Tier 1 clinical characteristic, it was assigned a
+severity class of ``Severe''. A ``Severe'' class assignment was also
+assigned if the phenotype often or always caused three or more Tier 2
+and Tier3 clinical characteristics. If the phenotype often or always
+caused at least one Tier 2 clinical characteristic, it was assigned a
+severity class of ``Moderate''. All remaining phenotypes were was
+assigned a severity class of ``Mild''. In cases where the phenotype
+mapped to more than one class, only the most severe class was used. This
+procedure is implemented within the function
 \texttt{HPOExplorer::gpt\_annot\_class}.
 
-\subsubsection{Correlations between severity
-metrics}\label{correlations-between-severity-metrics-1}
-
-To assess the correlation structure between each severity metric, as
-well as between the composite severity score and each metric, we
-computed Pearson correlation coefficients for all pairwise combinations
-of these variables using the numerically encoded metric values. The
-correlation matrix was visualised using a heatmap, with the colour
-intensity representing the strength of the correlation
+\subsubsection{Correlations between clinical characteristic severity
+metrics}\label{correlations-between-clinical-characteristic-severity-metrics-1}
+
+To assess the correlation structure between each clinical characteristic
+severity metric, as well as between the composite severity score and
+each metric, we computed Pearson correlation coefficients for all
+pairwise combinations of these variables using the numerically encoded
+metric values. The correlation matrix was visualised using a heatmap,
+with the colour intensity representing the strength of the correlation
 (Figure~\ref{fig-metric-corplot}).
@@
 
-The GPT-4 annotations for all HPO phenotypes are made available through
-the R function \texttt{HPOExplorer::gpt\_annot\_read} or in CSV format
-at:
+The GPT-4 clinical characteristic annotations for all HPO phenotypes are
+made available through the R function
+\texttt{HPOExplorer::gpt\_annot\_read} or in CSV format at:
 
@@
 \caption{\label{fig-consist-vs-ontLvl}Relationship between the
-consistency of GPT-4 annotations (using the stringent criterion) and the
-level of each phenotype within the HPO ontology (with the number of
-phenotypes in parentheses). Greater ontology levels (x-axis) indicate
-more specific phenotypes. The subtitle indicates summary statistics for
-the overall relationship between HPO level and the proportion of
-phenotypes that were annotated consistently. The p-values above each bar
-indicate whether the distribution of consistent/inconsistent
-annotations, within a given HPO level, significantly deviate from the
-expected null distribution.}
+consistency of GPT-4 clinical characterstic annotations (using the
+stringent criterion) and the level of each phenotype within the HPO
+ontology (with the number of phenotypes in parentheses). Greater
+ontology levels (x-axis) indicate more specific phenotypes. The subtitle
+indicates summary statistics for the overall relationship between HPO
+level and the proportion of phenotypes that were annotated consistently.
+The p-values above each bar indicate whether the distribution of
+consistent/inconsistent annotations, within a given HPO level,
+significantly deviate from the expected null distribution.}
 
@@
 between composite severity score (y-axis) and the frequency response
-categories within each annotation type.}
+categories within each clinical characteristic type.}
 
@@
 \caption{\label{fig-metric-corplot}Pearson correlations between each
-individual severity metric and the composite severity score
-(`severity\_score\_gpt').}
+individual clinical characteristic severity metric and the composite
+severity score (`severity\_score\_gpt').}
 
@@
 
-\caption{\label{tbl-annotations}Table of GPT-4 annotations for all Human
-Phenotype Ontology (HPO) phenotypes in Figure~\ref{fig-top-phenos}. For
-each phenotype, this includes the name of the phenotype (`hpo\_name'),
-the ID of the phenotype (`hpo\_id'), the frequency of each annotation
-(always, often, rarely, never), and the justification for each
-annotation ('\ldots\_justification'). These results can also be
-downloaded programmatically using the R function
+\caption{\label{tbl-annotations}Table of GPT-4 clinical characterstic
+annotations for all Human Phenotype Ontology (HPO) phenotypes in
+Figure~\ref{fig-top-phenos}. For each phenotype, this includes the name
+of the phenotype (`hpo\_name'), the ID of the phenotype (`hpo\_id'), the
+frequency of each annotation (always, often, rarely, never), and the
+justification for each annotation ('\ldots\_justification'). These
+results can also be downloaded programmatically using the R function
 \texttt{HPOExplorer::gpt\_annot\_check}.}
@@
 
-\caption{\label{tbl-fas}Severity nnotations generated for GPT-4
-annotations for the HPO phenotype `Fetal akinesia sequence'
-(HP:000198).}
+\caption{\label{tbl-fas}Severity nnotations generated for GPT-4 clinical
+characteristic annotations for the HPO phenotype `Fetal akinesia
+sequence' (HP:000198).}
 

### 2024-06-10 docs/manuscript.qmd
KBM manuscript edits

@@
                             save_dir=save_dir)
+
 hpo <- KGExplorer::filter_ontology(ont = hpo, 
@@
 to prioritise which make the best candidates for gene therapies.
-However, there is currently limited metadata describing the effects / severity of these phenotypes.
+However, there is currently limited metadata describing the clinical characteristics / severity of these phenotypes.
 With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming.
 Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 large language model (LLM) to systematically annotate the severity of all phenotypic abnormalities in the HPO.
-First, we benchmarked the generative LLM annotations against ground-truth labels within the HPO (e.g.
+Phenotypic severity was defined using a set of clinical characteristics and their frequency of occurrence.
+First, we benchmarked the generative LLM clinical characteristic annotations against ground-truth labels within the HPO (e.g.
 phenotypes in the 'Cancer' HPO branch were annotating as causing cancer by GPT-4).
-True positive recall rates across different annotations ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity.
-Using a novel approach, we developed a severity scoring system that incorporates both the nature of the phenotype outcome and the frequency of its occurrence.
-These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.
+True positive recall rates across different clinical characteristics ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity.
+Using a novel approach, we developed a severity scoring system that incorporates both the nature of the clinical characteristic and the frequency of its occurrence.
+These clinical characteristic severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.
 
@@
 To evaluate the consistency of responses generated by GPT-4 `r checks$consistency_count[[1]]` phenotypes were annotated multiple times.
-For a subset of phenotypes with known expected annotations, true positive rates were calculated to assess recall.
+For a subset of phenotypes with known expected clinical characteristics, true positive rates were calculated to assess recall.
 Additionally, based on the clinical characteristics and their occurrence, we have quantified the severity of each phenotype, providing an example of how these clinical characteristic annotations can be used to guide prioritisation of gene therapy trials.
@@
 #| fig-cap: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics.
-**a**, An example of the prompt input given to to GPT-4.
+**a** An example of the prompt input given to to GPT-4.
 The phenotypes listed in the second to last sentence (*italicised*) were changed to allow all HPO phenotypes to be annotated.
-**b**, Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all annotated HPO phenotypes.
+**b** Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all annotated HPO phenotypes.
 The terms shown on the x-axis are the clinical characteristics for which GPT-4 was asked to determine whether each phenotype caused them.
@@
 We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset.
-Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
-For each effect, we asked about the frequency of its occurrence - whether it never, rarely, often, or always occurred.
+Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each clinical characteristic and required a justification for each response.
+For each clinical characteristic, we asked about the frequency of its occurrence - whether it never, rarely, often, or always occurred.
 Framing the queries in this way served two purposes.
@@
 
-Phenotype outcome occurrence varied across annotation categories.
+Clinical characteristic occurrence varied across annotation categories.
 \>`r never_thresh`% of phenotypes never caused `r paste(never_dt$category[-1], collapse=", ")` or `r never_dt$category[1]`.
-Only a minority of phenotypes (`r round(sum(subset(count_df_long, condition %in% c("never") & category=="congenital onset")$percentage),1)`%) never had a congenital onset, which is expected as rare disorders tend to be early onset genetic conditions [Fig.
-@fig-occurrence].
+Only a minority of phenotypes (`r round(sum(subset(count_df_long, condition %in% c("never") & category=="congenital onset")$percentage),1)`%) never had a congenital onset, which is expected as rare disorders tend to be early onset genetic conditions ([Fig.
+@fig-occurrence]).
 
@@
 #| fig-cap: GPT-4 annotations are consistent and accurate across annotations.
-**a**, Barplot showing the annotation consistency within phenotypes that were annotated more than once.
+**a** Barplot showing the annotation consistency within phenotypes that were annotated more than once.
 In the lenient metric, annotations were collapsed into two groups ('always'/'often' and 'never'/'rarely').
@@
 The gold dashed line is the probability of two annotations being consistent by chance in the stringent metric (~1/4).
-**b**, Bar plot of the true positive rate for each annotation.
+**b** Bar plot of the true positive rate for each annotation.
 The labels above each bar indicate the number of phenotypes tested.
@@
 Quantifying the overall severity of phenotypes can have important implications for diagnosis, prognosis, and treatment.
-It may also guide the prioritisation of gene therapy trials for phenotypes with the most severe outcomes and thus the most urgent need.
-Importantly, the values reflected the severity of each outcome based on both the type of outcome itself and its frequency within a particular phenotype.
+It may also guide the prioritisation of gene therapy trials for phenotypes with the most severe clinical characteristics and thus the most urgent need.
+Importantly, the values reflected the severity of each clinical characteristic based on both the type of characteristic itself and its frequency within a particular phenotype.
 For instance, a phenotype always causing death would have a higher multiplied value than a phenotype often causing reduced fertility (see @tbl-metric-weights).
-First, we created a dictionary to map each phenotype outcome (e.g.
+First, we created a dictionary to map each clinical characteristic (e.g.
 blindness) and its frequency (always, often, rarely, never) to numeric values from 0-3.
-Then, the phenotype outcome values were multiplied by phenotype outcome weights.
-Next, we computed an average score for each phenotype by aggregating the multiplied values across all phenotype outcomes and then calculating the mean.
+Then, the clinical characteristic values were multiplied by weights.
+Next, we computed an average score for each phenotype by aggregating the multiplied values across all clinical characteristics and then calculating the mean.
 This was then normalised by the theoretical maximum severity score, so that all phenotypes were on a 0-100 severity scale (where 100 is the most severe phenotype possible).
-This average normalised score represents the overall severity of the phenotype based on the severity of its individual outcomes.
+This average normalised score represents the overall severity of the phenotype based on the severity of its individual clinical characteristics.
 
@@
 In fact, many of the most severe phenotypes were related to developmental brain and neural tube defects.
-Comparison of the severity scores for each response, across the phenotype outcomes annotated, revealed consistent trends: as the response of the phenotype outcome increased (from never to always), the severity score also increased ([Supplementary Fig.
+Comparison of the severity scores for each response, across the clinical characteristics annotated, revealed consistent trends: as the response of the clinical characteristic increased (from never to always), the severity score also increased ([Supplementary Fig.
 @fig-severity-boxplot]).
@@
 
-### Correlations between severity metrics
+### Correlations between clinical characteristic severity metrics
 
@@
 
-We found that some severity metrics were correlated with one another, with a mean Pearson correlation of `r round(mean(cor_metrics$r[-nrow(cor_metrics$r),-ncol(cor_metrics$r)]),2)` across all individual metrics (see @fig-metric-corplot).
+We found that some clinical characteristic severity metrics were correlated with one another, with a mean Pearson correlation of `r round(mean(cor_metrics$r[-nrow(cor_metrics$r),-ncol(cor_metrics$r)]),2)` across all individual metrics (see @fig-metric-corplot).
 In particular, blindness and sensory impairment were highly correlated with one another (r=`r round(cor_metrics$r["blindness","sensory_impairments"],2)`, p=`r round(cor_metrics$P["blindness","sensory_impairments"],4)`).
@@
 Even if all human annotations were provided on a volunteer basis, this would still require hundreds if not thousands of hours of cumulative manual human labour.
-Using our approach, severity annotations for the entire HPO can be generated automatically in a matter of several hours.
+Using our approach, severity annotations for the entire HPO can be generated automatically at a rate of ~100 phenotypes/hour.
+Further optimisation of the annotation process and increased API rate limits could potentially accelerate this even further.
 
@@
 
-A key contribution of our study is the introduction of a quantitative severity scoring system that integrates both the nature of the phenotype outcome and the frequency of its occurrence.
+A key contribution of our study is the introduction of a quantitative severity scoring system that integrates both the nature of the clinical characteristic and the frequency of its occurrence.
 By encoding the concept of severity in this way, we are able to prioritise phenotypes based on their impact on patients.
@@
 hydrocephalus, cerebellar hypoplasia) that causes severe impairments across multiple sensory systems [@chenPrenatalDiagnosisGenetic2012].
-This demonstrates that GPT-4 was able to recover the correct chain of causality from phenotype to clinical outcome.
+This demonstrates that GPT-4 was able to recover the correct chain of causality from phenotype to clinical characteristic.
 
@@
 We sought to evaluate whether each phenotype directly caused a given severity-related clinical characteristic, including: intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and/or had a congenital onset.
-Through prompt engineering we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response.
+Through prompt engineering we found that the performance of GPT-4 improved when we incorporated a scale associated with each clinical characteristic and required a justification for each response.
 We asked how frequently the given phenotype directly causes each clinical characteristic - whether it never, rarely, often, or always occurred.
@@
 #| label: tbl-true-positives
-#| tbl-cap: For each clinical characteristic, any terms that matched 
+#| tbl-cap: The HPO branches and their descendants used as true positives for each clinical characteristic.   
 #| tbl-colwidths: [40,40,20]
@@
 
-The GPT-4 generated annotation occurrences were converted into a semi-quantitative scoring system, with 'always' corresponding to 3, 'often' to 2, 'rarely' to 1, and 'never' to 0.
-These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided annotations (@tbl-metric-weights).
+The GPT-4 generated clinical characteristic occurrences were converted into a semi-quantitative scoring system, with 'always' corresponding to 3, 'often' to 2, 'rarely' to 1, and 'never' to 0.
+These scores were then weighted by a severity metric on a scale of 1-5, with 5 representing the highest severity, as determined by the provided clinical characteristics (@tbl-metric-weights).
 Subsequently, the weighted scores underwent normalisation to yield a final quantitative severity score ranging from 0-100, with 100 signifying the maximum severity score attainable.
@@
 This allows for direct comparability of severity scores across studies with different sets of phenotypes.
+
+\vspace{2em}
 
@@
 
-The decision tree algorithm used in @Lazarin2014-rz was adapted here for use with the GPT-4 annotations.
-This algorithm first assigned each annotation metric to a tier, where Tier 1 indicated the most severe annotation metrics and Tier 4 indicated the least severe annotation metrics (`r paste(paste0(shQuote(gsub("_"," ",names(tiers_dict))),"=",tiers_dict), collapse = ", ")`).
-If a phenotype often or always caused more than one Tier 1 annotation, it was assigned a severity class of "Profound".
-If the phenotype often or always caused only one Tier 1 annotation, it was assigned a severity class of "Severe".
-A "Severe" class assignment was also assigned if the phenotype often or always caused three or more Tier 2 and Tier3 annotations.
-If the phenotype often or always caused at least one Tier 2 annotation, it was assigned a severity class of "Moderate".
+The decision tree algorithm used in @Lazarin2014-rz was adapted here for use with the GPT-4 clinical characteristic annotations.
+This algorithm first assigned each clinical chacteristic to a tier, where Tier 1 indicated the most severe clinical characteristics and Tier 4 indicated the least severe clinical characteristics (`r paste(paste0(shQuote(gsub("_"," ",names(tiers_dict))),"=",tiers_dict), collapse = ", ")`).
+If a phenotype often or always caused more than one Tier 1 clinical characteristic, it was assigned a severity class of "Profound".
+If the phenotype often or always caused only one Tier 1 clinical characteristic, it was assigned a severity class of "Severe".
+A "Severe" class assignment was also assigned if the phenotype often or always caused three or more Tier 2 and Tier3 clinical characteristics.
+If the phenotype often or always caused at least one Tier 2 clinical characteristic, it was assigned a severity class of "Moderate".
 All remaining phenotypes were was assigned a severity class of "Mild".
@@
 
-### Correlations between severity metrics
+### Correlations between clinical characteristic severity metrics
 
-To assess the correlation structure between each severity metric, as well as between the composite severity score and each metric, we computed Pearson correlation coefficients for all pairwise combinations of these variables using the numerically encoded metric values.
+To assess the correlation structure between each clinical characteristic severity metric, as well as between the composite severity score and each metric, we computed Pearson correlation coefficients for all pairwise combinations of these variables using the numerically encoded metric values.
 The correlation matrix was visualised using a heatmap, with the colour intensity representing the strength of the correlation (@fig-metric-corplot).
@@
 
-The GPT-4 annotations for all HPO phenotypes are made available through the R function `HPOExplorer::gpt_annot_read` or in CSV format at:
+The GPT-4 clinical characteristic annotations for all HPO phenotypes are made available through the R function `HPOExplorer::gpt_annot_read` or in CSV format at:
 
@@
 #| label: fig-consist-vs-ontLvl
-#| fig-cap: Relationship between the consistency of GPT-4 annotations (using the stringent criterion) and the level of each phenotype within the HPO ontology (with the number of phenotypes in parentheses).
+#| fig-cap: Relationship between the consistency of GPT-4 clinical characterstic annotations (using the stringent criterion) and the level of each phenotype within the HPO ontology (with the number of phenotypes in parentheses).
 Greater ontology levels (x-axis) indicate more specific phenotypes.
@@
 #| label: fig-severity-boxplot
-#| fig-cap: Boxplot showing the relationship between composite severity score (y-axis) and the frequency response categories within each annotation type.
+#| fig-cap: Boxplot showing the relationship between composite severity score (y-axis) and the frequency response categories within each clinical characteristic type.
 #| fig-height: 5
@@
 #| label: fig-metric-corplot
-#| fig-cap: Pearson correlations between each individual severity metric and the composite severity score ('severity_score_gpt').
+#| fig-cap: Pearson correlations between each individual clinical characteristic severity metric and the composite severity score ('severity_score_gpt').
 #| fig-height: 8
@@
 
-Table of GPT-4 annotations for all Human Phenotype Ontology (HPO) phenotypes in @fig-top-phenos.
+Table of GPT-4 clinical characterstic annotations for all Human Phenotype Ontology (HPO) phenotypes in @fig-top-phenos.
 For each phenotype, this includes the name of the phenotype ('hpo_name'), the ID of the phenotype ('hpo_id'), the frequency of each annotation (always, often, rarely, never), and the justification for each annotation ('..._justification').
@@
 #| label: tbl-fas
-#| tbl-cap: Severity nnotations generated for GPT-4 annotations for the HPO phenotype 'Fetal akinesia sequence' (HP:000198).
+#| tbl-cap: Severity nnotations generated for GPT-4 clinical characteristic annotations for the HPO phenotype 'Fetal akinesia sequence' (HP:000198).
 #| tbl-colwidths: [20,20,60]

### 2024-06-10 manuscript.qmd
KBM manuscript edits

@@
 ## Abstract
+
 There are thousands of human phenotypes which are linked to genetic variation.
@@
 `r uniqueN(gpt_annot[death %in% c("never")]$hpo_name)` phenotypes were annotated as never causing death.
-Examples of phenotypes that never cause death included `r sum(grepl("syndactyly",gpt_annot[death %in% c("never")]$hpo_name, ignore.case = TRUE))` unique forms of syndactyly, a non-lethal condition that causes fused or webbed fingers (occurring 1 in 1,200–15,000 live births).
+Examples of phenotypes that never cause death included `r sum(grepl("syndactyly",gpt_annot[death %in% c("never")]$hpo_name, ignore.case = TRUE))` unique forms of syndactyly, a non-lethal condition that causes fused or webbed fingers (occurring 1 in 1,200--15,000 live births).
 While not life-threatening itself, syndactyly is a symptom of genetic disorders that can cause life-threatening cardiovascular and neurodevelopmental defects, such as Apert Syndrome [@garagnaniSyndromesAssociatedSyndactyly2013].
@@
 We approximated this approach using our GPT-4 annotations.
-This categorical approach showed a strong degree of positive correspondence with the continuous severity score ($\hat{\omega_{p}^2}$=`r round(fig_severity_class_dt$summary_data$estimate,2)`, p<`r format(.Machine$double.xmin, digits = 2)`).
-In other words, severity score increased with severity class level (mild < moderate < severe < profound) as expected.
+This categorical approach showed a strong degree of positive correspondence with the continuous severity score ($\hat{\omega_{p}^2}$=`r round(fig_severity_class_dt$summary_data$estimate,2)`, p\<`r format(.Machine$double.xmin, digits = 2)`).
+In other words, severity score increased with severity class level (mild \< moderate \< severe \< profound) as expected.
 The distribution of severity classes is shown in @fig-severity-class.
 
-### Correlations between severity metrics
+### Correlations between clinical characteristic severity metrics
 
@@
 Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.
-In total, the entire study cost $296.27 in queries to the OpenAI API.
+In total, the entire study cost \$296.27 in queries to the OpenAI API.
 While we do not have a direct comparison, this likely represents a extremely small fraction of the total costs of such a study if performed manually by human experts charging at an hourly rate.
 Even if all human annotations were provided on a volunteer basis, this would still require hundreds if not thousands of hours of cumulative manual human labour.
-Using our approach, severity annotations for the entire HPO can be generated automatically at a rate of ~100 phenotypes/hour.
+Using our approach, severity annotations for the entire HPO can be generated automatically at a rate of \~100 phenotypes/hour.
 Further optimisation of the annotation process and increased API rate limits could potentially accelerate this even further.
@@
 -   $NSS_p$: the final composite severity score for phenotype $p$ after applying normalisation to align values to a 0-100 scale and ensure equivalent meaning regardless of which other phenotypes are being analysed in addition to $p$.
-This allows for direct comparability of severity scores across studies with different sets of phenotypes.
+This allows for direct comparability of severity scores across studies with different sets of phenotypes.\
 
@@
 :::
-
-\
-\
 
@@
 
-This work was supported by a UK Dementia Research Institute (UK DRI) Future Leaders Fellowship \[MR/T04327X/1\] and the UK DRI which receives its funding from UK DRI Ltd, funded by the UK Medical Research Council, Alzheimer’s Society and Alzheimer's Research UK.
+This work was supported by a UK Dementia Research Institute (UK DRI) Future Leaders Fellowship \[MR/T04327X/1\] and the UK DRI which receives its funding from UK DRI Ltd, funded by the UK Medical Research Council, Alzheimer's Society and Alzheimer's Research UK.
 
@@
 ::: {#tbl-annotations}
-[**Top phenotype annotations table**](https://github.com/neurogenomics/gpt_hpo_annotations/raw/master/data/`r basename(top_phenos_path)`)
+[**Top phenotype annotations table**](https://github.com/neurogenomics/gpt_hpo_annotations/raw/master/data/%60r%20basename(top_phenos_path)%60)
 
-Table of GPT-4 annotations for all Human Phenotype Ontology (HPO) phenotypes in @fig-top-phenos.
-For each phenotype, this includes the name of the phenotype ('hpo_name'), the ID of the phenotype ('hpo_id'), the frequency of each annotation (always, often, rarely, never), and the justification for each annotation ('..._justification').
+Table of GPT-4 clinical characterstic annotations for all Human Phenotype Ontology (HPO) phenotypes in @fig-top-phenos.
+For each phenotype, this includes the name of the phenotype ('hpo_name'), the ID of the phenotype ('hpo_id'), the frequency of each annotation (always, often, rarely, never), and the justification for each annotation ('...\_justification').
 These results can also be downloaded programmatically using the R function `HPOExplorer::gpt_annot_check`.

### 2024-06-10 manuscript.qmd
Typo fix

@@
 These range from the benign (white eyelashes) to the deadly (respiratory failure).
-The Human Phenotype Ontology has categorised all human phenotypic variation into an unified framework that defines the relationships between them (e.g.
+The Human Phenotype Ontology has categorised all human phenotypic variation into a unified framework that defines the relationships between them (e.g.
 missing arms and missing legs are both abnormalities of the limb).
@@
 First, we benchmarked the generative LLM clinical characteristic annotations against ground-truth labels within the HPO (e.g.
-phenotypes in the 'Cancer' HPO branch were annotating as causing cancer by GPT-4).
+phenotypes in the 'Cancer' HPO branch were annotated as causing cancer by GPT-4).
 True positive recall rates across different clinical characteristics ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity.

### 2024-11-15 manuscript.qmd
Edit manuscript

@@
 Since 2008, the Human Phenotype Ontology (HPO) has been instrumental in healthcare and biomedical research by providing a framework for comprehensively describing human phenotypes and the relationships between them [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024].
-By expanding its depth and breadth over time, the HPO now contains \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases.
-Some HPO phenotypes also contain metadata annotations such typical age of onset, frequency, triggers, time course, mortality rate and typical severity.
+By expanding its depth and breadth over time, the HPO now contains \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities that are associated with \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases as captured in HPOA [REF].
+Some HPO phenotypes also contain information related to typical age of onset, frequency, triggers, time course, mortality rate and typical severity.
 Describing the severity-related attributes of a disease is crucial for both research and clinical care of individuals with rare diseases.
@@
 
-plot_top_phenos_out <- HPOExplorer::plot_top_phenos(res_class = res_class)
+plot_top_phenos_out <- HPOExplorer::plot_top_phenos(res_class = res_class, axis.text.x=c(TRUE,TRUE))
 plot_top_phenos_out$plot
@@
 -   $NSS_p$: the final composite severity score for phenotype $p$ after applying normalisation to align values to a 0-100 scale and ensure equivalent meaning regardless of which other phenotypes are being analysed in addition to $p$.
-This allows for direct comparability of severity scores across studies with different sets of phenotypes.\
+This allows for direct comparability of severity scores across studies with different sets of phenotypes.
+
+{{< pagebreak >}
 

### 2025-01-07 manuscript.qmd
edit manuscript

@@
 ---
-title: Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
+title: Using GPT-4 to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
 author:
@@
 These range from the benign (white eyelashes) to the deadly (respiratory failure).
-The Human Phenotype Ontology has categorised all human phenotypic variation into a unified framework that defines the relationships between them (e.g.
+The Human Phenotype Ontology has categorised a vast array of human phenotypic variation into a unified framework that defines the relationships between them (e.g.
 missing arms and missing legs are both abnormalities of the limb).
@@
 This high level of recall underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information.
-However, we acknowledge that the number of testable true positive phenotypes for some of these categories are low, especially `r shQuote(names(checks$checkable_count)[checks$checkable_count==min(checks$checkable_count)])` for which there is only `r min(checks$checkable_count)` phenotype in the HPO (after excluding terms pertaining to colour or night blindness).
+However, we acknowledge that the number of testable true positive phenotypes for some of these categories are low, especially `r shQuote(names(checks$checkable_count)[checks$checkable_count==min(checks$checkable_count)])` for which there is only `r min(checks$checkable_count)` phenotype in the HPO (after excluding terms pertaining to colour or night blindness, as these are not typically classified as causes of complete blindness).
 Furthermore, some of the true positive phenotypes are lexically similar to the name of the clinical characteristic itself.
@@
 This average normalised score represents the overall severity of the phenotype based on the severity of its individual clinical characteristics.
+@fig-severity-histo illustrates the distribution of severity scores across HPO terms by branch, with 'Abnormal cellular phenotype' being the branch that contains phenotypes with the highest severity scores on average.
+
+```{r fig-severity-histo}
+#| label: fig-severity-histo
+#| fig-cap: Distribution of composite GPT-4 severity scores across HPO terms, grouped by branch.
+The figure shows the distribution of phenotypes for each severity score (ranging from 0 to 100) across HPO branches.
+The vertical line indicates the average severity score for each branch.
+The branch 'Abnormal cellular phenotype' contains the highest concentration of terms associated with the most severe clinical characteristics.
+#| fig-height: 8
+#| fig-width: 10
+
+gpt_annot_plot_out$gp3 +
+  labs(x="Severity score") 
+```
 
@@
 #| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions.
-Heatmap of 10 represetantive phenotypes from each severity class (Profound, Severe, Moderate, Mild) stratified by whether the phenotypes are often/always congenital (**a**-**b**) or rarely/never congenital (**c**-**d**).
-Continuous severity scores are shown as bars (**b**,**d**) and were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights.
+Heatmap of 10 represetantive phenotypes from each severity class (Profound, Severe, Moderate, Mild) stratified by whether the phenotypes are often/always congenital **(a-b)** or rarely/never congenital **(c-d)**.
+Continuous severity scores are shown as bars **(b,d)** and were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights.
 The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score.
@@
 Overall, `r gsub("_"," ",names(cor_composite[1]))` seemed to be the strongest driver of the composite severity score with a Pearson correlation of `r mean(cor_composite[1],2)`, followed by `r gsub("_"," ",names(cor_composite[2]))` (r=`r round(cor_composite[2],2)`) and `r gsub("_"," ",names(cor_composite[3]))` (r=`r round(cor_composite[3],2)`).
+This is driven by the greater number of phenotypes 'always' causing impaired mobility (422), compared with, for instance, death (66), and intellectual disability (64).
 
@@
 
-{{< pagebreak >}
+{{< pagebreak >}}
 
@@
 
-```{r fig-severity-histo}
-#| label: fig-severity-histo
-#| fig-cap: Distribution of the composite GPT-4 severity score of the severity scores for all HPO terms.
-#| fig-height: 8
-#| fig-width: 10
-
-gpt_annot_plot_out$gp3 +
-  labs(x="Severity score") 
-```
-
 ```{r fig-severity-boxplot}

### 2025-09-10 docs/manuscript.qmd
condense discussion and modify

@@
 ## Abstract
+
 There are thousands of human phenotypes which are linked to genetic variation.
 These range from the benign (white eyelashes) to the deadly (respiratory failure).
-The Human Phenotype Ontology has categorised all human phenotypic variation into an unified framework that defines the relationships between them (e.g.
+The Human Phenotype Ontology has categorised a vast array of human phenotypic variation into a unified framework that defines the relationships between them (e.g.
 missing arms and missing legs are both abnormalities of the limb).
@@
 We approximated this approach using our GPT-4 annotations.
-This categorical approach showed a strong degree of positive correspondence with the continuous severity score ($\hat{\omega_{p}^2}$=`r round(fig_severity_class_dt$summary_data$estimate,2)`, p<`r format(.Machine$double.xmin, digits = 2)`).
-In other words, severity score increased with severity class level (mild < moderate < severe < profound) as expected.
+This categorical approach showed a strong degree of positive correspondence with the continuous severity score ($\hat{\omega_{p}^2}$=`r round(fig_severity_class_dt$summary_data$estimate,2)`, p\<`r format(.Machine$double.xmin, digits = 2)`).
+In other words, severity score increased with severity class level (mild \< moderate \< severe \< profound) as expected.
 The distribution of severity classes is shown in @fig-severity-class.
@@
 
-Here, we introduce a novel framework to leverage the current best-in-class LLM, GPT-4 [@openaiGPT4TechnicalReport2024], to systematically annotate the severity of `r n_ids_gpt` phenotypic abnormalities within the HPO.
+Here, we introduce a novel framework to leverage GPT-4 [@openaiGPT4TechnicalReport2024] to systematically annotate the severity of `r n_ids_gpt` phenotypic abnormalities within the HPO.
 By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without substantially compromising accuracy.
@@
 
-One key limitation of our study is the fact that we did not explicitly interrogate GPT-4 to assess how the availability of treatments affected the annotations it produced.
-For example, there are some very severe conditions for which highly effective treatments and early detection screens are widely available (e.g.
-syphilis, some forms of melanoma), thus rendering them fully treatable or even curable provided access to modern healthcare.
-It would therefore be useful to further interrogate GPT-4 to uncover how the availability of treatments influences its responses.
-Many of our findings here seem to indicate that GPT-4 does take into account quality of care to the extent that health services increase the likelihood of desired outcomes.
-For example, many of the cancer phenotypes are justified as always or often causing death unless detected and treated early in the disease course.
-On the other hand, some cancers are justified as rarely causing death if appropriate treatment is provided, which may not always be the case for individuals or populations with access to less access to quality healthcare services.
-Future efforts could more explicitly ask GPT-4 whether the phenotype would cause death with no or suboptimal treatment.
+One limitation of this study is that it did not test how GPT-4 considers treatment availability when making annotations.
+Some severe conditions, like syphilis or certain melanomas, are curable with early detection and modern healthcare.
+GPT-4 often linked outcomes to access and quality of care, describing some cancers as fatal unless caught early while noting others are rarely lethal if treatment is available.
+This suggests the model accounts for healthcare quality but not consistently.
+The study also did not capture differences in severity within the same phenotype, such as light sensitivity ranging from mild to disabling.
+Nor did it consider onset, disease course, or how phenotypes change in specific contexts, such as hypertension being mild in most people but severe in rare genetic disorders.
+These gaps highlight areas for future work.
 
-Another limitation with the present dataset is that phenotypes themselves can manifest with different degrees of severity, in the sense that they are more pronounced or intense.
-For example, sensitivity to light could range from a mild inconvenience to a severe disability that prevents the individual from leaving their home during the day.
-The effect of onset (beyond congenital vs.
-non-congenital) and time course (acute, slowly progression, relapse-remitting) were also not explicitly considered.
-Finally, we did not ask GPT-4 to consider phenotypes as they present within particular diseases.
-For example, while the phenotype 'Hypertension' may be mild to moderate in the general population and not present until middle-age, it can also present early in life as very severe in the context of a rare genetic disorder such as Liddle syndrome.
-Future work could explore these nuances in more detail.
-
-In addition to these technical challenges, there are multiple factors that need to be considered when trying to prioritise phenotypes for their suitability for gene therapy development.
-First, while we have attempted to formalise severity here, this is an inherently subjective concept that may vary considerably across different individuals and contexts.
-For instance, one could ask whether a condition that always causes death is worse than a condition that causes a lifetime of severe disability (e.g.
-paralysis, blindness, intellectual disability).
-Metrics such as quality-adjusted life years (QALYs) have been proposed in the past to address these dilemmas by defining health as a function of both the length and quality of life [@prietoProblemsSolutionsCalculating2003].
-With regards to the financial burden of diseases, in some situations phenotypes which require many years of expensive medical care may be prioritised over those that result in extremely early onset lethality and little opportunity for therapeutic intervention.
-Another factor that affects the viability of a therapeutic program is the speed, cost and other practical considerations of a clinical trial.
-For instance, measuring risk of ageing-related respiratory failure over a ten-year period may be impractical in some cases.
-However, testing for total reversal of an existing severe phenotype could potentially yield faster and more immediately impactful results.
-If performed in close collaboration with medical ethicists, governmental organisations, advocacy groups and patient families, such cost/benefit assessments could be aided by LLMs through the scalable gathering of relevant data.
-As AI capabilities continue to advance, the range of applications in which they can be used effectively will continue to grow.
-
-While our study demonstrates the feasibility and utility of AI-driven phenotypic annotation, several limitations must be acknowledged.
-The reliance on computational algorithms may introduce biases or inaccuracies inherent to the training data, necessitating ongoing validation and refinement of our approach.
-Additionally, our severity scoring system, while comprehensive, may not capture the full spectrum of phenotypic variability or account for complex gene-environment interactions.
-Future research should focus on further optimising AI-driven annotation methodologies, incorporating additional data modalities such as genomic and clinical data to enhance accuracy.
+Therapeutic prioritization adds further challenges.
+Severity is subjective; a disease that always causes death may not seem “worse” than lifelong paralysis or blindness.
+Cost also shapes priorities, as some conditions require years of expensive care while others cause early death with little chance of intervention.
+Practical constraints must also be considered as long-term trials are difficult, whereas testing therapies that quickly reverse severe symptoms may yield faster results.
+These choices should involve ethicists, policymakers, advocacy groups, and families, with LLMs helping gather data at scale.
+While this study shows the promise of AI-driven phenotypic annotation, it also reveals risks of bias, gaps in severity scoring, and the need to factor in gene–environment interactions.
+Finally, a flurry of new LLMs continue to be released regularly.
+Here, we use GPT-4 as an example but note that our framework is equally applicable to most other LLMs.
+Future work should refine methods, add genomic and clinical data, and validate results to make AI a stronger tool in precision medicine.
 
@@
 
-\vspace{2em}
+{{< pagebreak >}}
 
@@
 
-```
-
-```{r fig-severity-histo}
-#| label: fig-severity-histo
-#| fig-cap: Distribution of the composite GPT-4 severity score of the severity scores for all HPO terms.
-#| fig-height: 8
-#| fig-width: 10
-
-gpt_annot_plot_out$gp3 +
-  labs(x="Severity score") 
 ```
@@
 ::: {#tbl-annotations}
-[**Top phenotype annotations table**](https://github.com/neurogenomics/gpt_hpo_annotations/raw/master/data/`r basename(top_phenos_path)`)
+[**Top phenotype annotations table**](https://github.com/neurogenomics/gpt_hpo_annotations/raw/master/data/%60r%20basename(top_phenos_path)%60)
 
 Table of GPT-4 clinical characterstic annotations for all Human Phenotype Ontology (HPO) phenotypes in @fig-top-phenos.
-For each phenotype, this includes the name of the phenotype ('hpo_name'), the ID of the phenotype ('hpo_id'), the frequency of each annotation (always, often, rarely, never), and the justification for each annotation ('..._justification').
+For each phenotype, this includes the name of the phenotype ('hpo_name'), the ID of the phenotype ('hpo_id'), the frequency of each annotation (always, often, rarely, never), and the justification for each annotation ('...\_justification').
 These results can also be downloaded programmatically using the R function `HPOExplorer::gpt_annot_check`.
